---
title: 'Figure 1: PerIS in homeostasis'
author: "Andrew Y.F. Li Yim"
date: '2023-07-26'
output: html_document
---

```{r libraries, include=FALSE}
library(Seurat)
library(ggplot2)
library(dplyr)
library(ggrastr)
library(ggrepel)
library(ggpubr)
library(ggalt)
library(viridis)
library(readr)
```

If the snakemake pipeline was run as-is, the following files will be generated in the folder from which the pipeline was run. If necessary, adjust paths accordingly.

## Setup

### Paths

```{r paths}
#base_path <- "/mnt/smb/ltytgat/ayliyim/MI-group/ayliyim/projects/TIPRJ0000015_MAPS/maps_crc" 
base_path <- "."

# Subsets
seurat_hc_pbmc_pf_rds <- file.path(base_path, "output/q1_pf_characterization/subsets/hc_pbmc_pf_SeuratObject.Rds")
seurat_hc_pf_rds <- file.path(base_path, "output/q1_pf_characterization/subsets/hc_pf_SeuratObject.Rds")
seurat_hc_pf_myeloid_rds <- file.path(base_path, "output/q1_pf_characterization/subsets/hc_pf_Myeloid_SeuratObject.Rds")
seurat_hc_pf_macrophages_rds <- file.path(base_path, "output/q1_pf_characterization/subsets/hc_pf_macrophages_SeuratObject.Rds")

# Plotting parameters
celltype_markers_xlsx <- file.path(base_path, "config/order/celltype_markers.xlsx")
pf_heatmap_order_xlsx <- file.path(base_path, "config/order/q1_pf_heatmap_order.xlsx")
```

### Import data

```{r load subsets}
seurat_hc_pbmc_pf <- readRDS(seurat_hc_pbmc_pf_rds)
seurat_hc_pf <- readRDS(seurat_hc_pf_rds)
seurat_hc_pf_myeloid <- readRDS(seurat_hc_pf_myeloid_rds)
seurat_hc_pf_macrophages <- readRDS(seurat_hc_pf_macrophages_rds)
```

### Output

```{r figures}
fig1Dir <- file.path(base_path, "docs/manuscript/figures/fig1")
dir.create(fig1Dir, recursive = T)
```

### Plotting parameters

```{r plotting order}
celltype_order_pf_l2_l3 <- readxl::read_excel(celltype_markers_xlsx, col_names = T) %>%
  dplyr::filter(level == "manual_l3",
                modality == "gene") %>%
  dplyr::select(celltype, color) %>%
  unique() %>%
  dplyr::filter(celltype %in% unique(seurat_hc_pf@meta.data[,"manual_l3"])) %>%
  dplyr::mutate(celltype = ifelse(celltype %in% c("Classical monocytes", "Macrophages VCAN+", "Macrophages C1Q+", "Macrophages VCAN+C1Q+", "Macrophages SPP1+"), "Mono-macs", celltype),
                color = ifelse(celltype %in% c("Mono-macs"), "#E31A1C", color)) %>%
  unique() %>%
  dplyr::mutate(number_subset = paste0(1:nrow(.), ". ", celltype))

celltype_order_pf_l4 <- readxl::read_excel(celltype_markers_xlsx, col_names = T) %>%
  dplyr::filter(level == "manual_l4",
                modality == "gene") %>%
  dplyr::select(celltype, color) %>%
  unique() %>%
  dplyr::filter(celltype %in% unique(seurat_hc_pf@meta.data[,"manual_l4"])) %>%
  dplyr::mutate(number_subset = paste0(1:nrow(.), ". ", celltype))
```

```{r colors}
tissue_colors <- c(PF = "#224FBD", PBMC = "#F30000", Colon = "#7B3F00", Liver = "#AA336A")

manual_l2_l3_colors_pf <- celltype_order_pf_l2_l3$color
names(manual_l2_l3_colors_pf) <- celltype_order_pf_l2_l3$celltype

manual_l4_colors_pf <- celltype_order_pf_l4$color
names(manual_l4_colors_pf) <- celltype_order_pf_l4$celltype
```

## Figures

### UMAP HC PBMC PF colored by tissue

Here we show the unsupervised clustering analysis using UMAP of PBMC- and PF-derived immune cells colored by tissue.

```{r umap hc pbmc pf coltissue, fig.width=7.5, fig.height=7.5}
umap_hc_pbmc_pf_coltissue_ggplotobj <- data.frame(CB = colnames(seurat_hc_pbmc_pf),
                                             Embeddings(seurat_hc_pbmc_pf[["umap"]]),
                                             seurat_hc_pbmc_pf@meta.data) %>%
  ggplot(aes(x = UMAP_1, y = UMAP_2)) +
  geom_point_rast(show.legend = T, size = 0.25, aes(col = Tissue)) +
  labs(y = "",
       x = "") +
  guides(colour = guide_legend(override.aes = list(size = 3))) +
  scale_color_manual(values = tissue_colors) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.x = element_text(face = "bold"))

pdf(file.path(figS1Dir, "umap_hc_pbmc_pf_coltissue.pdf"), width = 7.5, height = 7.5)
print(umap_hc_pbmc_pf_coltissue_ggplotobj)
dev.off()
```

### UMAP HC PF colored by celltype

Here we show the unsupervised clustering analysis using UMAP of PF-derived immune cells annotated using the celltype.

```{r umap hc pf colcelltype, fig.width=7.5, fig.height=7.5}
umap_hc_pf_colcelltype_ggplotobj <- data.frame(CB = colnames(seurat_hc_pf),
                                               Embeddings(seurat_hc_pf[["wnn.umap"]]),
                                               seurat_hc_pf@meta.data) %>%
  dplyr::mutate(celltype = manual_l3,
                celltype = ifelse(celltype %in% c("Classical monocytes", "Macrophages C1Q+", "Macrophages SPP1+", "Macrophages VCAN+", "Macrophages VCAN+C1Q+"), "Mono-macs", celltype),
                celltype = factor(celltype, levels = celltype_order_pf_l2_l3$celltype),
                celltype_number = as.numeric(celltype),
                celltype_w_number = paste0(as.numeric(celltype), ". ", celltype),
                celltype_w_number = factor(celltype_w_number, levels = celltype_order_pf_l2_l3$number_subset)) %>% 
  ggplot(aes(x = wnnUMAP_1, y = wnnUMAP_2)) +
  geom_point_rast(show.legend = T, size = 0.25, aes(col = celltype)) +
  labs(y = "",
       x = "") +
  guides(colour = guide_legend(override.aes = list(size = 3))) +
  scale_color_manual(values = manual_l2_l3_colors_pf) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.x = element_text(face = "bold"))

pdf(file.path(fig1Dir, "umap_hc_pf_colcelltype.pdf"), width = 7.5, height = 7.5)
print(umap_hc_pf_colcelltype_ggplotobj)
dev.off()
```

### UMAP HC PF Myeloid colored by myeloid GEX markers

Here we characterize the PF-derived ITGAM+ myeloid cells by superimposing the gene expression of the following genes on the respective UMAP configuration:
- Myeloid (ITGAM)
- PDCs (CLEC4C) 
- CDC1s (XCR1)
- CDC2s (CDC1)
- Macrophages (MARCO)
- Macrophages (CD163)
- Monocytes/Monocyte-macrophages (CCR2)
- Classical monocytes/Monocyte-macrophages (CD14)

```{r umap hc pf col_myeloidgexmarkers, fig.width=9, fig.height=9}
umap_hc_pf_myeloid_colmyeloidgenemarkers_ggplotobj_list <- lapply(
  data.frame(CB = colnames(seurat_hc_pf_myeloid),
             Embeddings(seurat_hc_pf_myeloid[["wnn.umap"]]),
             seurat_hc_pf_myeloid@meta.data,
             expr = GetAssayData(seurat_hc_pf_myeloid, assay = "RNA")["CLEC4C",],
             Feature = "CLEC4C") %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_hc_pf_myeloid),
                                  Embeddings(seurat_hc_pf_myeloid[["wnn.umap"]]),
                                  seurat_hc_pf_myeloid@meta.data,
                                  expr = GetAssayData(seurat_hc_pf_myeloid, assay = "RNA")["XCR1",],
                                  Feature = "XCR1")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_hc_pf_myeloid),
                                  Embeddings(seurat_hc_pf_myeloid[["wnn.umap"]]),
                                  seurat_hc_pf_myeloid@meta.data,
                                  expr = GetAssayData(seurat_hc_pf_myeloid, assay = "RNA")["CD1C",],
                                  Feature = "CD1C")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_hc_pf_myeloid),
                                  Embeddings(seurat_hc_pf_myeloid[["wnn.umap"]]),
                                  seurat_hc_pf_myeloid@meta.data,
                                  expr = GetAssayData(seurat_hc_pf_myeloid, assay = "RNA")["ITGAM",],
                                  Feature = "ITGAM")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_hc_pf_myeloid),
                                  Embeddings(seurat_hc_pf_myeloid[["wnn.umap"]]),
                                  seurat_hc_pf_myeloid@meta.data,
                                  expr = GetAssayData(seurat_hc_pf_myeloid, assay = "RNA")["MARCO",],
                                  Feature = "MARCO")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_hc_pf_myeloid),
                                  Embeddings(seurat_hc_pf_myeloid[["wnn.umap"]]),
                                  seurat_hc_pf_myeloid@meta.data,
                                  expr = GetAssayData(seurat_hc_pf_myeloid, assay = "RNA")["CD14",],
                                  Feature = "CD14")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_hc_pf_myeloid),
                                  Embeddings(seurat_hc_pf_myeloid[["wnn.umap"]]),
                                  seurat_hc_pf_myeloid@meta.data,
                                  expr = GetAssayData(seurat_hc_pf_myeloid, assay = "RNA")["CCR2",],
                                  Feature = "CCR2")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_hc_pf_myeloid),
                                  Embeddings(seurat_hc_pf_myeloid[["wnn.umap"]]),
                                  seurat_hc_pf_myeloid@meta.data,
                                  expr = GetAssayData(seurat_hc_pf_myeloid, assay = "RNA")["CD163",],
                                  Feature = "CD163")) %>%
    dplyr::mutate(expr_rank = rank(expr, ties.method="first"),
                  Feature = factor(Feature, levels = c("CLEC4C", "XCR1", "CD1C", "ITGAM", "MARCO", "CD14", "CCR2", "CD163"))) %>%
    dplyr::filter(manual_l2 %in% c("Macrophages", "CDCs", "PDCs")) %>%
    split(., .$Feature), FUN = function(plotdf){
      plotdf %>% dplyr::arrange(expr) %>%
        ggplot(aes(x = wnnUMAP_1, y = wnnUMAP_2)) +
        geom_point_rast(show.legend = F, size = 0.25, col = "#d3d3d3") +
        geom_point_rast(data = . %>%
                          dplyr::filter(expr>0), aes(col = expr), show.legend = T) +
        guides(colour = guide_legend(override.aes = list(size = 3))) +
        facet_wrap(~Feature) +
        guides(colour = guide_legend(override.aes = list(size = 3),
                                     title = "Expression")) +
        xlim(-6,11) +
        ylim(-5,8) +
        scale_color_viridis() +
        theme_bw() +
        theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              legend.position = "none",
              axis.title.x = element_blank(),
              axis.title.y = element_blank(),
              strip.text.x = element_text(face = "bold"))
    }
)

pdf(file.path(fig1Dir, "umap_hc_pf_myeloid_colmyeloidgenemarkers.pdf"), width = 9, height = 9)
print(ggarrange(plotlist = umap_hc_pf_myeloid_colmyeloidgenemarkers_ggplotobj_list, nrow = 3, ncol = 3))
dev.off()
```

### UMAP HC PF Myeloid colored by myeloid CITE markers

Here we characterize the PF-derived ITGAM+ myeloid cells by superimposing the protein expression of the following genes on the respective UMAP configuration:
- CDCs (CD1C)
- Macrophages (CD163)
- Monocytes/Monocyte-macrophages (CD192 [CCR2])
- Classical monocytes/Monocyte-macrophages (CD14)

```{r umap hc pf col_myeloidcitemarkers, fig.width=12, fig.height=3}
umap_hc_pf_myeloid_colmyeloidcitemarkers_ggplotobj_list <- lapply(
  data.frame(CB = colnames(seurat_hc_pf_myeloid),
             Embeddings(seurat_hc_pf_myeloid[["wnn.umap"]]),
             seurat_hc_pf_myeloid@meta.data,
             expr = GetAssayData(seurat_hc_pf_myeloid, assay = "CITE")["Hu.CD1c",],
             Feature = "CD1C") %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_hc_pf_myeloid),
                                  Embeddings(seurat_hc_pf_myeloid[["wnn.umap"]]),
                                  seurat_hc_pf_myeloid@meta.data,
                                  expr = GetAssayData(seurat_hc_pf_myeloid, assay = "CITE")["Hu.CD14-M5E2",],
                                  Feature = "CD14")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_hc_pf_myeloid),
                                  Embeddings(seurat_hc_pf_myeloid[["wnn.umap"]]),
                                  seurat_hc_pf_myeloid@meta.data,
                                  expr = GetAssayData(seurat_hc_pf_myeloid, assay = "CITE")["Hu.CD192",],
                                  Feature = "CD192")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_hc_pf_myeloid),
                                  Embeddings(seurat_hc_pf_myeloid[["wnn.umap"]]),
                                  seurat_hc_pf_myeloid@meta.data,
                                  expr = GetAssayData(seurat_hc_pf_myeloid, assay = "CITE")["Hu.CD163",],
                                  Feature = "CD163")) %>%
    dplyr::mutate(expr_rank = rank(expr, ties.method="first"),
                  Feature = factor(Feature, levels = c("CD1C", "CD14", "CD192", "CD163"))) %>%
    dplyr::filter(manual_l2 %in% c("Macrophages", "CDCs", "PDCs")) %>%
    split(., .$Feature), FUN = function(plotdf){
      plotdf %>% 
        dplyr::arrange(expr) %>%
        ggplot(aes(x = wnnUMAP_1, y = wnnUMAP_2)) +
        geom_point_rast(show.legend = F, size = 0.25, col = "#d3d3d3") +
        geom_point_rast(data = . %>%
                          dplyr::filter(expr>0), aes(col = expr), show.legend = T) +
        guides(colour = guide_legend(override.aes = list(size = 3))) +
        facet_wrap(~Feature) +
        guides(colour = guide_legend(override.aes = list(size = 3),
                                     title = "Expression")) +
        xlim(-6,11) +
        ylim(-5,8) +
        scale_colour_gradient(low = "#ffffff", high = "#FF0000") +
        theme_bw() +
        theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              legend.position = "none",
              axis.title.x = element_blank(),
              axis.title.y = element_blank(),
              strip.text.x = element_text(face = "bold"))
    }
)

pdf(file.path(fig1Dir, "umap_hc_pf_myeloid_colmyeloidcitemarkers.pdf"), width = 12, height = 3)
print(ggarrange(plotlist = umap_hc_pf_myeloid_colmyeloidcitemarkers_ggplotobj_list, nrow = 1, ncol = 4))
dev.off()
```

### UMAP HC PF macrophages colored by macrophage GEX markers

Here we characterize the PF-derived CD163+ macrophages by superimposing the gene expression of the following genes on the respective UMAP configuration:
- [tissue residency] (VSIG4)
- Macrophages (CD163)
- Monocytes/Monocyte-macrophages/Macrophages VCAN+ (VCAN)
- Monocytes/Monocyte-macrophages (CCR2)
- Macrophages C1Q+ (C1QA)
- Macrophages SPP1+ (SPP1)

```{r umap hc pf macrophages col_macrophagegexmarkers, fig.width=9, fig.height=6}
umap_hc_pf_macrophages_colmacrophagegexmarkers_ggplotobj_list <- lapply(
  data.frame(CB = colnames(seurat_hc_pf_macrophages),
             Embeddings(seurat_hc_pf_macrophages[["gex.umap"]]),
             seurat_hc_pf_macrophages@meta.data,
             expr = GetAssayData(seurat_hc_pf_macrophages, assay = "RNA")["VSIG4",],
             Feature = "VSIG4") %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_hc_pf_macrophages),
                                  Embeddings(seurat_hc_pf_macrophages[["gex.umap"]]),
                                  seurat_hc_pf_macrophages@meta.data,
                                  expr = GetAssayData(seurat_hc_pf_macrophages, assay = "RNA")["CD163",],
                                  Feature = "CD163")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_hc_pf_macrophages),
                                  Embeddings(seurat_hc_pf_macrophages[["gex.umap"]]),
                                  seurat_hc_pf_macrophages@meta.data,
                                  expr = GetAssayData(seurat_hc_pf_macrophages, assay = "RNA")["VCAN",],
                                  Feature = "VCAN")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_hc_pf_macrophages),
                                  Embeddings(seurat_hc_pf_macrophages[["gex.umap"]]),
                                  seurat_hc_pf_macrophages@meta.data,
                                  expr = GetAssayData(seurat_hc_pf_macrophages, assay = "RNA")["CCR2",],
                                  Feature = "CCR2")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_hc_pf_macrophages),
                                  Embeddings(seurat_hc_pf_macrophages[["gex.umap"]]),
                                  seurat_hc_pf_macrophages@meta.data,
                                  expr = GetAssayData(seurat_hc_pf_macrophages, assay = "RNA")["C1QA",],
                                  Feature = "C1QA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_hc_pf_macrophages),
                                  Embeddings(seurat_hc_pf_macrophages[["gex.umap"]]),
                                  seurat_hc_pf_macrophages@meta.data,
                                  expr = GetAssayData(seurat_hc_pf_macrophages, assay = "RNA")["SPP1",],
                                  Feature = "SPP1")) %>%
    dplyr::mutate(expr_rank = rank(expr, ties.method="first"),
                  Feature = factor(Feature, levels = c("VSIG4", "CD163", "VCAN", "CCR2", "C1QA", "SPP1"))) %>%
    split(., .$Feature), FUN = function(plotdf){
      plotdf %>% dplyr::arrange(expr) %>%
        ggplot(aes(x = gexUMAP_1, y = gexUMAP_2)) +
        geom_point_rast(show.legend = F, size = 0.25, col = "#d3d3d3") +
        geom_point_rast(data = . %>%
                          dplyr::filter(expr>0), aes(col = expr), show.legend = T) +
        guides(colour = guide_legend(override.aes = list(size = 3))) +
        facet_wrap(~Feature) +
        guides(colour = guide_legend(override.aes = list(size = 3),
                                     title = "Expression")) +
        scale_color_viridis() +
        theme_bw() +
        theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              legend.position = "none",
              axis.title.x = element_blank(),
              axis.title.y = element_blank(),
              strip.text.x = element_text(face = "bold"))
    }
)

pdf(file.path(fig1Dir, "umap_hc_pf_macrophages_colmacrophagegexmarkers.pdf"), width = 9, height = 6)
print(ggarrange(plotlist = umap_hc_pf_macrophages_colmacrophagegexmarkers_ggplotobj_list, nrow = 2, ncol = 3))
dev.off()
```

### UMAP HC PF macrophages colored by macrophage CITE markers

Here we characterize the CD163+ macrophages by superimposing the protein expression of the following genes on the respective UMAP configuration:
- Macrophages (CD163)

```{r umap hc pf macrophages col_macrophagecitemarkers, fig.width=3, fig.height=3}
umap_hc_pf_macrophages_colmacrophagecitemarkers_ggplotobj <- 
  data.frame(CB = colnames(seurat_hc_pf_macrophages),
             Embeddings(seurat_hc_pf_macrophages[["gex.umap"]]),
             seurat_hc_pf_macrophages@meta.data,
             expr = GetAssayData(seurat_hc_pf_macrophages, assay = "CITE")["Hu.CD163",],
             Feature = "CD163") %>%
  dplyr::mutate(expr_rank = rank(expr, ties.method="first"),
                Feature = factor(Feature, levels = c("CD163"))) %>%
  dplyr::arrange(expr) %>%
  ggplot(aes(x = gexUMAP_1, y = gexUMAP_2)) +
  geom_point_rast(show.legend = F, size = 0.25, col = "#d3d3d3") +
  geom_point_rast(data = . %>%
                    dplyr::filter(expr>0), aes(col = expr), show.legend = T) +
  guides(colour = guide_legend(override.aes = list(size = 3))) +
  facet_wrap(~Feature) +
  guides(colour = guide_legend(override.aes = list(size = 3),
                               title = "Expression")) +
  scale_colour_gradient(low = "#ffffff", high = "#FF0000") +
  theme_bw() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        strip.text.x = element_text(face = "bold"))

pdf(file.path(fig1Dir, "umap_hc_pf_macrophages_colmacrophagecitemarkers.pdf"), width = 3, height = 3)
print(umap_hc_pf_macrophages_colmacrophagecitemarkers_ggplotobj)
dev.off()
```

### Boxplot HC PF myeloid percentage relative to all myeloid

Here we prepare the data for scattered boxplots depicting the PF-derived myeloid subsets relative to the total PF-derived myeloid population. Actual data was plotted in Graphpad Prism.

```{r hc pf myeloid l3rl1 data}
seurat_hc_pf_myeloid@meta.data %>%
  dplyr::group_by(manual_l3, SampleID, .drop = T) %>% 
  dplyr::summarise(Ncells = n()) %>%
  dplyr::ungroup() %>%
  tidyr::complete(tidyr::nesting(SampleID), tidyr::nesting(manual_l3), fill = list(Ncells = 0)) %>%
  dplyr::group_by(SampleID) %>%
  dplyr::mutate(Nlrefsample = sum(Ncells),
                Ncellprop = Ncells/Nlrefsample,
                Ncellprop = ifelse(is.na(Ncellprop), 0, Ncellprop),
                Ncellperc = Ncellprop*100,
                Donor = gsub("^(pt[0-9]{2})_.+$", "\\1", SampleID)) %>%
  readr::write_csv(file.path(fig1Dir, "hc_pf_mnp_l3rl1.csv"))
```
