---
title: 'Figure 3: TIME of PM-CRC'
author: "Andrew Y.F. Li Yim"
date: '2023-07-31'
output: html_document
---

```{r libraries, include=FALSE}
library(Seurat)
library(ggplot2)
library(dplyr)
library(ggrastr)
library(ggrepel)
library(ggpubr)
library(viridis)
library(DESeq2)
library(ggalt)
```

If the snakemake pipeline was run as-is, the following files will be generated in the folder from which the pipeline was run. If not, then adjust paths accordingly.    

## Setup

### Paths

```{r paths}
base_path <- "/mnt/smb/ltytgat/ayliyim/MI-group/ayliyim/projects/TIPRJ0000015_MAPS/maps_crc" # L-drive

# Subsets
seurat_crcpmp_tx_immune_rds <- file.path(base_path, "output/q3_pm_tx_characterization/subsets/crcpmp_tx_immune_SeuratObject.Rds")
seurat_crcpmp_tx_monocytes_macrophages_rds <- file.path(base_path, "output/q3_pm_tx_characterization/subsets/crcpmp_tx_monocytes_macrophages_SeuratObject.Rds")
seurat_crcpmp_tx_macrophages_rds <- file.path(base_path, "output/q3_pm_tx_characterization/subsets/crcpmp_tx_macrophages_SeuratObject.Rds")

# Analyses
crcpmp_pf_tx_paired_macrophages_tamannotation_rds <- file.path(base_path, "output/q3_pm_tx_characterization/analyses/crcpmp_pf_tx_paired_macrophages_tamannotation.Rds")

# Plotting parameters
celltype_markers_xlsx <- file.path(base_path, "config/order/celltype_markers.xlsx")

# Genes of interest
tam_markers_xlsx <- file.path(base_path, "config/genes_of_interest/tams_ma_2022.xlsx")
azizi2018_xlsx <- file.path(base_path, "config/genes_of_interest/m1_m2_azizi2018.xlsx")
```

### Import data

```{r load subsets}
seurat_crcpmp_tx_monocytes_macrophages <- readRDS(seurat_crcpmp_tx_monocytes_macrophages_rds)
seurat_crcpmp_tx_macrophages <- readRDS(seurat_crcpmp_tx_macrophages_rds)
seurat_crcpmp_tx_immune <- readRDS(seurat_crcpmp_tx_immune_rds)
```

```{r load macrophage markers}
m1_markers <- readxl::read_excel(azizi2018_xlsx, sheet = 2)
m2_markers <- readxl::read_excel(azizi2018_xlsx, sheet = 3)
```

```{r load tam markers}
tam_markers <- readxl::read_excel(tam_markers_xlsx)
crcpmp_pf_tx_paired_macrophages_tamannotation <- readRDS(crcpmp_pf_tx_paired_macrophages_tamannotation_rds)
```

### Output

```{r figures}
fig3Dir <- file.path(base_path, "docs/manuscript/figures/fig3")
dir.create(fig3Dir, recursive = T)
```

### Plotting parameters

```{r plotting order}
pf_heatmap_order_xlsx <- file.path(base_path, "config/order/q1_pf_heatmap_order.xlsx")

celltype_order_l3_tx_immune <- readxl::read_excel(celltype_markers_xlsx, col_names = T) %>%
  dplyr::filter(level == "manual_l3",
                modality == "gene") %>%
  dplyr::select(celltype, color) %>%
  unique() %>%
  dplyr::filter(!is.na(color),
                celltype %in% seurat_crcpmp_tx_immune$manual_l3) %>%
  dplyr::mutate(number_subset = paste0(1:nrow(.), ". ", celltype))

celltype_order_l4_monocytes_macrophages <- readxl::read_excel(celltype_markers_xlsx, col_names = T) %>%
  dplyr::filter(level == "manual_l4",
                celltype %in% c("Classical monocytes", "Non-classical monocytes", "Macrophages VCAN+", "Macrophages C1Q+", "Macrophages VCAN+C1Q+", "Macrophages SPP1+"),
                modality == "gene") %>%
  dplyr::select(celltype, color) %>%
  unique() %>%
  dplyr::filter(!is.na(color)) %>%
  dplyr::mutate(number_subset = paste0(1:nrow(.), ". ", celltype))

celltype_order_l4_macrophages <- readxl::read_excel(celltype_markers_xlsx, col_names = T) %>%
  dplyr::filter(level == "manual_l4",
                celltype %in% c("Macrophages C1Q+", "Macrophages VCAN+", "Macrophages VCAN+C1Q+", "Macrophages SPP1+"),
                modality == "gene") %>%
  dplyr::select(celltype, color) %>%
  unique() %>%
  dplyr::filter(!is.na(color)) %>%
  dplyr::mutate(number_subset = paste0(1:nrow(.), ". ", celltype))
```

```{r colors}
tissue_colors <- c(PBMC = "#F30000", PF = "#224FBD", TX = "#855522")

manual_l3_tx_immune_colors <- celltype_order_l3_tx_immune$color
names(manual_l3_tx_immune_colors) <- celltype_order_l3_tx_immune$number_subset

manual_l4_monocytes_macrophages_number_colors <- celltype_order_l4_monocytes_macrophages$color
names(manual_l4_monocytes_macrophages_number_colors) <- celltype_order_l4_monocytes_macrophages$number_subset

manual_l4_macrophages_colors <- celltype_order_l4_macrophages$color
names(manual_l4_macrophages_colors) <- celltype_order_l4_macrophages$celltype
```

## Figures

### UMAP PM-CRC TX monocytes and macrophages colored by l4

```{r umap crcpmp pf monocytes macrophages col_l4, fig.width=5, fig.height=5}
umap_crcpmp_tx_monocytes_macrophages_coll4_ggplotobj <- data.frame(CB = colnames(seurat_crcpmp_tx_monocytes_macrophages),
                                                                Embeddings(seurat_crcpmp_tx_monocytes_macrophages[["umap"]]),
                                                                seurat_crcpmp_tx_monocytes_macrophages@meta.data) %>%
  dplyr::mutate(Group = factor(Group, levels = c("HC", "CRC+")),
                celltype = factor(manual_l4, levels = celltype_order_l4_monocytes_macrophages$celltype),
                celltype_number = as.numeric(celltype),
                celltype_w_number = paste0(as.numeric(celltype), ". ", celltype),
                celltype_w_number = factor(celltype_w_number)) %>% 
  ggplot(aes(x = UMAP_1, y = UMAP_2)) +
  #geom_point_rast(show.legend = T, size = 0.5, col = "black") +
  geom_point_rast(show.legend = T, size = 3, aes(col = celltype_w_number)) +
  labs(y = "",
       x = "") +
  geom_label_repel(data = . %>%
                     dplyr::group_by(celltype, celltype_number, celltype_w_number) %>%
                     summarize(x = median(x = UMAP_1),
                               y = median(x = UMAP_2)),
                   mapping = aes(label = celltype_number, x = x, y = y),
                   alpha = 1, 
                   size = 8,
                   label.size=0.25,
                   show.legend = F) +
  guides(colour = guide_legend(override.aes = list(size = 3))) +
  scale_color_manual(values = manual_l4_monocytes_macrophages_number_colors) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.x = element_text(face = "bold"))

print(umap_crcpmp_tx_monocytes_macrophages_coll4_ggplotobj)

pdf(file.path(fig3Dir, "umap_crcpmp_tx_monocytes_macrophages_coll4.pdf"), width = 5, height = 5)
print(umap_crcpmp_tx_monocytes_macrophages_coll4_ggplotobj)
dev.off()
```
### Boxplot PM-CRC TX macrophage subsets percentage relative to all macrophages

Here we prepare the data for scattered boxplots depicting the TX-derived macrophage subsets relative to the total TX-derived macrophage population. Actual data was plotted in Graphpad Prism.

```{r crcpmp tx myeloid l3rl1 data}
seurat_crcpmp_tx_macrophages@meta.data %>%
  dplyr::group_by(manual_l3, SampleID, .drop = T) %>% 
  dplyr::summarise(Ncells = n()) %>%
  dplyr::ungroup() %>%
  tidyr::complete(tidyr::nesting(SampleID), tidyr::nesting(manual_l3), fill = list(Ncells = 0)) %>%
  dplyr::group_by(SampleID) %>%
  dplyr::mutate(Nlrefsample = sum(Ncells),
                Ncellprop = Ncells/Nlrefsample,
                Ncellprop = ifelse(is.na(Ncellprop), 0, Ncellprop),
                Ncellperc = Ncellprop*100,
                Donor = gsub("^(pt[0-9]{2})_.+$", "\\1", SampleID)) %>%
  readr::write_csv(file = file.path(fig3Dir, "crcpmp_tx_macrophages_l3rl2.csv"))
```




### UMAP PM-CRC TX monocytes and macrophages colored by myeloid markers

```{r umap crcpmp tx monocytes macrophages col_myeloidmarkers, fig.width=3, fig.height=3}
umap_crcpmp_tx_monocytes_macrophages_colmyeloidgenemarkers_ggplotobj <- data.frame(
  CB = colnames(seurat_crcpmp_tx_monocytes_macrophages),
  Embeddings(seurat_crcpmp_tx_monocytes_macrophages[["umap"]]),
  seurat_crcpmp_tx_monocytes_macrophages@meta.data,
  expr = GetAssayData(seurat_crcpmp_tx_monocytes_macrophages, assay = "RNA")["VSIG4",],
  Feature = "VSIG4") %>%
  dplyr::mutate(expr_rank = rank(expr, ties.method="first")) %>% 
  dplyr::arrange(expr) %>%
  ggplot(aes(x = UMAP_1, y = UMAP_2)) +
  geom_point_rast(show.legend = F, size = 0.25, col = "#d3d3d3") +
  geom_point_rast(data = . %>%
                    dplyr::filter(expr>0), aes(col = expr), show.legend = T) +
  guides(colour = guide_legend(override.aes = list(size = 3))) +
  facet_wrap(~Feature) +
  guides(colour = guide_legend(override.aes = list(size = 3),
                               title = "Expression")) +
  scale_color_viridis() +
  theme_bw() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        strip.text.x = element_text(face = "bold"))

pdf(file.path(fig3Dir, "umap_crcpmp_tx_monocytes_macrophages_colmyeloidgenemarkers.pdf"), width = 3, height = 3)
print(umap_crcpmp_tx_monocytes_macrophages_colmyeloidgenemarkers_ggplotobj)
dev.off()
```

### Dotplot GEX PM-CRC TX macrophages chemotaxis receptors

```{r dotplot gex crcpmp tx macrophages col_chemotaxisreceptors, fig.width=1.7, fig.height=4.5}
chemotaxis_receptors <- c("CCR1", "CCR3", "CCR4", "CCR5", "CCR6", "CXCR3", "CXCR4", "CSF1R", "CSF2RA")

crcpmp_tx_macrophages_chemotaxisreceptors_gex <- GetAssayData(seurat_crcpmp_tx_macrophages, assay = "RNA")[which(rownames(GetAssayData(seurat_crcpmp_tx_macrophages, assay = "RNA")) %in% chemotaxis_receptors), ]

dotplot_gex_crcpmp_tx_macrophages_colchemotaxisreceptors_ggplotobj <- data.frame(FeatureID = chemotaxis_receptors, crcpmp_tx_macrophages_chemotaxisreceptors_gex) %>%
  tidyr::pivot_longer(-c(FeatureID), names_to = "CB", values_to = "nUMIs") %>%
  dplyr::mutate(CB = stringr::str_replace(CB, "\\.", "-")) %>%
  dplyr::inner_join(data.frame(CB = rownames(seurat_crcpmp_tx_macrophages@meta.data), 
                               Celltype = seurat_crcpmp_tx_macrophages@meta.data[,"manual_l4"]), 
                    by = "CB") %>%
  dplyr::mutate(Celltype = factor(Celltype), 
                FeatureID = factor(FeatureID)) %>%
  dplyr::group_by(Celltype, FeatureID, .drop = F) %>%
  dplyr::summarize(Median = median(log1p(nUMIs)),
                   Percentage = mean(nUMIs>0)*100) %>%
  dplyr::mutate(Median = ifelse(is.na(Median), 0, Median),
                Percentage = ifelse(is.na(Percentage), 0, Percentage),
                Celltype = factor(Celltype, celltype_order_l4_macrophages$celltype),
                FeatureID = factor(FeatureID, rev(chemotaxis_receptors))) %>% 
  ggplot(aes(x = Celltype, y = FeatureID, col = Median)) +
  geom_tile(alpha = 0, col = "grey") +
  geom_point(aes(size = Percentage)) +
  scale_color_continuous(low = "#132B43", high = "#56B1F7") + 
  theme_bw() +
  theme(legend.position = "bottom", 
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.spacing.y=unit(0, "lines"))

pdf(file.path(fig3Dir, "dotplot_gex_crcpmp_tx_macrophages_colchemotaxisreceptors.pdf"), width = 1.7, height = 4.5)
print(dotplot_gex_crcpmp_tx_macrophages_colchemotaxisreceptors_ggplotobj)
dev.off()
```

## Radarplot PM-CRC TX Macrophages median TAM classification

```{r radarplot crcpmp tx macrophages median tam classification, fig.width = 6, fig.height = 5}
radarplot_crcpmp_tx_macrophages_mediantamclassification_ggplotobj <- crcpmp_pf_tx_paired_macrophages_tamannotation %>%
  dplyr::select(c("CellID", "SampleID", "Donor", "Tissue", "manual_l1", "manual_l2", "manual_l3", "manual_l4", "Angio", "IFN", "INFLAM", "LA", "Prolif", "Reg", "RTM")) %>%
  tidyr::pivot_longer(-c("CellID", "SampleID", "Donor", "Tissue", "manual_l1", "manual_l2", "manual_l3", "manual_l4"), names_to = "TAM", values_to = "UCell") %>%
  dplyr::filter(Tissue == "TX") %>%
  dplyr::group_by(manual_l4, TAM) %>%
  dplyr::summarize(median_UCell = median(UCell)) %>%
  dplyr::mutate(TAM = factor(TAM, levels = c("Angio", "IFN", "INFLAM", "LA", "Prolif", "Reg", "RTM"))) %>%
  ggplot(aes(x = TAM, y = median_UCell, col = manual_l4, group = manual_l4)) +
  geom_point() +
  geom_encircle(expand = 0, s_shape=1, size = 2) +
  scale_color_manual(values = manual_l4_macrophages_colors) +
  labs(y = "Median UCell Score") +
  coord_polar() +
  theme_bw() +
  theme(legend.title = element_blank(),
        axis.title.x = element_blank())

pdf(file.path(fig3Dir, "radarplot_crcpmp_pf_tx_macrophages_mediantamclassification.pdf"), width = 5, height=5.5)
print(radarplot_crcpmp_tx_macrophages_mediantamclassification_ggplotobj)
dev.off()
```

### Dotplot GEX PM-CRC TX macrophages TAM markers

```{r dotplot gex crcpmp tx col_tammarkers, fig.width = 7, fig.height=2.75}
seurat_crcpmp_tx_macrophages_tamgenes <- subset(seurat_crcpmp_tx_macrophages, features = unique(tam_markers$Gene))
Idents(seurat_crcpmp_tx_macrophages_tamgenes) <- "manual_l4"

seurat_crcpmp_tx_macrophages_tamgenes_markergenes <- FindAllMarkers(seurat_crcpmp_tx_macrophages_tamgenes, only.pos = T)
seurat_crcpmp_tx_macrophages_tamgenes_markergenes <- split(seurat_crcpmp_tx_macrophages_tamgenes_markergenes, seurat_crcpmp_tx_macrophages_tamgenes_markergenes$cluster)
seurat_crcpmp_tx_macrophages_tamgenes_markergenes <- seurat_crcpmp_tx_macrophages_tamgenes_markergenes[c("Macrophages SPP1+", "Macrophages VCAN+C1Q+", "Macrophages C1Q+", "Macrophages VCAN+")]

tammarkergenes <- unique(unlist(lapply(seurat_crcpmp_tx_macrophages_tamgenes_markergenes, function(macrophage){
  macrophage$gene[1:5]
})))

macrophages_tammarkergenes_gex <- GetAssayData(seurat_crcpmp_tx_macrophages, assay = "RNA")[which(rownames(GetAssayData(seurat_crcpmp_tx_macrophages, assay = "RNA")) %in% tammarkergenes), ]

dotplot_crcpmp_tx_macrophages_coltammarkers_ggplotobj <- data.frame(
  FeatureID = rownames(macrophages_tammarkergenes_gex), macrophages_tammarkergenes_gex) %>%
  tidyr::pivot_longer(-c(FeatureID), names_to = "CB", values_to = "nUMIs") %>%
  dplyr::mutate(CB = stringr::str_replace(CB, "\\.", "-")) %>%
  dplyr::inner_join(data.frame(CB = rownames(seurat_crcpmp_tx_macrophages@meta.data), 
                               Celltype = seurat_crcpmp_tx_macrophages@meta.data[,"manual_l4"]), 
                    by = "CB") %>%
  dplyr::mutate(Celltype = factor(Celltype), 
                FeatureID = factor(FeatureID)) %>%
  dplyr::group_by(Celltype, FeatureID, .drop = F) %>%
  dplyr::summarize(Median = median(log1p(nUMIs)),
                   Percentage = mean(nUMIs>0)*100) %>%
  dplyr::mutate(Median = ifelse(is.na(Median), 0, Median),
                Percentage = ifelse(is.na(Percentage), 0, Percentage),
                Celltype = factor(Celltype, c("Macrophages SPP1+", "Macrophages VCAN+C1Q+", "Macrophages C1Q+", "Macrophages VCAN+")),
                FeatureID = factor(FeatureID, rev(tammarkergenes[which(tammarkergenes %in% FeatureID)]))) %>% 
  ggplot(aes(x = FeatureID, y = Celltype, col = Median)) +
  geom_tile(alpha = 0, col = "grey") +
  geom_point(aes(size = Percentage)) +
  scale_color_viridis() +
  theme_bw() +
  theme(legend.position = "bottom", 
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.spacing.x=unit(0, "lines"))

pdf(file.path(fig3Dir, "dotplot_crcpmp_tx_macrophages_coltammarkers.pdf"), width = 7, height=2.75)
print(dotplot_crcpmp_tx_macrophages_coltammarkers_ggplotobj)
dev.off()
```
