---
title: 'Figure 3: PerIS HC vs PM-CRC'
author: "Andrew Y.F. Li Yim"
date: '2023-07-31'
output: html_document
---

```{r libraries, include=FALSE}
library(Seurat)
library(DESeq2)
library(ggplot2)
library(dplyr)
library(ggrastr)
library(ggrepel)
library(ggpubr)
library(viridis)
library(ggblend)
library(SummarizedExperiment)
library(msigdbr)
```

If the snakemake pipeline was run as-is, the following files will be generated in the folder from which the pipeline was run. If necessary, adjust paths accordingly.

## Setup

### Paths

```{r paths}
base_path <- "/mnt/smb/ltytgat/ayliyim/MI-group/ayliyim/projects/TIPRJ0000015_MAPS/maps_crc" # 

# Subsets
seurat_hc_crcpmp_pf_allcells_rds <- file.path(base_path, "output/q2_crc_vs_hc/subsets/hc_crcpmp_pf_allcells_SeuratObject.Rds")
seurat_hc_crcpmp_pf_rds <- file.path(base_path, "output/q2_crc_vs_hc/subsets/hc_crcpmp_pf_SeuratObject.Rds")
seurat_hc_crcpmp_pf_macrophages_rds <- file.path(base_path, "output/q2_crc_vs_hc/subsets/hc_crcpmp_pf_macrophages_SeuratObject.Rds")
seurat_hc_crcpmp_gcpmp_pf_macrophages_rds <- file.path(base_path, "resources/gastric_cancer/hc_crcpmp_gcpmp_pf_macrophages_SeuratObject.Rds") 

# Analyses
degs_pf_l4_crcpmpvhc_rds <- file.path(base_path, "output/q2_crc_vs_hc/analyses/degs_crcvhc_PF_manual_l4_list.Rds")

# Plotting parameters
celltype_markers_xlsx <- file.path(base_path, "config/order/celltype_markers.xlsx")
```

### Import data

```{r load subsets}
seurat_hc_crcpmp_pf_allcells <- readRDS(seurat_hc_crcpmp_pf_allcells_rds)
seurat_hc_crcpmp_pf <- readRDS(seurat_hc_crcpmp_pf_rds)
seurat_hc_crcpmp_pf_macrophages <- readRDS(seurat_hc_crcpmp_pf_macrophages_rds)
seurat_hc_crcpmp_gcpmp_pf_macrophages <- readRDS(seurat_hc_crcpmp_gcpmp_pf_macrophages_rds)
```

```{r load de analyses}
degs_pf_l4_crcpmpvhc <- readRDS(degs_pf_l4_crcpmpvhc_rds)
```

### Output

```{r figures}
fig3Dir <- file.path(base_path, "docs/manuscript/figures/fig3")
dir.create(fig3Dir, recursive = T)
```

### Plotting parameters

```{r plotting order}
celltype_order_pf_l1l2 <- readxl::read_excel(celltype_markers_xlsx, col_names = T) %>%
  dplyr::filter(level == "manual_l1",
                modality == "gene") %>%
  dplyr::select(celltype, color) %>%
  unique() %>%
  dplyr::filter(celltype %in% unique(seurat_hc_crcpmp_pf_allcells@meta.data[,"manual_l1"]),
                !celltype %in% c("Mesenchymal", "Myeloid")) %>%
  dplyr::rows_append(data.frame(celltype = "Mono-macs", 
                                color = "#E31A1C")) %>%
  dplyr::rows_append(data.frame(celltype = "DCs", 
                                color = "#FB9A99")) %>% 
  dplyr::rows_append(data.frame(celltype = "Granulocytes", 
                                color = "#33A02C")) %>%
  dplyr::arrange(factor(celltype, levels = c("T", "NK/ILC", "B", "Mono-macs", "DCs", "Granulocytes", "Erythroblasts", "Epithelial", "(Myo)Fibroblasts"))) %>%
  # dplyr::rows_append(data.frame(celltype = "DCs", 
  #                               color = "#FDBF6F")) %>%
  unique() %>%
  dplyr::mutate(number_subset = paste0(1:nrow(.), ". ", celltype)) %>%
  dplyr::rows_append(data.frame(celltype = c("(Myo)Fibroblasts"), 
                                color = c("#9ACD32"), 
                                number_subset = c("10. (Myo)Fibroblasts")))

celltype_order_pf_macrophages_l4 <- readxl::read_excel(celltype_markers_xlsx, col_names = T) %>%
  dplyr::filter(level == "manual_l4",
                modality == "gene",
                celltype %in% c("Macrophages VCAN+", "Macrophages C1Q+", "Macrophages VCAN+C1Q+", "Macrophages SPP1+", "Macrophages SPP1+")) %>%
  dplyr::select(celltype, color) %>%
  unique() %>%
  dplyr::filter(celltype %in% unique(seurat_hc_crcpmp_pf_macrophages@meta.data[,"manual_l4"])) %>%
  dplyr::mutate(number_subset = paste0(1:nrow(.), ". ", celltype))
```

```{r colors}
tissue_colors <- c(PF = "#224FBD", PBMC = "#F30000")
group_colors <- c(`HC` = "#93CEC1", `CRC+` = "#996633")

manual_l1l2_colors_pf <- celltype_order_pf_l1l2$color
names(manual_l1l2_colors_pf) <- celltype_order_pf_l1l2$celltype

manual_l4_colors_pf_macrophages <- celltype_order_pf_macrophages_l4$color
names(manual_l4_colors_pf_macrophages) <- celltype_order_pf_macrophages_l4$celltype

manual_l4_number_colors_pf_macrophages <- celltype_order_pf_macrophages_l4$color
names(manual_l4_number_colors_pf_macrophages) <- celltype_order_pf_macrophages_l4$number_subset
```

## Figures

### UMAP HC PM-CRC PF colored by l4

Here we show the unsupervised clustering analysis using UMAP of PF-derived macrophages obtained from PM-CRC patients annotated using the celltype.

```{r umap hc crcpmp pf macrophages col_l4, fig.width=10, fig.height=5}
umap_hc_crcpmp_pf_macrophages_coll4_ggplotobj <- data.frame(CB = colnames(seurat_hc_crcpmp_pf_macrophages),
                                                            Embeddings(seurat_hc_crcpmp_pf_macrophages[["umap"]]),
                                                            seurat_hc_crcpmp_pf_macrophages@meta.data) %>%
  dplyr::mutate(Group = factor(Group, levels = c("HC", "CRC+")),
                celltype = factor(manual_l4, levels = celltype_order_pf_macrophages_l4$celltype),
                celltype_number = as.numeric(celltype),
                celltype_w_number = paste0(as.numeric(celltype), ". ", celltype),
                celltype_w_number = factor(celltype_w_number)) %>% 
  ggplot(aes(x = UMAP_1, y = UMAP_2)) +
  geom_point_rast(show.legend = T, size = 2, aes(col = celltype_w_number)) +
  labs(y = "",
       x = "") +
  geom_label_repel(data = . %>%
                     dplyr::group_by(celltype, celltype_number, celltype_w_number) %>%
                     summarize(x = median(x = UMAP_1),
                               y = median(x = UMAP_2)),
                   mapping = aes(label = celltype_number, x = x, y = y),
                   alpha = 1, 
                   size = 8,
                   show.legend = F,
                   label.size = 0.25,
  ) +
  guides(colour = guide_legend(override.aes = list(size = 3))) +
  scale_color_manual(values = manual_l4_number_colors_pf_macrophages) +
  facet_wrap(~Group, nrow = 1) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.x = element_text(face = "bold"))

pdf(file.path(fig3Dir, "umap_hc_crcpmp_pf_macrophages_coll4.pdf"), width = 7.5, height=3.75)
print(umap_hc_crcpmp_pf_macrophages_coll4_ggplotobj)
dev.off()
```

### Stackedbarplot PM-CRC PF macrophages colored by l4

Here we show a stacked barplot of the PF-derived macrophage subsets relative to the total PF-derived macrophage population  obtained from HC, and PM-CRC patients separately.

```{r stackedbarplot crcpmp pf macrophages col_l4, fig.width=3, fig.height=5}
stackedbarplot_crcpmp_pf_macrophages_coll4_ggplotobj <- seurat_hc_crcpmp_pf_macrophages@meta.data %>%
  dplyr::group_by(manual_l4, Group, .drop = T) %>% 
  dplyr::summarise(Ncells = n()) %>%
  dplyr::ungroup() %>%
  tidyr::complete(tidyr::nesting(Group), tidyr::nesting(manual_l4), fill = list(Ncells = 0)) %>%
  dplyr::group_by(Group) %>%
  dplyr::mutate(manual_l4 = factor(manual_l4, levels = celltype_order_pf_macrophages_l4$celltype),
                Nsample = sum(Ncells),
                Ncellprop = Ncells/Nsample,
                Ncellprop = ifelse(is.na(Ncellprop), 0, Ncellprop),
                Ncellperc = Ncellprop*100,
                Group = factor(Group, levels = c("HC", "CRC+"))) %>%
  dplyr::rename(Celltype = manual_l4) %>%
  ggplot(aes(x = Group, y = Ncellperc)) +
  geom_bar(position="stack", stat="identity", aes(fill = Celltype)) +
  labs(y = "%Macrophages") +
  scale_fill_manual(values = manual_l4_colors_pf_macrophages) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "bottom",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

pdf(file.path(fig3Dir, "stackedbarplot_crcpmp_pf_macrophages_coll4.pdf"), width = 3, height = 5)
print(stackedbarplot_crcpmp_pf_macrophages_coll4_ggplotobj)
dev.off()

stackedbarplot_crcpmp_pf_macrophages_coll4_ggplotobj <- seurat_hc_crcpmp_pf_macrophages@meta.data %>%
  dplyr::group_by(manual_l4, Group, .drop = T) %>% 
  dplyr::summarise(Ncells = n()) %>%
  dplyr::ungroup() %>%
  tidyr::complete(tidyr::nesting(Group), tidyr::nesting(manual_l4), fill = list(Ncells = 0)) %>%
  dplyr::group_by(Group) %>%
  dplyr::mutate(manual_l4 = factor(manual_l4, levels = celltype_order_pf_macrophages_l4$celltype),
                Nsample = sum(Ncells),
                Ncellprop = Ncells/Nsample,
                Ncellprop = ifelse(is.na(Ncellprop), 0, Ncellprop),
                Ncellperc = Ncellprop*100,
                Group = factor(Group, levels = c("HC", "CRC+"))) %>%
  dplyr::rename(Celltype = manual_l4) %>%
  readr::write_csv(file.path(fig3Dir, "stackedbarplot_crcpmp_pf_macrophages.csv"))
```

### Heatmap pathways PM-CRC vs HC PF macrophages

Here we show a heatmap of the depicting the effect size (Wald statistic) for differentially expressed genes from selected differentially expressed KEGG gene sets, where we compare all macrophage subsets from PM-CRC with their counterpart from HC of the PF-derived macrophages.

```{r heatmap gex crcpmpvhc pf macrophages split_dpwoi, fig.width = 3, fig.height = 5.5}
pwoi_for_genes <- c("KEGG_CHEMOKINE_SIGNALING_PATHWAY",
                    "KEGG_TGF_BETA_SIGNALING_PATHWAY",
                    "KEGG_CYTOKINE_CYTOKINE_RECEPTOR_INTERACTION",
                    "KEGG_ANTIGEN_PROCESSING_AND_PRESENTATION",
                    "KEGG_COLORECTAL_CANCER",
                    "KEGG_PATHWAYS_IN_CANCER")

gs_hs <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:KEGG")
gs_hs_list <- split(x = gs_hs$gene_symbol, f = gs_hs$gs_name)
genes_pwoi <- unique(unlist(gs_hs_list[pwoi_for_genes]))

genes_pwoi_de <- lapply(degs_pf_l4_crcpmpvhc[celltype_order_pf_macrophages_l4$celltype], function(macrophage){
  macrophage$degs %>%
    data.frame() %>%
    dplyr::filter(padj<0.05 & gene %in% genes_pwoi)
})

genes_pwoi_de_sig <- unique(do.call(rbind, genes_pwoi_de)$gene)

genes_pwoi_bin <- data.frame(do.call(cbind, lapply(gs_hs_list[pwoi_for_genes], function(pwoi){genes_pwoi_de_sig %in% pwoi})))
rownames(genes_pwoi_bin) <- genes_pwoi_de_sig
colnames(genes_pwoi_bin) <- pwoi_for_genes

pwoi_pw <- data.frame(gene = rownames(genes_pwoi_bin), 
                      pathway = apply(genes_pwoi_bin, MARGIN = 1, function(gene){colnames(genes_pwoi_bin)[min(which(gene))]}))

heatmap_gex_crcpmpvhc_pf_macrophages_dpwois_ggplotobj <- do.call(rbind, lapply(celltype_order_pf_macrophages_l4$celltype, function(macrophage){
  degs_pf_l4_crcpmpvhc[[macrophage]]$degs %>%
    data.frame() %>%
    dplyr::filter(gene %in% genes_pwoi_de_sig) %>%
    dplyr::select(stat, pvalue, padj, gene) %>%
    dplyr::mutate(macrophages = macrophage)
})) %>%
  dplyr::mutate(direction = ifelse(stat<0, "HC", "CRC+"),
                significance = ifelse(pvalue<0.05, "Significant", "NS"),
                macrophages = factor(macrophages, levels = celltype_order_pf_macrophages_l4$celltype)) %>%
  dplyr::left_join(pwoi_pw, by = "gene") %>%
  ggplot(aes(x = macrophages, y = gene)) +
  geom_tile(aes(fill = stat, alpha = significance)) +
  theme_bw() +
  facet_grid(pathway~., space = "free", scales = "free") +
  scale_fill_gradient2(low = group_colors["HC"],
                       mid = "white",
                       high = group_colors["CRC+"],
                       midpoint = 0) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "right",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

pdf(file.path(fig3Dir, "heatmap_gex_crcpmpvhc_pf_macrophages_dpwois.pdf"), width = 3, height = 9)
print(heatmap_gex_crcpmpvhc_pf_macrophages_dpwois_ggplotobj)
dev.off()
```

### Boxplot HC and PM-CRC PF pseudobulk differential expression of macrophage subsets at l4

Here we prepare the data for scattered boxplots comparing the HC- and PM-CRC PF-derived macrophages at l4. Actual data was plotted in Graphpad Prism.

```{r crcpmpvhc pf macrophages l4 data}
macrophages <- c("Macrophages VCAN+", "Macrophages C1Q+", "Macrophages VCAN+C1Q+", "Macrophages SPP1+")

macrophage_genes_of_interest <- c("HLA-DRB1", "HLA-C", "IL10", "VEGFA", "IL6", "TNF", "IFNG", "IL1B")
macrophage_genes_of_interest <- c("HLA-DRB1", "HLA-C", "IL10", "VEGFA", "IL6", "TNF", "IFNG")

lapply(macrophages, function(macrophage){
  data.frame(Donor = colData(degs_pf_l4_crcpmpvhc[[macrophage]]$dds)$Donor,
             Group = colData(degs_pf_l4_crcpmpvhc[[macrophage]]$dds)$Group,
             t(assay(rlog(degs_pf_l4_crcpmpvhc[[macrophage]]$dds)[rownames(degs_pf_l4_crcpmpvhc[[macrophage]]$dds) %in% macrophage_genes_of_interest,])),
             Macrophage = macrophage) %>%
    tidyr::pivot_longer(-c("Donor", "Group", "Macrophage"), names_to = "Gene", values_to = "Expression") %>%
    dplyr::mutate(Gene = gsub("\\.", "-", Gene)) %>%
    readr::write_csv(file.path(fig3Dir, paste0("crcpmpvhc_pf_", macrophage, "_exprs.csv")))
  
  degs_pf_l4_crcpmpvhc[[macrophage]]$degs %>%
    data.frame() %>% 
    dplyr::filter(gene %in% macrophage_genes_of_interest) %>%
    readr::write_csv(file.path(fig3Dir, paste0("crcpmpvhc_pf_", macrophage, "_de.csv")))
})

```

### UMAP HC PM-CRC PF all cells colored by l1/l2

Here we show the unsupervised clustering analysis using UMAP of all PF-derived cells obtained from HC and PM-CRC patients annotated using the celltype.

```{r umap hc crcpmp pf allcells col_l1l2, fig.width=8, fig.height=7.5}
umap_hc_crcpmp_pf_allcells_coll1l2_ggplotobj <- data.frame(CB = colnames(seurat_hc_crcpmp_pf_allcells),
                                                           Embeddings(seurat_hc_crcpmp_pf_allcells[["umap"]]),
                                                           seurat_hc_crcpmp_pf_allcells@meta.data) %>%
  dplyr::mutate(celltype = manual_l1l2,
                celltype = factor(celltype, levels = celltype_order_pf_l1l2$celltype),
                celltype_number = as.numeric(celltype),
                celltype_w_number = paste0(as.numeric(celltype), ". ", celltype),
                celltype_w_number = factor(celltype_w_number, levels = celltype_order_pf_l1l2$number_subset)) %>% 
  ggplot(aes(x = UMAP_1, y = UMAP_2)) +
  geom_point_rast(show.legend = T, size = 0.25, aes(col = celltype)) +
  labs(y = "",
       x = "") +
  geom_label_repel(data = . %>%
                     dplyr::group_by(celltype, celltype_number, celltype_w_number) %>%
                     summarize(x = median(x = UMAP_1),
                               y = median(x = UMAP_2)),
                   mapping = aes(label = celltype, x = x, y = y),
                   size = 9,
                   alpha = 1,
                   show.legend = F) +
  guides(colour = guide_legend(override.aes = list(size = 3))) +
  scale_color_manual(values = manual_l1l2_colors_pf) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.x = element_text(face = "bold"))

pdf(file.path(fig3Dir, "umap_hc_crcpmp_pf_allcells_coll1l2.pdf"), width = 7.5, height = 7.5)
print(umap_hc_crcpmp_pf_allcells_coll1l2_ggplotobj)
dev.off()
```

### UMAP HC PM-CRC PF all cells colored by immunosuppressive genes

Here we characterize the PF-derived cells by superimposing the gene expression of the following genes (GEX) on the respective UMAP configuration:
- Immunosuppression (IL10)
- Immunosuppression (VEGFA)

```{r umap_hc_crcpmp_pf_allcells_colimmunosuppressivemarkers, fig.width = 10, fig.height = 5.5}
umap_hc_crcpmp_pf_allcells_colimmunosuppressivemarkers_ggplotobj_list <- lapply(
  data.frame(CB = colnames(seurat_hc_crcpmp_pf_allcells),
             Embeddings(seurat_hc_crcpmp_pf_allcells[["umap"]]),
             seurat_hc_crcpmp_pf_allcells@meta.data,
             expr = GetAssayData(seurat_hc_crcpmp_pf_allcells, layers = "data")["IL10",],
             Feature = "IL10") %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_hc_crcpmp_pf_allcells),
                                  Embeddings(seurat_hc_crcpmp_pf_allcells[["umap"]]),
                                  seurat_hc_crcpmp_pf_allcells@meta.data,
                                  expr = GetAssayData(seurat_hc_crcpmp_pf_allcells, layers = "data")["VEGFA",],
                                  Feature = "VEGFA")) %>%
    dplyr::mutate(expr_rank = rank(expr, ties.method="first"),
                  Feature = as.factor(Feature)) %>%
    split(., .$Feature), FUN = function(plotdf){
      plotdf %>% dplyr::arrange(expr) %>%
        ggplot(aes(x = UMAP_1, y = UMAP_2)) +
        geom_point_rast(show.legend = F, size = 0.25, col = "#d3d3d3") +
        geom_point_rast(data = . %>%
                          dplyr::filter(expr>0), aes(col = expr), show.legend = T) +
        guides(colour = guide_legend(override.aes = list(size = 3))) +
        facet_wrap(~Feature) +
        guides(colour = guide_legend(override.aes = list(size = 3),
                                     title = "Expression")) +
        scale_color_viridis() +
        theme_bw() +
        theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              legend.position = "bottom",
              axis.text.x = element_blank(),
              axis.text.y = element_blank(),
              axis.ticks = element_blank(),
              axis.title.x = element_blank(),
              axis.title.y = element_blank(),
              strip.text.x = element_text(face = "bold"))
    }
)

umap_hc_crcpmp_pf_allcells_colimmunosuppressivemarkers_ggplotobj <- ggarrange(plotlist = umap_hc_crcpmp_pf_allcells_colimmunosuppressivemarkers_ggplotobj_list, nrow = 1, ncol = 2)

pdf(file.path(fig3Dir, "umap_hc_crcpmp_pf_allcells_colimmunosuppressivemarkers.pdf"), width = 6, height = 3.5)
print(umap_hc_crcpmp_pf_allcells_colimmunosuppressivemarkers_ggplotobj)
dev.off()
```
