---
title: 'Figure 4: TIME of PM-CRC'
author: "Andrew Y.F. Li Yim"
date: '2023-07-31'
output: html_document
---

```{r libraries, include=FALSE}
library(Seurat)
library(ggplot2)
library(dplyr)
library(ggrastr)
library(ggrepel)
library(ggpubr)
library(viridis)
library(DESeq2)
library(ggalt)
library(ComplexHeatmap)
```

If the snakemake pipeline was run as-is, the following files will be generated in the folder from which the pipeline was run. If not, then adjust paths accordingly.    

## Setup

### Paths

```{r paths}
base_path <- "/mnt/smb/ltytgat/ayliyim/MI-group/ayliyim/projects/TIPRJ0000015_MAPS/maps_crc"

# Subsets
seurat_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages_rds <- file.path(base_path, "output/q3_pm_tx_characterization/subsets/hc_crcpmp_pbmc_pf_tx_paired_monocytes_macrophages_SeuratObject.Rds")
seurat_crcpmp_tx_immune_rds <- file.path(base_path, "output/q3_pm_tx_characterization/subsets/crcpmp_tx_immune_SeuratObject.Rds")
seurat_crcpmp_tx_monocytes_macrophages_rds <- file.path(base_path, "output/q3_pm_tx_characterization/subsets/crcpmp_tx_monocytes_macrophages_SeuratObject.Rds")
seurat_crcpmp_tx_macrophages_rds <- file.path(base_path, "output/q3_pm_tx_characterization/subsets/crcpmp_tx_macrophages_SeuratObject.Rds")

# Analyses
crcpmp_pf_tx_paired_macrophages_tamannotation_rds <- file.path(base_path, "output/q3_pm_tx_characterization/analyses/crcpmp_pf_tx_paired_macrophages_tamannotation.Rds")

# Plotting parameters
celltype_markers_xlsx <- file.path(base_path, "config/order/celltype_markers.xlsx")

# Genes of interest
tam_markers_xlsx <- file.path(base_path, "config/genes_of_interest/tams_ma_2022.xlsx")
azizi2018_xlsx <- file.path(base_path, "config/genes_of_interest/m1_m2_azizi2018.xlsx")
```

### Import data

```{r load subsets}
seurat_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages <- readRDS(seurat_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages_rds)
seurat_crcpmp_tx_monocytes_macrophages <- readRDS(seurat_crcpmp_tx_monocytes_macrophages_rds)
seurat_crcpmp_tx_macrophages <- readRDS(seurat_crcpmp_tx_macrophages_rds)
seurat_crcpmp_tx_immune <- readRDS(seurat_crcpmp_tx_immune_rds)
```

```{r load macrophage markers}
m1_markers <- readxl::read_excel(azizi2018_xlsx, sheet = 2)
m2_markers <- readxl::read_excel(azizi2018_xlsx, sheet = 3)
```

```{r load tam markers}
tam_markers <- readxl::read_excel(tam_markers_xlsx)
crcpmp_pf_tx_paired_macrophages_tamannotation <- readRDS(crcpmp_pf_tx_paired_macrophages_tamannotation_rds)
```

### Output

```{r figures}
fig4Dir <- file.path(base_path, "docs/manuscript/figures/fig4")
dir.create(fig4Dir, recursive = T)
```

### Plotting parameters

```{r plotting order}
pf_heatmap_order_xlsx <- file.path(base_path, "config/order/q1_pf_heatmap_order.xlsx")

celltype_order_l2_monocytes_macrophages <- readxl::read_excel(celltype_markers_xlsx, col_names = T) %>%
  dplyr::filter(level == "manual_l2",
                modality == "gene",
                celltype %in% c("Monocytes", "Macrophages")) %>%
  dplyr::select(celltype, color) %>%
  unique()

celltype_order_l3_tx_immune <- readxl::read_excel(celltype_markers_xlsx, col_names = T) %>%
  dplyr::filter(level == "manual_l3",
                modality == "gene") %>%
  dplyr::select(celltype, color) %>%
  unique() %>%
  dplyr::filter(!is.na(color),
                celltype %in% seurat_crcpmp_tx_immune$manual_l3) %>%
  dplyr::mutate(number_subset = paste0(1:nrow(.), ". ", celltype))

celltype_order_l4_monocytes_macrophages <- readxl::read_excel(celltype_markers_xlsx, col_names = T) %>%
  dplyr::filter(level == "manual_l4",
                celltype %in% c("Classical monocytes", "Non-classical monocytes", "Macrophages VCAN+", "Macrophages C1Q+", "Macrophages VCAN+C1Q+", "Macrophages SPP1+"),
                modality == "gene") %>%
  dplyr::select(celltype, color) %>%
  unique() %>%
  dplyr::filter(!is.na(color)) %>%
  dplyr::mutate(number_subset = paste0(1:nrow(.), ". ", celltype))

celltype_order_l4_macrophages <- readxl::read_excel(celltype_markers_xlsx, col_names = T) %>%
  dplyr::filter(level == "manual_l4",
                celltype %in% c("Macrophages C1Q+", "Macrophages VCAN+", "Macrophages VCAN+C1Q+", "Macrophages SPP1+"),
                modality == "gene") %>%
  dplyr::select(celltype, color) %>%
  unique() %>%
  dplyr::filter(!is.na(color)) %>%
  dplyr::mutate(number_subset = paste0(1:nrow(.), ". ", celltype))

gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}

```

```{r colors}
tissue_colors <- c(PBMC = "#F30000", PF = "#224FBD", TX = "#808080")
group_colors <- c(`HC` = "#93CEC1", `CRC+` = "#996633")
tissue_group_colors <- c("PBMC HC" = "#58508d", "PBMC CRC+" = "#bc5090", "PF HC" = "#ff6361", "PF CRC+" = "#ffa600", "TX CRC+" = "#003f5c")
donor_colors <- c(pt31 = "#e6194b", pt35 = "#3cb44b", pt37 = "#ffe119", pt71 = "#4363d8", pt73 = "#f58231", pt74 = "#911eb4", pt76 = "#46f0f0", pt78 = "#f032e6", pt87 = "#bcf60c", pt88 = "#fabebe", pt89 = "#008080", pt91 = "#e6beff", pt92 = "#9a6324")

manual_l2_monocytes_macrophages_colors <- celltype_order_l2_monocytes_macrophages$color
names(manual_l2_monocytes_macrophages_colors) <- celltype_order_l2_monocytes_macrophages$celltype

manual_l3_tx_immune_colors <- celltype_order_l3_tx_immune$color
names(manual_l3_tx_immune_colors) <- celltype_order_l3_tx_immune$number_subset

manual_l4_monocytes_macrophages_number_colors <- celltype_order_l4_monocytes_macrophages$color
names(manual_l4_monocytes_macrophages_number_colors) <- celltype_order_l4_monocytes_macrophages$number_subset

manual_l4_macrophages_colors <- celltype_order_l4_macrophages$color
names(manual_l4_macrophages_colors) <- celltype_order_l4_macrophages$celltype
```

## Figures

### UMAP PM-CRC TX monocytes and macrophages colored by l4

Here we show the unsupervised clustering analysis using UMAP of CRC-PM patients PF-derived immune cells annotated using the celltype.

```{r umap crcpmp pf monocytes macrophages col_l4, fig.width=5, fig.height=5}
umap_crcpmp_tx_monocytes_macrophages_coll4_ggplotobj <- data.frame(CB = colnames(seurat_crcpmp_tx_monocytes_macrophages),
                                                                   Embeddings(seurat_crcpmp_tx_monocytes_macrophages[["umap"]]),
                                                                   seurat_crcpmp_tx_monocytes_macrophages@meta.data) %>%
  dplyr::mutate(Group = factor(Group, levels = c("HC", "CRC+")),
                celltype = factor(manual_l4, levels = celltype_order_l4_monocytes_macrophages$celltype),
                celltype_number = as.numeric(celltype),
                celltype_w_number = paste0(as.numeric(celltype), ". ", celltype),
                celltype_w_number = factor(celltype_w_number)) %>% 
  ggplot(aes(x = UMAP_1, y = UMAP_2)) +
  #geom_point_rast(show.legend = T, size = 0.5, col = "black") +
  geom_point_rast(show.legend = T, size = 3, aes(col = celltype_w_number)) +
  labs(y = "",
       x = "") +
  geom_label_repel(data = . %>%
                     dplyr::group_by(celltype, celltype_number, celltype_w_number) %>%
                     summarize(x = median(x = UMAP_1),
                               y = median(x = UMAP_2)),
                   mapping = aes(label = celltype_number, x = x, y = y),
                   alpha = 1, 
                   size = 8,
                   label.size=0.25,
                   show.legend = F) +
  guides(colour = guide_legend(override.aes = list(size = 3))) +
  scale_color_manual(values = manual_l4_monocytes_macrophages_number_colors) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.x = element_text(face = "bold"))

print(umap_crcpmp_tx_monocytes_macrophages_coll4_ggplotobj)

pdf(file.path(fig4Dir, "umap_crcpmp_tx_monocytes_macrophages_coll4.pdf"), width = 5, height = 5)
print(umap_crcpmp_tx_monocytes_macrophages_coll4_ggplotobj)
dev.off()
```

### Boxplot PM-CRC TX macrophage subsets percentage relative to all macrophages

Here we prepare the data for scattered boxplots depicting the TX-derived macrophage subsets relative to the total TX-derived macrophage population. Actual data was plotted in Graphpad Prism.

```{r crcpmp tx myeloid l3rl1 data}
seurat_crcpmp_tx_macrophages@meta.data %>%
  dplyr::group_by(manual_l3, SampleID, .drop = T) %>% 
  dplyr::summarise(Ncells = n()) %>%
  dplyr::ungroup() %>%
  tidyr::complete(tidyr::nesting(SampleID), tidyr::nesting(manual_l3), fill = list(Ncells = 0)) %>%
  dplyr::group_by(SampleID) %>%
  dplyr::mutate(Nlrefsample = sum(Ncells),
                Ncellprop = Ncells/Nlrefsample,
                Ncellprop = ifelse(is.na(Ncellprop), 0, Ncellprop),
                Ncellperc = Ncellprop*100,
                Donor = gsub("^(pt[0-9]{2})_.+$", "\\1", SampleID)) %>%
  readr::write_csv(file = file.path(fig4Dir, "crcpmp_tx_macrophages_l3rl2.csv"))
```

### UMAP PM-CRC TX monocytes and macrophages colored by myeloid markers

Here we characterize the TX-derived monocytes and macrophages by superimposing the gene expression of the following genes on the respective UMAP configuration:
- Resident macrophage (VSIG4)

```{r umap crcpmp tx monocytes macrophages col_myeloidmarkers, fig.width=3, fig.height=3}
umap_crcpmp_tx_monocytes_macrophages_colmyeloidgenemarkers_ggplotobj <- data.frame(
  CB = colnames(seurat_crcpmp_tx_monocytes_macrophages),
  Embeddings(seurat_crcpmp_tx_monocytes_macrophages[["umap"]]),
  seurat_crcpmp_tx_monocytes_macrophages@meta.data,
  expr = GetAssayData(seurat_crcpmp_tx_monocytes_macrophages, assay = "RNA")["VSIG4",],
  Feature = "VSIG4") %>%
  dplyr::mutate(expr_rank = rank(expr, ties.method="first")) %>% 
  dplyr::arrange(expr) %>%
  ggplot(aes(x = UMAP_1, y = UMAP_2)) +
  geom_point_rast(show.legend = F, size = 0.25, col = "#d3d3d3") +
  geom_point_rast(data = . %>%
                    dplyr::filter(expr>0), aes(col = expr), show.legend = T) +
  guides(colour = guide_legend(override.aes = list(size = 3))) +
  facet_wrap(~Feature) +
  guides(colour = guide_legend(override.aes = list(size = 3),
                               title = "Expression")) +
  scale_color_viridis() +
  theme_bw() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        strip.text.x = element_text(face = "bold"))

pdf(file.path(fig4Dir, "umap_crcpmp_tx_monocytes_macrophages_colmyeloidgenemarkers.pdf"), width = 3, height = 3)
print(umap_crcpmp_tx_monocytes_macrophages_colmyeloidgenemarkers_ggplotobj)
dev.off()
```

### UMAP HC PM-CRC PBMC PF TX monocytes and macrophages colored by l2

Here we show the unsupervised clustering analysis using UMAP of HC and CRC-PM patients PBMC-, PF-, and TX-derived monocytes and macrophages annotated using the celltype.

```{r umap crcpmp pf monocytes macrophages col_l2, fig.width=5, fig.height=5}
umap_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages_coll2_ggplotobj <- data.frame(
  CB = colnames(seurat_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages),
  Embeddings(seurat_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages[["umap"]]),
  seurat_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages@meta.data) %>%
  ggplot(aes(x = umap_1, y = umap_2, col = manual_l2)) +
  geom_point_rast(show.legend = T, size = 1) +
  labs(y = "",
       x = "") +
  guides(colour = guide_legend(override.aes = list(size = 3))) +
  scale_color_manual(values = manual_l2_monocytes_macrophages_colors) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.x = element_text(face = "bold"))

pdf(file.path(fig4Dir, "umap_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages_coll2.pdf"), width = 7.5, height = 7.5)
print(umap_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages_coll2_ggplotobj)
dev.off()
```

### UMAP HC PM-CRC PBMC PF TX monocytes and macrophages colored by tissue_group

Here we show the unsupervised clustering analysis using UMAP of HC and CRC-PM patients PBMC-, PF-, and TX-derived monocytes and macrophages annotated using a combination of the group and celltype.

```{r umap hc crcpmp pbmc pf tx monocytes macrophages col_tissuegroup, fig.width=5, fig.height=5}
umap_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages_coltissuegroup_ggplotobj <- data.frame(
  CB = colnames(seurat_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages),
  Embeddings(seurat_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages[["umap"]]),
  seurat_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages@meta.data) %>%
  dplyr::mutate(Tissue_group = factor(paste0(Tissue, " ", Group), levels = c("PBMC HC", "PBMC CRC+", "PF HC", "PF CRC+", "TX CRC+"))) %>%
  ggplot(aes(x = umap_1, y = umap_2, col = Tissue_group)) +
  geom_point_rast(show.legend = T, size = 1) +
  labs(y = "",
       x = "") +
  guides(colour = guide_legend(override.aes = list(size = 3))) +
  scale_color_manual(values = tissue_group_colors) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.x = element_text(face = "bold"))

pdf(file.path(fig4Dir, "umap_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages_coltissuegroup.pdf"), width = 7.5, height = 7.5)
print(umap_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages_coltissuegroup_ggplotobj)
dev.off()
```

### Heatmap HC PM-CRC PBMC PF TX monocytes and macrophages

Here we show a dendrogram/heatmap of the HC and CRC-PM patients PBMC-, PF-, and TX-derived monocytes and macrophages to show the distance between each celltype using the Euclidian distance.

```{r heatmap HC PM-CRC PBMC PF TX monocytes and macrophages, fig.height = 12, fig.width = 12}
seurat_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages@meta.data$Tissue_group_donor_l2 <- paste0(seurat_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages@meta.data$Tissue, " ", seurat_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages@meta.data$Group, " ", seurat_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages@meta.data$Donor, " ", seurat_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages@meta.data$manual_l2)

counts_tissue_group <- GetAssayData(seurat_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages, assay = "RNA", layer = "counts") %*% model.matrix(~0+Tissue_group_donor_l2, data = seurat_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages@meta.data)
colnames(counts_tissue_group) <- gsub("Tissue_group_donor_l2", "", colnames(counts_tissue_group))

counts_tissue_group <- counts_tissue_group[which(Matrix::rowSums(counts_tissue_group) != 0),]

rld_tissue_group <- DESeq2::rlog(as.matrix(counts_tissue_group))
colnames(rld_tissue_group) <- gsub("TX", "PM", colnames(rld_tissue_group))
colnames(rld_tissue_group) <- gsub("CRC\\+", "PM-CRC", colnames(rld_tissue_group))

dist_rld_tissue_group <- dist(t(rld_tissue_group), method = "euclidean")
hclust_rld_tissue_group <- hclust(dist_rld_tissue_group)
dend_hclust_dist_rld_tissue_group <- as.dendrogram(hclust_rld_tissue_group)

metadata_tissue_group <- data.frame(do.call(rbind, strsplit(colnames(counts_tissue_group), split = " ")), row.names = colnames(counts_tissue_group)) %>%
  dplyr::rename(Tissue = X1,
                Group = X2,
                Donor = X3,
                Celltype = X4) %>%
  dplyr::mutate(Group = ifelse(Group == "CRC", "CRC+", Group))

heatmap_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages_coleucdist <- Heatmap(as.matrix(dist_rld_tissue_group), 
                  col = colorRampPalette(c("blue", "white", "red"))(10000), 
                  cluster_rows = dend_hclust_dist_rld_tissue_group, 
                  cluster_columns = dend_hclust_dist_rld_tissue_group,
                  top_annotation = HeatmapAnnotation(
                    Tissue = metadata_tissue_group$Tissue,
                    Group = metadata_tissue_group$Group,
                    Celltype = metadata_tissue_group$Celltype,
                    col = list(Tissue = tissue_colors,
                               Group = group_colors,
                               Celltype = c(Macrophages = "#F8766D", Monocytes = "#00BFC4")),
                    gp = gpar(col = "black")),
                  heatmap_legend_param = list(title = "Euclidian Distance"))

pdf(file.path(fig4Dir, "heatmap_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages_coleucdist.pdf"), width = 14, height = 12)
print(heatmap_hc_crcpmp_pbmc_pf_tx_monocytes_macrophages_coleucdist)
dev.off()

# Circular
phylo_hclust_rld_tissue_group <- as.phylo(hclust_rld_tissue_group)

dend_hclust_dist_rld_tissue_group_ggtree <- ggtree(phylo_hclust_rld_tissue_group, layout = "circular")
dend_hclust_dist_rld_tissue_group_ggtree <- gheatmap(dend_hclust_dist_rld_tissue_group_ggtree, metadata_tissue_group %>% 
           dplyr::select(Celltype),
           offset = 1, 
           width = .1,
           colnames_angle = 90, 
           colnames_offset_y = .35)
dend_hclust_dist_rld_tissue_group_ggtree <- gheatmap(dend_hclust_dist_rld_tissue_group_ggtree, metadata_tissue_group %>% 
           dplyr::select(Group),
           offset = 11, 
           width = .1,
           colnames_angle = 90, 
           colnames_offset_y = .25)
dend_hclust_dist_rld_tissue_group_ggtree <- gheatmap(dend_hclust_dist_rld_tissue_group_ggtree, metadata_tissue_group %>% 
           dplyr::select(Tissue),
           offset = 21, 
           width = .1,
           colnames_angle = 90, 
           colnames_offset_y = .15)
dend_hclust_dist_rld_tissue_group_ggtree <- gheatmap(dend_hclust_dist_rld_tissue_group_ggtree, metadata_tissue_group %>% 
           dplyr::select(Donor),
           offset = 31, 
           width = .1,
           colnames_angle = 90, 
           colnames_offset_y = .15) +
  scale_fill_manual(values = c(tissue_colors, group_colors, manual_l2_monocytes_macrophages_colors, donor_colors))

pdf(file.path(fig4Dir, "dendrogram_hclust_dist_rld_tissue_group_circular_ptcol.pdf"), width = 10, height = 10)
print(dend_hclust_dist_rld_tissue_group_ggtree)
dev.off()
```
