---
title: 'Supplementary Figure 4: Peritoneal immune system in PM-CRC patients'
author: "Andrew Y.F. Li Yim"
date: '2023-07-31'
output: html_document
---

```{r libraries, include=FALSE}
library(Seurat)
library(DESeq2)
library(ggplot2)
library(dplyr)
library(ggrastr)
library(ggrepel)
library(ggpubr)
library(viridis)
```

If the snakemake pipeline was run as-is, the following files will be generated in the folder from which the pipeline was run. If necessary, adjust paths accordingly.

## Setup

### Paths

```{r paths}
base_path <- "/mnt/smb/ltytgat/ayliyim/MI-group/ayliyim/projects/TIPRJ0000015_MAPS/maps_crc"

# Subsets
seurat_hc_crcpmp_pf_rds <- file.path(base_path, "output/q2_crc_vs_hc/subsets/hc_crcpmp_pf_SeuratObject.Rds")
seurat_hc_crcpmp_pf_macrophages_rds <- file.path(base_path, "output/q2_crc_vs_hc/subsets/hc_crcpmp_pf_macrophages_SeuratObject.Rds")

# Analyses
degs_pf_l4_crcpmpvhc_rds <- file.path(base_path, "output/q2_crc_vs_hc/analyses/degs_crcvhc_PF_manual_l4_list.Rds")
fgsea_pf_l4_crcpmpvhc_rds <- file.path(base_path, "output/q2_crc_vs_hc/analyses/fgsea_crcvhc_PF_manual_l4_list.Rds")

# Plotting parameters
celltype_markers_xlsx <- file.path(base_path, "config/order/celltype_markers.xlsx")
azizi2018_markers_xlsx <- file.path(base_path, "config/genes_of_interest/m1_m2_azizi2018.xlsx")
pf_heatmap_order_xlsx <- file.path(base_path, "config/order/q1_pf_heatmap_order.xlsx")
```

### Import data

```{r load subsets}
seurat_hc_crcpmp_pf <- readRDS(seurat_hc_crcpmp_pf_rds)
seurat_hc_crcpmp_pf_macrophages <- readRDS(seurat_hc_crcpmp_pf_macrophages_rds)
```

```{r load de analyses}
degs_pf_l4_crcpmpvhc <- readRDS(degs_pf_l4_crcpmpvhc_rds)
fgsea_pf_l4_crcpmpvhc <- readRDS(fgsea_pf_l4_crcpmpvhc_rds)
```

```{r load macrophage markers}
m1_markers <- readxl::read_excel(azizi2018_markers_xlsx, sheet = 2)
m2_markers <- readxl::read_excel(azizi2018_markers_xlsx, sheet = 3)
```

### Output

```{r figures}
figS4Dir <- file.path(base_path, "docs/manuscript/figures/figS4")
dir.create(figS4Dir, recursive = T)
```

### Plotting parameters

```{r plotting order}
celltype_order_pf_macrophages_l4 <- readxl::read_excel(celltype_markers_xlsx, col_names = T) %>%
  dplyr::filter(level == "manual_l4",
                modality == "gene",
                celltype %in% c("Macrophages VCAN+", "Macrophages C1Q+", "Macrophages VCAN+C1Q+", "Macrophages SPP1+", "Macrophages SPP1+")) %>%
  dplyr::select(celltype, color) %>%
  unique() %>%
  dplyr::filter(celltype %in% unique(seurat_hc_crcpmp_pf_macrophages@meta.data[,"manual_l4"])) %>%
  dplyr::mutate(number_subset = paste0(1:nrow(.), ". ", celltype))
```

```{r colors}
tissue_colors <- c(PF = "#224FBD", PBMC = "#F30000")
group_colors <- c(`HC` = "#93CEC1", `CRC+` = "#996633")

manual_l4_colors_pf_macrophages <- celltype_order_pf_macrophages_l4$color
names(manual_l4_colors_pf_macrophages) <- celltype_order_pf_macrophages_l4$celltype

manual_l4_number_colors_pf_macrophages <- celltype_order_pf_macrophages_l4$color
names(manual_l4_number_colors_pf_macrophages) <- celltype_order_pf_macrophages_l4$number_subset
```

## Figures

### UMAP HC PM-CRC PF macrophages blend by Donor split Group

```{r umap hc crcpmp pf coldonor splitgroup, fig.width=15, fig.height=7.5}
umap_hc_crcpmp_pf_coldonor_splitgroup_ggplotobj <- data.frame(CB = colnames(seurat_hc_crcpmp_pf),
                                                              Embeddings(seurat_hc_crcpmp_pf[["umap"]]),
                                                              seurat_hc_crcpmp_pf@meta.data) %>%
  dplyr::mutate(Group = factor(ifelse(Group == "CRC+", "CRC", "HC"), levels = c("HC", "CRC"))) %>%
  ggplot(aes(x = UMAP_1, y = UMAP_2, col = Donor, col = Donor)) +
  geom_point_rast(show.legend = T, size = 0.25)+
  labs(y = "",
       x = "") +
  guides(colour = guide_legend(override.aes = list(size = 3))) +
  facet_wrap(~Group) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.x = element_text(face = "bold"))

pdf(file.path(figS4Dir, "umap_hc_crcpmp_pf_coldonor_splitgroup.pdf"), width = 15, height=7.5)
print(umap_hc_crcpmp_pf_coldonor_splitgroup_ggplotobj)
dev.off()
```

### Heatmap PM-CRC vs HC PF macrophages M1/M2 markers colored by tstat

Here we visualize a the statistically significant genes belonging to the M1 and M2 macrophage markers that were statistically significantly different based on pseudobulk analysis in PF-derived macrophages when comparing PM-CRC and HC. 

```{r heatmap crcpmpvhc pf l4 m1m2gexmarkers col_tstat, fig.width = 3, fig.height = 7}
macrophage_markers <- data.frame(gene = c(m1_markers$HGNC, m2_markers$HGNC),
                                 set = c(rep("M1", length(m1_markers$HGNC)), rep("M2", length(m2_markers$HGNC))))

heatmap_crcpmpvhc_pf_l4_m1m2gexmarkers_ggplotobj <- do.call(rbind, lapply(celltype_order_pf_macrophages_l4$celltype, function(macrophage){
  degs_pf_l4_crcpmpvhc[[macrophage]]$degs %>%
    data.frame() %>%
    dplyr::filter(gene %in% macrophage_markers$gene) %>%
    dplyr::select(stat, pvalue, padj, gene) %>%
    dplyr::mutate(macrophages = macrophage)
})) %>%
  dplyr::mutate(direction = ifelse(stat<0, "HC", "CRC+"),
                significance = ifelse(pvalue<0.05, "Significant", "NS"),
                macrophages = factor(macrophages, levels = celltype_order_pf_macrophages_l4$celltype)) %>%
  dplyr::left_join(macrophage_markers, by = "gene") %>%
  dplyr::filter(pvalue<0.05) %>%
  ggplot(aes(x = macrophages, y = gene)) +
  geom_tile(aes(fill = stat)) +
  facet_grid(set~., space = "free", scales = "free") +
  theme_bw() +
  scale_fill_gradient2(low = group_colors["HC"],
                       mid = "white",
                       high = group_colors["CRC+"],
                       midpoint = 0) +
  scale_alpha_manual(values = c("NS" = "#d3d3d3", "Significant" = "#000000")) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "right",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

pdf(file.path(figS4Dir, "heatmap_crcpmpvhc_pf_l4_m1m2gexmarkers.pdf"), width = 3.5, height = 9)
print(heatmap_crcpmpvhc_pf_l4_m1m2gexmarkers_ggplotobj)
dev.off()
```

### Dotplot pathways PM-CRC vs HC PF macrophages colored by direction that were statistically significant

Here we visualize a subset of statistically significant pathways that were different in PF-derived macrophages when comparing PM-CRC and HC. Importantly, we define pathways of interest based on it being statistically significantly different in one or more macrophage subsets. 

```{r crcpmpvhc pathways of intereset}
crcpmpvhc_pf_macrophages_l4_fgsea_sig <- unique(unlist(lapply(fgsea_pf_l4_crcpmpvhc[celltype_order_pf_macrophages_l4$celltype], function(macrophage){
  macrophage %>%
    data.frame() %>%
    dplyr::select(1:7) %>%
    dplyr::filter(padj<0.05) %>%
    dplyr::pull(pathway)
})))

crcpmpvhc_pf_macrophages_l4_fgsea_sig_df <- do.call(rbind, lapply(celltype_order_pf_macrophages_l4$celltype, function(macrophage){
  fgsea_pf_l4_crcpmpvhc[[macrophage]] %>%
    data.frame() %>%
    dplyr::filter(pathway %in% crcpmpvhc_pf_macrophages_l4_fgsea_sig) %>%
    dplyr::select(pathway, NES, pval, padj) %>%
    dplyr::mutate(macrophages = macrophage)
}))
```

Here, we define the following significant pathways as interesting: 
- KEGG_OXIDATIVE_PHOSPHORYLATION
- KEGG_T_CELL_RECEPTOR_SIGNALING_PATHWAY
- KEGG_ANTIGEN_PROCESSING_AND_PRESENTATION
- KEGG_PATHWAYS_IN_CANCER
- KEGG_CHEMOKINE_SIGNALING_PATHWAY
- KEGG_JAK_STAT_SIGNALING_PATHWAY
- KEGG_CYTOKINE_CYTOKINE_RECEPTOR_INTERACTION
- KEGG_TOLL_LIKE_RECEPTOR_SIGNALING_PATHWAY
- KEGG_FC_GAMMA_R_MEDIATED_PHAGOCYTOSIS
- KEGG_TGF_BETA_SIGNALING_PATHWAY
- KEGG_COLORECTAL_CANCER
- KEGG_CELL_ADHESION_MOLECULES_CAMS
- KEGG_MAPK_SIGNALING_PATHWAY

```{r dotplot pws crcpmpvhc pf macrophages col_pws, fig.width = 6.25, fig.height = 5}
pwoi_macrophages <- c("KEGG_OXIDATIVE_PHOSPHORYLATION",
                      "KEGG_T_CELL_RECEPTOR_SIGNALING_PATHWAY",
                      "KEGG_ANTIGEN_PROCESSING_AND_PRESENTATION",
                      "KEGG_PATHWAYS_IN_CANCER",
                      "KEGG_CHEMOKINE_SIGNALING_PATHWAY",
                      "KEGG_JAK_STAT_SIGNALING_PATHWAY",
                      "KEGG_CYTOKINE_CYTOKINE_RECEPTOR_INTERACTION",
                      "KEGG_TOLL_LIKE_RECEPTOR_SIGNALING_PATHWAY",
                      "KEGG_FC_GAMMA_R_MEDIATED_PHAGOCYTOSIS",
                      "KEGG_TGF_BETA_SIGNALING_PATHWAY",
                      "KEGG_COLORECTAL_CANCER",
                      "KEGG_CELL_ADHESION_MOLECULES_CAMS",
                      "KEGG_MAPK_SIGNALING_PATHWAY")

dotplot_pws_crcpmpvhc_pf_macrophages_colpwoi_ggplotobj <- crcpmpvhc_pf_macrophages_l4_fgsea_sig_df %>%
  dplyr::filter(pathway %in% pwoi_macrophages) %>%
  dplyr::mutate(direction = ifelse(NES<0, "HC", "CRC+"),
                significance = ifelse(pval<0.05, "Significant", "NS"),
                pathway = factor(pathway, levels = unique(pathway)),
                macrophages = factor(macrophages, levels = celltype_order_pf_macrophages_l4$celltype)) %>%
  ggplot(aes(x = macrophages, y = pathway)) +
  geom_point(aes(size = -log10(pval), col = direction, alpha = significance)) +
  theme_bw() +
  scale_color_manual(values = group_colors) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "right",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

pdf(file.path(figS4Dir, "dotplot_pws_crcpmpvhc_pf_macrophages_colpwoi.pdf"), width = 6.25, height = 5)
print(dotplot_pws_crcpmpvhc_pf_macrophages_colpwoi_ggplotobj)
dev.off()
```

Here we extract the GSEA statistics from the analyses performed in a previous script. 

```{r data pws crcpmpvhc pf macrophages}
crcpmpvhc_pf_macrophages_l4_fgsea_sig_df %>%
  dplyr::filter(pathway %in% pwoi_macrophages) %>%
  readr::write_csv(file.path(figS4Dir, "pws_crcpmpvhc_pf_macrophages_pwoi.csv"))
```
