---
title: 'Supplementary Figure 5: Characterization of macrophages in peritoneal fluid of peritoneal metastasized gastric cancer'
author: "Andrew Y.F. Li Yim"
date: '2024-07-26'
output: html_document
---

```{r libraries, include=FALSE}
library(Seurat)
library(ggplot2)
library(dplyr)
library(ggrastr)
library(ggrepel)
library(ggpubr)
```

If the snakemake pipeline was run as-is, the following files will be generated in the folder from which the pipeline was run. If necessary, adjust paths accordingly.

## Setup

### Paths

```{r paths}
base_path <- "/mnt/smb/ltytgat/ayliyim/MI-group/ayliyim/projects/TIPRJ0000015_MAPS/maps_crc"

# Subsets
seurat_gc_pf_rds <- file.path(base_path, "resources/gastric_cancer/maps_gc_pf_reannotated_stable.RDS")
seurat_gc_pf_myeloid_rds <- file.path(base_path, "resources/gastric_cancer/maps_gc_pf_reannotated_stable_mnp.RDS")
seurat_gcpmp_pf_macrophages_rds <- file.path(base_path, "resources/gastric_cancer/maps_gc_pf_macrophages_txpts.Rds")
seurat_gcpmp_tx_macrophages_rds <- file.path(base_path, "resources/gastric_cancer/maps_gc_tx_macrophages.Rds")
seurat_hc_crcpmp_gcpmp_pf_macrophages_rds <- file.path(base_path, "resources/gastric_cancer/hc_crcpmp_gcpmp_pf_macrophages_SeuratObject.Rds")

# Plotting parameters
celltype_markers_xlsx <- file.path(base_path, "config/order/celltype_markers.xlsx")
pf_heatmap_order_xlsx <- file.path(base_path, "config/order/q1_pf_heatmap_order.xlsx")
```

### Import data

```{r load subsets}
seurat_gc_pf <- readRDS(seurat_gc_pf_rds)
seurat_gc_pf_myeloid <- readRDS(seurat_gc_pf_myeloid_rds)
seurat_gcpmp_pf_macrophages <- readRDS(seurat_gcpmp_pf_macrophages_rds)
seurat_gcpmp_tx_macrophages <- readRDS(seurat_gcpmp_tx_macrophages_rds)
seurat_hc_crcpmp_gcpmp_pf_macrophages <- readRDS(seurat_hc_crcpmp_gcpmp_pf_macrophages_rds)
```

```{r load marker genes}
marker_genes <- readxl::read_excel(pf_heatmap_order_xlsx) %>%
  dplyr::filter(Assay == "RNA")
```

### Output

```{r figures}
figS6Dir <- file.path(base_path, "docs/manuscript/figures/figS6")
dir.create(figS6Dir, recursive = T)
```

### Plotting parameters

```{r plotting order}
celltype_order_pf_l2_l3 <- readxl::read_excel(celltype_markers_xlsx, col_names = T) %>%
  dplyr::filter(level == "manual_l3",
                modality == "gene") %>%
  dplyr::select(celltype, color) %>%
  unique() %>%
  dplyr::filter(celltype %in% unique(seurat_gc_pf@meta.data[,"manual_l3"])) %>%
  dplyr::mutate(celltype = ifelse(celltype %in% c("Classical monocytes", "Macrophages VCAN+", "Macrophages C1Q+", "Macrophages VCAN+C1Q+", "Macrophages SPP1+"), "Mono-macs", celltype),
                color = ifelse(celltype %in% c("Mono-macs"), "#E31A1C", color)) %>%
  unique() %>%
  dplyr::mutate(number_subset = paste0(1:nrow(.), ". ", celltype))

celltype_order_pf_macrophages_l4 <- readxl::read_excel(celltype_markers_xlsx, col_names = T) %>%
  dplyr::filter(level == "manual_l4",
                modality == "gene",
                celltype %in% c("Macrophages VCAN+", "Macrophages C1Q+", "Macrophages VCAN+C1Q+", "Macrophages SPP1+", "Macrophages SPP1+")) %>%
  dplyr::select(celltype, color) %>%
  unique() %>%
  dplyr::filter(celltype %in% unique(seurat_hc_crcpmp_gcpmp_pf_macrophages@meta.data[,"manual_l4"])) %>%
  dplyr::mutate(number_subset = paste0(1:nrow(.), ". ", celltype))
```

```{r colors}
manual_l4_colors_pf_macrophages <- celltype_order_pf_macrophages_l4$color
names(manual_l4_colors_pf_macrophages) <- celltype_order_pf_macrophages_l4$celltype

manual_l4_number_colors_pf_macrophages <- celltype_order_pf_macrophages_l4$color
names(manual_l4_number_colors_pf_macrophages) <- celltype_order_pf_macrophages_l4$number_subset
```

## Prepare data

```{r prepare subset}
seurat_gc_pf_l2l3 <- seurat_gc_pf

seurat_gc_pf_l2l3@meta.data <- seurat_gc_pf_l2l3@meta.data %>%
  dplyr::mutate(celltype = manual_l3,
                celltype = ifelse(celltype %in% c("Classical monocytes", "Macrophages C1Q+", "Macrophages SPP1+", "Macrophages VCAN+", "Monocytes", "Macrophages VCAN+C1Q+"), "Mono-macs", as.character(celltype)))
```

## Figures

### UMAP PM-GC PF colored by lineage genes

Here we characterize the GC+ PF-derived immune cells by superimposing the gene expression of the following genes on the respective UMAP configuration:
- T (CD3D)
- NK/ILC (NCAM1)
- B (MS4A1)
- Myeloid (ITGAM)

```{r umap gc pf col_lineage_markers, fig.width=10, fig.height=3}
umap_gc_pf_collineagegenemarkers_ggplotobj_list <- lapply(
  data.frame(CB = colnames(seurat_gc_pf),
             Embeddings(seurat_gc_pf[["umap"]]),
             seurat_gc_pf@meta.data,
             expr = GetAssayData(seurat_gc_pf, layers = "data")["CD3D",],
             Feature = "CD3/CD3D",
             Modality = "RNA") %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_gc_pf),
                                  Embeddings(seurat_gc_pf[["umap"]]),
                                  seurat_gc_pf@meta.data,
                                  expr = GetAssayData(seurat_gc_pf, layers = "data")["NCAM1",],
                                  Feature = "CD56/NCAM1",
                                  Modality = "RNA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_gc_pf),
                                  Embeddings(seurat_gc_pf[["umap"]]),
                                  seurat_gc_pf@meta.data,
                                  expr = GetAssayData(seurat_gc_pf, layers = "data")["MS4A1",],
                                  Feature = "CD20/MS4A1",
                                  Modality = "RNA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_gc_pf),
                                  Embeddings(seurat_gc_pf[["umap"]]),
                                  seurat_gc_pf@meta.data,
                                  expr = GetAssayData(seurat_gc_pf, layers = "data")["ITGAM",],
                                  Feature = "CD11b/ITGAM",
                                  Modality = "RNA")) %>%
    dplyr::mutate(expr_rank = rank(expr, ties.method="first"),
                  Feature = factor(Feature, levels = c("CD3/CD3D", "CD56/NCAM1", "CD20/MS4A1", "CD11b/ITGAM"))) %>%
    split(., .$Feature), FUN = function(plotdf){
      plotdf %>% dplyr::arrange(expr) %>%
        ggplot(aes(x = umap_1, y = umap_2)) +
        geom_point_rast(show.legend = F, size = 0.25, col = "#d3d3d3") +
        geom_point_rast(data = . %>%
                          dplyr::filter(expr>0), aes(col = expr), show.legend = T) +
        guides(colour = guide_legend(override.aes = list(size = 3))) +
        facet_wrap(~Feature) +
        guides(colour = guide_legend(override.aes = list(size = 3),
                                     title = "Expression")) +
        scale_color_viridis() +
        theme_bw() +
        theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              legend.position = "none",
              axis.text.x = element_blank(),
              axis.text.y = element_blank(),
              axis.ticks = element_blank(),
              axis.title.x = element_blank(),
              axis.title.y = element_blank(),
              strip.text.x = element_text(face = "bold"))
    }
)

pdf(file.path(figS6Dir, "umap_gc_pf_collineagegenemarkers.pdf"), width = 10, height = 3)
print(ggarrange(plotlist = umap_gc_pf_collineagegenemarkers_ggplotobj_list, nrow = 1, ncol = 4))
dev.off()
```

### UMAP PM-GC PF mononuclear phagocytes colored by mononuclear phagocytes marker genes

Here we characterize the PM-GC PF-derived mononuclear phagocytes by superimposing the gene expression of the following genes on the respective UMAP configuration:
- Plasmacytoid dendritic cells (CLEC4C)
- Conventional dendritic cells type 1 (XCR1)
- Conventional dendritic cells type 2 (CD1C)
- Myeloid (ITGAM)
- Macrophages (MARCO)
- Macrophages (CD163)
- Monocyte-macrophages (CD14)
- Monocytes (CCR2)

```{r umap gc pf col_myeloidgenemarkers, fig.width=6, fig.height=12}
umap_gc_pf_mnp_colmnpgenemarkers_ggplotobj_list <- lapply(
  data.frame(CB = colnames(seurat_gc_pf_myeloid),
             Embeddings(seurat_gc_pf_myeloid[["umap"]]),
             seurat_gc_pf_myeloid@meta.data,
             expr = GetAssayData(seurat_gc_pf_myeloid, layers = "data")["CLEC4C",],
             Feature = "CLEC4C") %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_gc_pf_myeloid ),
                                  Embeddings(seurat_gc_pf_myeloid[["umap"]]),
                                  seurat_gc_pf_myeloid@meta.data,
                                  expr = GetAssayData(seurat_gc_pf_myeloid, layers = "data")["XCR1",],
                                  Feature = "XCR1")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_gc_pf_myeloid ),
                                  Embeddings(seurat_gc_pf_myeloid[["umap"]]),
                                  seurat_gc_pf_myeloid@meta.data,
                                  expr = GetAssayData(seurat_gc_pf_myeloid, layers = "data")["CD1C",],
                                  Feature = "CD1C")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_gc_pf_myeloid ),
                                  Embeddings(seurat_gc_pf_myeloid[["umap"]]),
                                  seurat_gc_pf_myeloid@meta.data,
                                  expr = GetAssayData(seurat_gc_pf_myeloid, layers = "data")["ITGAM",],
                                  Feature = "ITGAM")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_gc_pf_myeloid ),
                                  Embeddings(seurat_gc_pf_myeloid[["umap"]]),
                                  seurat_gc_pf_myeloid@meta.data,
                                  expr = GetAssayData(seurat_gc_pf_myeloid, layers = "data")["MARCO",],
                                  Feature = "MARCO")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_gc_pf_myeloid ),
                                  Embeddings(seurat_gc_pf_myeloid[["umap"]]),
                                  seurat_gc_pf_myeloid@meta.data,
                                  expr = GetAssayData(seurat_gc_pf_myeloid, layers = "data")["CD163",],
                                  Feature = "CD163")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_gc_pf_myeloid ),
                                  Embeddings(seurat_gc_pf_myeloid[["umap"]]),
                                  seurat_gc_pf_myeloid@meta.data,
                                  expr = GetAssayData(seurat_gc_pf_myeloid, layers = "data")["CD14",],
                                  Feature = "CD14")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_gc_pf_myeloid ),
                                  Embeddings(seurat_gc_pf_myeloid[["umap"]]),
                                  seurat_gc_pf_myeloid@meta.data,
                                  expr = GetAssayData(seurat_gc_pf_myeloid, layers = "data")["CCR2",],
                                  Feature = "CCR2")) %>%
    dplyr::mutate(expr_rank = rank(expr, ties.method="first"),
                  Feature = factor(Feature, levels = c("CLEC4C", "XCR1", "CD1C", "ITGAM", "MARCO", "CD163", "CD14", "CCR2"))) %>%
    dplyr::filter(manual_l2 %in% c("Macrophages", "CDCs", "PDCs")) %>%
    split(., .$Feature), FUN = function(plotdf){
      plotdf %>% dplyr::arrange(expr) %>%
        ggplot(aes(x = umap_1, y = umap_2)) +
        geom_point_rast(show.legend = F, size = 0.25, col = "#d3d3d3") +
        geom_point_rast(data = . %>%
                          dplyr::filter(expr>0), aes(col = expr), show.legend = T) +
        guides(colour = guide_legend(override.aes = list(size = 3))) +
        facet_wrap(~Feature) +
        guides(colour = guide_legend(override.aes = list(size = 3),
                                     title = "Expression")) +
        scale_color_viridis() +
        theme_bw() +
        theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              legend.position = "none",
              axis.title.x = element_blank(),
              axis.title.y = element_blank(),
              strip.text.x = element_text(face = "bold"))
    }
)

pdf(file.path(figS6Dir, "umap_gc_pf_mnp_colmnpgenemarkers.pdf"), width = 6, height = 12)
print(ggarrange(plotlist = umap_gc_pf_mnp_colmnpgenemarkers_ggplotobj_list, nrow = 4, ncol = 2))
dev.off()
```

### UMAP PM-GC PF macrophages colored by macrophage marker genes

Here we characterize the PM-GC PF-derived macrophages by superimposing the gene expression of the following genes on the respective UMAP configuration:
- Macrophages (CD163)
- Macrophages C1Q+ [aka "CMs C1Q+"] (C1QA)
- Macrophages VCAN+ [aka "mono-CMs"] (VCAN)
- Monocytes (CCR2)
- Macrophages SPP1+ [aka "CMs SPP1+"] (SPP1)

```{r umap gc pm macrophages col_macrophagegenemarkers, fig.width=6, fig.height=9}
umap_gc_pf_macrophages_colmacrophagegenemarkers_ggplotobj_list <- lapply(
  data.frame(CB = colnames(seurat_gcpmp_pf_macrophages),
             Embeddings(seurat_gcpmp_pf_macrophages[["umap"]]),
             seurat_gcpmp_pf_macrophages@meta.data,
             expr = GetAssayData(seurat_gcpmp_pf_macrophages, layers = "data")["CD163",],
             Feature = "CD163") %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_gcpmp_pf_macrophages),
                                  Embeddings(seurat_gcpmp_pf_macrophages[["umap"]]),
                                  seurat_gcpmp_pf_macrophages@meta.data,
                                  expr = GetAssayData(seurat_gcpmp_pf_macrophages, layers = "data")["C1QA",],
                                  Feature = "C1QA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_gcpmp_pf_macrophages),
                                  Embeddings(seurat_gcpmp_pf_macrophages[["umap"]]),
                                  seurat_gcpmp_pf_macrophages@meta.data,
                                  expr = GetAssayData(seurat_gcpmp_pf_macrophages, layers = "data")["VCAN",],
                                  Feature = "VCAN")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_gcpmp_pf_macrophages),
                                  Embeddings(seurat_gcpmp_pf_macrophages[["umap"]]),
                                  seurat_gcpmp_pf_macrophages@meta.data,
                                  expr = GetAssayData(seurat_gcpmp_pf_macrophages, layers = "data")["CCR2",],
                                  Feature = "CCR2")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_gcpmp_pf_macrophages),
                                  Embeddings(seurat_gcpmp_pf_macrophages[["umap"]]),
                                  seurat_gcpmp_pf_macrophages@meta.data,
                                  expr = GetAssayData(seurat_gcpmp_pf_macrophages, layers = "data")["SPP1",],
                                  Feature = "SPP1")) %>%
    dplyr::mutate(expr_rank = rank(expr, ties.method="first"),
                  Feature = factor(Feature, levels = c("CD163", "C1QA", "VCAN", "CCR2", "SPP1"))) %>%
    split(., .$Feature), FUN = function(plotdf){
      plotdf %>% dplyr::arrange(expr) %>%
        ggplot(aes(x = umap_1, y = umap_2)) +
        geom_point_rast(show.legend = F, size = 0.25, col = "#d3d3d3") +
        geom_point_rast(data = . %>%
                          dplyr::filter(expr>0), aes(col = expr), show.legend = T) +
        guides(colour = guide_legend(override.aes = list(size = 3))) +
        facet_wrap(~Feature) +
        guides(colour = guide_legend(override.aes = list(size = 3),
                                     title = "Expression")) +
        scale_color_viridis() +
        theme_bw() +
        theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              legend.position = "none",
              axis.title.x = element_blank(),
              axis.title.y = element_blank(),
              strip.text.x = element_text(face = "bold"))
    }
)

pdf(file.path(figS6Dir, "umap_gc_pf_macrophages_colmacrophagegenemarkers.pdf"), width = 6, height = 9)
print(ggarrange(plotlist = umap_gc_pf_macrophages_colmacrophagegenemarkers_ggplotobj_list, nrow = 3, ncol = 2))
dev.off()
```

### Dotplot GEX HC PF colored by l2/l3

```{r dotplot gex hc pf col_l2l3, fig.width = 7.25, fig.height=15}
marker_gc_pf_gex <- GetAssayData(seurat_gc_pf_l2l3, assay = "RNA")[which(rownames(GetAssayData(seurat_gc_pf_l2l3, assay = "RNA")) %in% unique(marker_genes$FeatureID)), ]

dotplot_gc_pf_l2l3_cellgexmarkers_ggplotobj <- data.frame(FeatureID = rownames(marker_gc_pf_gex), marker_gc_pf_gex) %>%
  tidyr::pivot_longer(-c(FeatureID), names_to = "CB", values_to = "nUMIs") %>%
  dplyr::mutate(CB = stringr::str_replace(CB, "\\.", "-")) %>%
  dplyr::inner_join(data.frame(CB = rownames(seurat_gc_pf_l2l3@meta.data), 
                               Celltype = seurat_gc_pf_l2l3@meta.data[,"celltype"]), 
                    by = "CB") %>%
  dplyr::filter(Celltype %in% celltype_order_pf_l2_l3$celltype) %>%
  dplyr::group_by(Celltype, FeatureID, .drop = F) %>%
  dplyr::summarize(Median = median(log1p(nUMIs)),
                   Percentage = mean(nUMIs>0)*100) %>%
  dplyr::mutate(Median = ifelse(is.na(Median), 0, Median),
                Percentage = ifelse(is.na(Percentage), 0, Percentage),
                Celltype = factor(Celltype, celltype_order_pf_l2_l3$celltype),
                FeatureID = factor(FeatureID, rev(marker_genes$FeatureID))) %>% 
  ggplot(aes(x = Celltype, y = FeatureID, col = Median)) +
  geom_tile(alpha = 0, col = "grey") +
  geom_point(aes(size = Percentage)) +
  theme_bw() +
  theme(legend.position = "bottom", 
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.spacing.x=unit(0, "lines"))

pdf(file.path(figS6Dir, "dotplot_gc_pf_l2l3_cellgexmarkers_ggplotobj.pdf"), width = 7.5, height=15)
print(dotplot_gc_pf_l2l3_cellgexmarkers_ggplotobj)
dev.off()
```

### UMAP HC PM-GC PF colored by l4 

```{r umap hc crcpmp gcpmp pf macrophages col_l4, fig.width=10.25, fig.height=3.75}
umap_hc_crcpmp_gcpmp_pf_macrophages_l4_ggplotobj <- data.frame(CB = colnames(seurat_hc_crcpmp_gcpmp_pf_macrophages),
                                                               Embeddings(seurat_hc_crcpmp_gcpmp_pf_macrophages[["umap"]]),
                                                               seurat_hc_crcpmp_gcpmp_pf_macrophages@meta.data) %>%
  dplyr::mutate(Group = factor(Group, levels = c("HC", "CRC+", "GC+")),
                celltype = factor(manual_l4, levels = celltype_order_pf_macrophages_l4$celltype),
                celltype_number = as.numeric(celltype),
                celltype_w_number = paste0(as.numeric(celltype), ". ", celltype),
                celltype_w_number = factor(celltype_w_number)) %>% 
  ggplot(aes(x = umap_1, y = umap_2)) +
  geom_point_rast(show.legend = T, size = 2, aes(col = celltype_w_number)) +
  labs(y = "",
       x = "") +
  geom_label_repel(data = . %>%
                     dplyr::group_by(celltype, celltype_number, celltype_w_number) %>%
                     summarize(x = median(x = umap_1),
                               y = median(x = umap_2)),
                   mapping = aes(label = celltype_number, x = x, y = y),
                   alpha = 1, 
                   size = 8,
                   show.legend = F,
                   label.size = 0.25,
  ) +
  guides(colour = guide_legend(override.aes = list(size = 3))) +
  scale_color_manual(values = manual_l4_number_colors_pf_macrophages) +
  theme_minimal() +
  facet_wrap(~Group) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.x = element_text(face = "bold"))

pdf(file.path(figS6ir, "umap_hc_crcpmp_gcpmp_pf_macrophages_l4.pdf"), width = 10.25, height=3.75)
print(umap_hc_crcpmp_gcpmp_pf_macrophages_l4_ggplotobj)
dev.off()
```

### Stackedbarplot PM-CRC PM-GC PF macrophages colored by l4

```{r stackedbarplot crcpmp gcpmp pf macrophages col_l4, fig.width=3, fig.height=5}
stackedbarplot_hc_crcpmp_gcpmp_pf_macrophages_coll4_ggplotobj <- seurat_hc_crcpmp_gcpmp_pf_macrophages@meta.data %>%
  dplyr::group_by(manual_l4, Group, .drop = T) %>% 
  dplyr::summarise(Ncells = n()) %>%
  dplyr::ungroup() %>%
  tidyr::complete(tidyr::nesting(Group), tidyr::nesting(manual_l4), fill = list(Ncells = 0)) %>%
  dplyr::group_by(Group) %>%
  dplyr::mutate(manual_l4 = factor(manual_l4, levels = celltype_order_pf_macrophages_l4$celltype),
                Nsample = sum(Ncells),
                Ncellprop = Ncells/Nsample,
                Ncellprop = ifelse(is.na(Ncellprop), 0, Ncellprop),
                Ncellperc = Ncellprop*100,
                Group = factor(Group, levels = c("HC", "CRC+", "GC+"))) %>%
  dplyr::rename(Celltype = manual_l4) %>%
  ggplot(aes(x = Group, y = Ncellperc)) +
  geom_bar(position="stack", stat="identity", aes(fill = Celltype)) +
  labs(y = "%Macrophages") +
  scale_fill_manual(values = manual_l4_colors_pf_macrophages) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "bottom",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

pdf(file.path(figS6Dir, "stackedbarplot_hc_crcpmp_gcpmp_pf_macrophages_coll4.pdf"), width = 3, height = 5)
print(stackedbarplot_hc_crcpmp_gcpmp_pf_macrophages_coll4_ggplotobj)
dev.off()
```

### UMAP HC PM-GC PF macrophages colored by l4

```{r umap hc crcpmp gcpmp pf macrophages col_l4, fig.width=10.25, fig.height=3.75}
umap_gcpmp_pf_macrophages_l4_ggplotobj <- data.frame(CB = colnames(seurat_hc_crcpmp_gcpmp_pf_macrophages),
                                                     Embeddings(seurat_hc_crcpmp_gcpmp_pf_macrophages[["umap"]]),
                                                     seurat_hc_crcpmp_gcpmp_pf_macrophages@meta.data) %>%
  dplyr::mutate(Group = factor(Group, levels = c("HC", "CRC+", "GC+")),
                celltype = factor(manual_l4, levels = celltype_order_pf_macrophages_l4$celltype),
                celltype_number = as.numeric(celltype),
                celltype_w_number = paste0(as.numeric(celltype), ". ", celltype),
                celltype_w_number = factor(celltype_w_number)) %>% 
  ggplot(aes(x = umap_1, y = umap_2)) +
  geom_point_rast(show.legend = T, size = 2, aes(col = celltype_w_number)) +
  labs(y = "",
       x = "") +
  geom_label_repel(data = . %>%
                     dplyr::group_by(celltype, celltype_number, celltype_w_number) %>%
                     summarize(x = median(x = umap_1),
                               y = median(x = umap_2)),
                   mapping = aes(label = celltype_number, x = x, y = y),
                   alpha = 1, 
                   size = 8,
                   show.legend = F,
                   label.size = 0.25,
  ) +
  guides(colour = guide_legend(override.aes = list(size = 3))) +
  scale_color_manual(values = manual_l4_number_colors_pf_macrophages) +
  theme_minimal() +
  facet_wrap(~Group) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.x = element_text(face = "bold"))

pdf(file.path(figS6Dir, "umap_hc_crcpmp_gcpmp_pf_macrophages_l4.pdf"), width = 10.25, height=3.75)
print(umap_gcpmp_pf_macrophages_l4_ggplotobj)
dev.off()
```

### Boxplot PM-GC TX lineage relative to all immune cells

Here we prepare the data for scattered boxplots depicting the PM-GC-derived TX macrophage subsets relative to the total macrophage population. Actual data was plotted in Graphpad Prism.

```{r crcpmp tx macrophages l4rl2 data}
seurat_gcpmp_tx_macrophages@meta.data %>%
  dplyr::group_by(manual_l4, SampleID, .drop = T) %>% 
  dplyr::summarise(Ncells = n()) %>%
  dplyr::ungroup() %>%
  tidyr::complete(tidyr::nesting(SampleID), tidyr::nesting(manual_l4), fill = list(Ncells = 0)) %>%
  dplyr::group_by(SampleID) %>%
  dplyr::mutate(Nlrefsample = sum(Ncells),
                Ncellprop = Ncells/Nlrefsample,
                Ncellprop = ifelse(is.na(Ncellprop), 0, Ncellprop),
                Ncellperc = Ncellprop*100,
                Donor = gsub("^(pt[0-9]{2})_.+$", "\\1", SampleID)) %>%
  readr::write_csv(file.path(figS6Dir, "crcpmp_tx_macrophages_l4rl2.csv"))
```
