---
title: 'Supplementary Figure 6: Characterized peritoneal metastatic tumor immune infiltrate'
author: "Andrew Y.F. Li Yim"
date: '2023-07-31'
output: html_document
---

```{r libraries, include=FALSE}
library(Seurat)
library(ggplot2)
library(dplyr)
library(ggrastr)
library(ggrepel)
library(ggpubr)
library(viridis)
library(SummarizedExperiment)
library(slingshot)
library(SingleCellExperiment)
library(DESeq2)
library(ggdendro)
library(ggalt)
library(lme4)
library(nlme)
library(emmeans)
```

If the snakemake pipeline was run as-is, the following files will be generated in the folder from which the pipeline was run. If not, then adjust paths accordingly.    

## Setup

### Paths

```{r paths}
base_path <- "/mnt/smb/ltytgat/ayliyim/MI-group/ayliyim/projects/TIPRJ0000015_MAPS/maps_crc"

# Subsets
seurat_crcpmp_tx_immune_rds <- file.path(base_path, "output/q3_pm_tx_characterization/subsets/crcpmp_tx_immune_SeuratObject.Rds")
seurat_crcpmp_tx_monocytes_macrophages_rds <- file.path(base_path, "output/q3_pm_tx_characterization/subsets/crcpmp_tx_monocytes_macrophages_SeuratObject.Rds")
seurat_hc_pf_myeloid_rds <- file.path(base_path, "output/q1_pf_characterization/subsets/hc_pf_Myeloid_SeuratObject.Rds")

# Plotting parameters
celltype_markers_xlsx <- file.path(base_path, "config/order/celltype_markers.xlsx")
pf_heatmap_order_xlsx <- file.path(base_path, "config/order/q1_pf_heatmap_order.xlsx")
```

### Import data

```{r load subsets}
seurat_crcpmp_tx_immune <- readRDS(seurat_crcpmp_tx_immune_rds)
seurat_crcpmp_tx_monocytes_macrophages <- readRDS(seurat_crcpmp_tx_monocytes_macrophages_rds)
seurat_hc_pf_myeloid <- readRDS(seurat_hc_pf_myeloid_rds)
```

```{r load marker genes}
marker_genes <- readxl::read_excel(pf_heatmap_order_xlsx) %>%
  dplyr::filter(Assay == "RNA")
```

### Output

```{r figures}
figS5Dir <- file.path(base_path, "docs/manuscript/figures/figS5")
dir.create(figS5Dir, recursive = T)
```

### Plotting parameters

```{r plotting order}
celltype_order_l3_tx_immune <- readxl::read_excel(celltype_markers_xlsx, col_names = T) %>%
  dplyr::filter(level == "manual_l3",
                modality == "gene") %>%
  dplyr::select(celltype, color) %>%
  unique() %>%
  dplyr::filter(!is.na(color),
                celltype %in% seurat_crcpmp_tx_immune$manual_l3) %>%
  dplyr::mutate(number_subset = paste0(1:nrow(.), ". ", celltype))

celltype_order_pf_macrophages_l4 <- readxl::read_excel(celltype_markers_xlsx, col_names = T) %>%
  dplyr::filter(level == "manual_l4",
                modality == "gene",
                celltype %in% c("Macrophages VCAN+", "Macrophages C1Q+", "Macrophages VCAN+C1Q+", "Macrophages SPP1+", "Macrophages SPP1+")) %>%
  dplyr::select(celltype, color) %>%
  unique() %>%
  dplyr::filter(celltype %in% unique(seurat_gcpmp_tx_macrophages@meta.data[,"manual_l4"])) %>%
  dplyr::mutate(number_subset = paste0(1:nrow(.), ". ", celltype))
```

```{r colors}
tissue_colors <- c(PBMC = "#F30000", PF = "#224FBD", TX = "#855522")

manual_l3_tx_immune_colors <- celltype_order_l3_tx_immune$color
names(manual_l3_tx_immune_colors) <- celltype_order_l3_tx_immune$number_subset

manual_l4_number_colors_pf_macrophages <- celltype_order_pf_macrophages_l4$color
names(manual_l4_number_colors_pf_macrophages) <- celltype_order_pf_macrophages_l4$number_subset
```

## Figures

### Dotplot GEX CRC TX colored by l2/l3

```{r dotplot crcpmp tx, fig.width = 7.25, fig.height=15}
marker_genes_wvsig4 <- marker_genes[1:which(marker_genes$FeatureID == "CD163")-1,] %>%
  dplyr::rows_append(data.frame(FeatureID = "VSIG4",
                                Assay = "RNA",
                                protein = NA,
                                gene = "VSIG4")) %>%
  dplyr::rows_append(marker_genes[which(marker_genes$FeatureID == "CD163"):nrow(marker_genes),])

marker_crcpmp_tx_gex <- GetAssayData(seurat_crcpmp_tx_immune, assay = "RNA")[which(rownames(GetAssayData(seurat_crcpmp_tx_immune, assay = "RNA")) %in% unique(marker_genes_wvsig4$FeatureID)), ]

dotplot_crcpmp_tx_immune_cellmarkers_ggplotobj <- data.frame(FeatureID = rownames(marker_crcpmp_tx_gex), marker_crcpmp_tx_gex) %>%
  tidyr::pivot_longer(-c(FeatureID), names_to = "CB", values_to = "nUMIs") %>%
  dplyr::mutate(CB = stringr::str_replace(CB, "\\.", "-")) %>%
  dplyr::inner_join(data.frame(CB = rownames(seurat_crcpmp_tx_immune@meta.data), 
                               Celltype = seurat_crcpmp_tx_immune@meta.data[,"manual_l3"]), 
                    by = "CB") %>%
  dplyr::mutate(Celltype = factor(Celltype), 
                FeatureID = factor(FeatureID)) %>%
  dplyr::group_by(Celltype, FeatureID, .drop = F) %>%
  dplyr::summarize(Median = median(log1p(nUMIs)),
                   Percentage = mean(nUMIs>0)*100) %>%
  dplyr::mutate(Median = ifelse(is.na(Median), 0, Median),
                Percentage = ifelse(is.na(Percentage), 0, Percentage),
                Celltype = factor(Celltype, celltype_order_l3_tx_immune$celltype),
                FeatureID = factor(FeatureID, rev(marker_genes_wvsig4$FeatureID))) %>% 
  ggplot(aes(x = Celltype, y = FeatureID, col = Median)) +
  geom_tile(alpha = 0, col = "grey") +
  geom_point(aes(size = Percentage)) +
  theme_bw() +
  theme(legend.position = "bottom", 
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.spacing.x=unit(0, "lines"))

pdf(file.path(figS5Dir, "dotplot_crcpmp_tx_immune_cellmarkers.pdf"), width = 7.5, height = 15)
print(dotplot_crcpmp_tx_immune_cellmarkers_ggplotobj)
dev.off()
```

### UMAP PM-CRC TX colored by l3 labeled by l3

```{r umap crcpmp tx col_l3 lab_l3, fig.width=8, fig.height=9}
umap_crcpmp_tx_coll3_labl3_ggplotobj <- data.frame(CB = colnames(seurat_crcpmp_tx_immune),
                                                   Embeddings(seurat_crcpmp_tx_immune[["umap"]]),
                                                   seurat_crcpmp_tx_immune@meta.data) %>%
  dplyr::mutate(celltype = factor(manual_l3, levels = celltype_order_l3_tx_immune$celltype),
                celltype_number = as.numeric(celltype),
                celltype_w_number = paste0(as.numeric(celltype), ". ", celltype),
                celltype_w_number = factor(celltype_w_number, levels = celltype_order_l3_tx_immune$number_subset)) %>% 
  ggplot(aes(x = UMAP_1, y = UMAP_2)) +
  geom_point_rast(show.legend = T, size = 0.25, aes(col = celltype_w_number)) +
  labs(y = "",
       x = "") +
  geom_label_repel(data = . %>%
                     dplyr::group_by(celltype, celltype_number, celltype_w_number) %>%
                     summarize(x = median(x = UMAP_1),
                               y = median(x = UMAP_2)),
                   mapping = aes(label = celltype_number, x = x, y = y),
                   alpha = 1, 
                   size = 8,
                   label.size=0.25,
                   show.legend = F) +
  guides(colour = guide_legend(override.aes = list(size = 3))) +
  scale_color_manual(values = manual_l3_tx_immune_colors) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.x = element_text(face = "bold"))

print(umap_crcpmp_tx_coll3_labl3_ggplotobj)

pdf(file.path(figS5Dir, "umap_crcpmp_tx_coll3_labl3.pdf"), width = 7.5, height=7.5)
print(umap_crcpmp_tx_coll3_labl3_ggplotobj)
dev.off()
```

### UMAP PM-CRC TX monocytes and macrophages colored by myeloid markers

```{r umap crcpmp tx monocytes macrophages col_myeloidmarkers, fig.width=9, fig.height=9}
umap_crcpmp_tx_monocytes_macrophages_colmyeloidgenemarkers_ggplotobj_list <- lapply(
  data.frame(CB = colnames(seurat_crcpmp_tx_monocytes_macrophages),
             Embeddings(seurat_crcpmp_tx_monocytes_macrophages[["umap"]]),
             seurat_crcpmp_tx_monocytes_macrophages@meta.data,
             expr = GetAssayData(seurat_crcpmp_tx_monocytes_macrophages, assay = "RNA")["C1QA",],
             Feature = "C1QA") %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_crcpmp_tx_monocytes_macrophages),
                                  Embeddings(seurat_crcpmp_tx_monocytes_macrophages[["umap"]]),
                                  seurat_crcpmp_tx_monocytes_macrophages@meta.data,
                                  expr = GetAssayData(seurat_crcpmp_tx_monocytes_macrophages, assay = "RNA")["CCR2",],
                                  Feature = "CCR2")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_crcpmp_tx_monocytes_macrophages),
                                  Embeddings(seurat_crcpmp_tx_monocytes_macrophages[["umap"]]),
                                  seurat_crcpmp_tx_monocytes_macrophages@meta.data,
                                  expr = GetAssayData(seurat_crcpmp_tx_monocytes_macrophages, assay = "RNA")["CD14",],
                                  Feature = "CD14")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_crcpmp_tx_monocytes_macrophages),
                                  Embeddings(seurat_crcpmp_tx_monocytes_macrophages[["umap"]]),
                                  seurat_crcpmp_tx_monocytes_macrophages@meta.data,
                                  expr = GetAssayData(seurat_crcpmp_tx_monocytes_macrophages, assay = "RNA")["CD163",],
                                  Feature = "CD163")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_crcpmp_tx_monocytes_macrophages),
                                  Embeddings(seurat_crcpmp_tx_monocytes_macrophages[["umap"]]),
                                  seurat_crcpmp_tx_monocytes_macrophages@meta.data,
                                  expr = GetAssayData(seurat_crcpmp_tx_monocytes_macrophages, assay = "RNA")["FCGR3A",],
                                  Feature = "FCGR3A")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_crcpmp_tx_monocytes_macrophages),
                                  Embeddings(seurat_crcpmp_tx_monocytes_macrophages[["umap"]]),
                                  seurat_crcpmp_tx_monocytes_macrophages@meta.data,
                                  expr = GetAssayData(seurat_crcpmp_tx_monocytes_macrophages, assay = "RNA")["MARCO",],
                                  Feature = "MARCO")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_crcpmp_tx_monocytes_macrophages),
                                  Embeddings(seurat_crcpmp_tx_monocytes_macrophages[["umap"]]),
                                  seurat_crcpmp_tx_monocytes_macrophages@meta.data,
                                  expr = GetAssayData(seurat_crcpmp_tx_monocytes_macrophages, assay = "RNA")["SPP1",],
                                  Feature = "SPP1")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_crcpmp_tx_monocytes_macrophages),
                                  Embeddings(seurat_crcpmp_tx_monocytes_macrophages[["umap"]]),
                                  seurat_crcpmp_tx_monocytes_macrophages@meta.data,
                                  expr = GetAssayData(seurat_crcpmp_tx_monocytes_macrophages, assay = "RNA")["VCAN",],
                                  Feature = "VCAN")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_crcpmp_tx_monocytes_macrophages),
                                  Embeddings(seurat_crcpmp_tx_monocytes_macrophages[["umap"]]),
                                  seurat_crcpmp_tx_monocytes_macrophages@meta.data,
                                  expr = GetAssayData(seurat_crcpmp_tx_monocytes_macrophages, assay = "RNA")["VSIG4",],
                                  Feature = "VSIG4")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_crcpmp_tx_monocytes_macrophages),
                                  Embeddings(seurat_crcpmp_tx_monocytes_macrophages[["umap"]]),
                                  seurat_crcpmp_tx_monocytes_macrophages@meta.data,
                                  expr = GetAssayData(seurat_crcpmp_tx_monocytes_macrophages, assay = "RNA")["MRC1",],
                                  Feature = "MRC1")) %>%
    dplyr::mutate(expr_rank = rank(expr, ties.method="first"),
                  Feature = factor(Feature, levels = c("C1QA", "CCR2", "CD14", "CD163", "FCGR3A", "MARCO", "SPP1", "VCAN", "VSIG4", "MRC1"))) %>%
    split(., .$Feature), FUN = function(plotdf){
      plotdf %>% dplyr::arrange(expr) %>%
        ggplot(aes(x = UMAP_1, y = UMAP_2)) +
        geom_point_rast(show.legend = F, size = 0.25, col = "#d3d3d3") +
        geom_point_rast(data = . %>%
                          dplyr::filter(expr>0), aes(col = expr), show.legend = T) +
        guides(colour = guide_legend(override.aes = list(size = 3))) +
        facet_wrap(~Feature) +
        guides(colour = guide_legend(override.aes = list(size = 3),
                                     title = "Expression")) +
        scale_color_viridis() +
        theme_bw() +
        theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              legend.position = "none",
              axis.title.x = element_blank(),
              axis.title.y = element_blank(),
              strip.text.x = element_text(face = "bold"))
    }
)

pdf(file.path(figS5Dir, "umap_crcpmp_tx_monocytes_macrophages_colmyeloidgenemarkers_colviridis.pdf"), width = 9, height = 12)
print(ggarrange(plotlist = umap_crcpmp_tx_monocytes_macrophages_colmyeloidgenemarkers_ggplotobj_list, nrow = 4, ncol = 3))
dev.off()
```

### Boxplot PM-CRC TX lineage relative to all immune cells

Here we prepare the data for scattered boxplots depicting the PM-CRC-derived TX immune lineages relative to the total immune population. Actual data was plotted in Graphpad Prism.

```{r crcpmp tx l3rl0 data}
seurat_crcpmp_tx_immune@meta.data %>%
  dplyr::group_by(manual_l3, SampleID, .drop = T) %>% 
  dplyr::summarise(Ncells = n()) %>%
  dplyr::ungroup() %>%
  tidyr::complete(tidyr::nesting(SampleID), tidyr::nesting(manual_l3), fill = list(Ncells = 0)) %>%
  dplyr::group_by(SampleID) %>%
  dplyr::mutate(Nlrefsample = sum(Ncells),
                Ncellprop = Ncells/Nlrefsample,
                Ncellprop = ifelse(is.na(Ncellprop), 0, Ncellprop),
                Ncellperc = Ncellprop*100,
                Donor = gsub("^(pt[0-9]{2})_.+$", "\\1", SampleID)) %>%
  dplyr::left_join(unique(seurat_crcpmp_tx_immune@meta.data[,c("manual_l1", "manual_l3")]), by = "manual_l3") %>%
  readr::write_csv(file.path(figS5Dir, "crcpmp_tx_immune_l3rl0.csv"))
```

### UMAP HC PF macrophages colored by macrophage CITE markers

Here we characterize the PF-derived CD163+ macrophages by superimposing the gene expression of the following genes on the respective UMAP configuration:
- Monocytes/Monocyte-macrophages/Macrophages (CD194)
- Monocytes/Monocyte-macrophages/Macrophages (CD195)
- Monocytes/Monocyte-macrophages/Macrophages (CD196)

```{r umap hc pf myeloid col_cxcrcitemarkers, fig.width=6, fig.height=3}
umap_hc_pf_macrophages_colcxcrcitemarkers_ggplotobj_list <- lapply(
  data.frame(CB = colnames(seurat_hc_pf_myeloid),
             Embeddings(seurat_hc_pf_myeloid[["wnn.umap"]]),
             seurat_hc_pf_myeloid@meta.data,
             expr = GetAssayData(seurat_hc_pf_myeloid, assay = "CITE")["Hu.CD194",],
             Feature = "CD194") %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_hc_pf_myeloid),
                                  Embeddings(seurat_hc_pf_myeloid[["wnn.umap"]]),
                                  seurat_hc_pf_myeloid@meta.data,
                                  expr = GetAssayData(seurat_hc_pf_myeloid, assay = "CITE")["Hu.CD195",],
                                  Feature = "CD195"))  %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_hc_pf_myeloid),
                                  Embeddings(seurat_hc_pf_myeloid[["wnn.umap"]]),
                                  seurat_hc_pf_myeloid@meta.data,
                                  expr = GetAssayData(seurat_hc_pf_myeloid, assay = "CITE")["Hu.CD196",],
                                  Feature = "CD196")) %>%
    dplyr::mutate(expr_rank = rank(expr, ties.method="first"),
                  Feature = factor(Feature, levels = c("CD194", "CD195", "CD196"))) %>%
    split(., .$Feature), FUN = function(plotdf){
      plotdf %>% dplyr::arrange(expr) %>%
        ggplot(aes(x = wnnUMAP_1, y = wnnUMAP_2)) +
        geom_point_rast(show.legend = F, size = 0.25, col = "#d3d3d3") +
        geom_point_rast(data = . %>%
                          dplyr::filter(expr>0), aes(col = expr), show.legend = T) +
        guides(colour = guide_legend(override.aes = list(size = 3))) +
        facet_wrap(~Feature) +
        guides(colour = guide_legend(override.aes = list(size = 3),
                                     title = "Expression")) +
        scale_colour_gradient(low = "#ffffff", high = "#FF0000") +
        xlim(-6,11) +
        ylim(-5,8) +
        theme_bw() +
        theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              legend.position = "none",
              axis.title.x = element_blank(),
              axis.title.y = element_blank(),
              strip.text.x = element_text(face = "bold"))
    }
)

pdf(file.path(figS5Dir, "umap_hc_pf_macrophages_colcxcrcitemarkers.pdf"), width = 6, height = 3)
print(ggarrange(plotlist = umap_hc_pf_macrophages_colcxcrcitemarkers_ggplotobj_list, ncol = 3))
dev.off()
```