---
title: 'Supplementary Figure 2: PerIS HC vs PM-CRC'
author: "Andrew Y.F. Li Yim"
date: '2023-07-31'
output: html_document
---

```{r libraries, include=FALSE}
library(Seurat)
library(DESeq2)
library(ggplot2)
library(dplyr)
library(ggrastr)
library(ggrepel)
library(ggpubr)
library(viridis)
library(ggblend)
library(SummarizedExperiment)
library(SingleCellExperiment)
library(nlme)
library(emmeans)
```

If the snakemake pipeline was run as-is, the following files will be generated in the folder from which the pipeline was run. If necessary, adjust paths accordingly.

## Setup

### Paths

```{r paths}
base_path <- "/mnt/smb/ltytgat/ayliyim/MI-group/ayliyim/projects/TIPRJ0000015_MAPS/maps_crc"

# Subsets
seurat_hc_pbmc_rds <- file.path(base_path, "output/q1_pf_characterization/subsets/hc_pbmc_SeuratObject.Rds")
seurat_hc_pf_rds <- file.path(base_path, "output/q1_pf_characterization/subsets/hc_pf_SeuratObject.Rds")
seurat_hc_pbmc_pf_rds <- file.path(base_path, "output/q1_pf_characterization/subsets/hc_pbmc_pf_SeuratObject.Rds")
seurat_hc_pf_macrophages_rds <- file.path(base_path, "output/q1_pf_characterization/subsets/hc_pf_macrophages_SeuratObject.Rds")
sce_hc_pf_macrophages_trajectory_rds <- file.path(base_path, "output/q1_pf_characterization/subsets/hc_pf_macrophages_trajectory_sce.Rds")
seurat_pbmc_pf_liver_colon_rds <- file.path(base_path, "output/q1_pf_characterization/subsets/pbmc_pf_liver_colon_SeuratObject.Rds")

# Plotting parameters
celltype_markers_xlsx <- file.path(base_path, "config/order/celltype_markers.xlsx")
pf_heatmap_order_xlsx <- file.path(base_path, "config/order/q1_pf_heatmap_order.xlsx")
```

### Import data

```{r load subsets}
seurat_hc_pbmc <- readRDS(seurat_hc_pbmc_rds)
seurat_hc_pbmc_pf <- readRDS(seurat_hc_pbmc_pf_rds)
seurat_hc_pf <- readRDS(seurat_hc_pf_rds)
seurat_hc_pf_macrophages <- readRDS(seurat_hc_pf_macrophages_rds)
sce_hc_pf_macrophages_trajectory <- readRDS(sce_hc_pf_macrophages_trajectory_rds )
seurat_pbmc_pf_liver_colon <- readRDS(seurat_pbmc_pf_liver_colon_rds)
```

```{r load marker genes}
marker_genes <- readxl::read_excel(pf_heatmap_order_xlsx) %>%
  dplyr::filter(Assay == "RNA")
```

### Output

```{r figures}
figS2Dir <- file.path(base_path, "docs/manuscript/figures/figS2")
dir.create(figS2Dir, recursive = T)
```

### Plotting parameters

```{r plotting order}
pf_heatmap_order_xlsx <- file.path(base_path, "config/order/q1_pf_heatmap_order.xlsx")

manual_l3_order <- readxl::read_excel(celltype_markers_xlsx) %>%
  dplyr::filter(level == "manual_l3") %>%
  dplyr::pull(celltype) %>%
  unique()

celltype_order_l2 <- readxl::read_excel(celltype_markers_xlsx, col_names = T) %>%
  dplyr::filter(level == "manual_l2",
                modality == "gene") %>%
  dplyr::select(celltype, color) %>%
  unique() %>%
  dplyr::filter(celltype %in% unique(seurat_hc_pf@meta.data[,"manual_l2"])) %>%
  dplyr::mutate(number_subset = paste0(1:nrow(.), ". ", celltype))

celltype_order_pf_l2_l3 <- readxl::read_excel(celltype_markers_xlsx, col_names = T) %>%
  dplyr::filter(level == "manual_l3",
                modality == "gene") %>%
  dplyr::select(celltype, color) %>%
  unique() %>%
  dplyr::filter(celltype %in% unique(seurat_hc_pf@meta.data[,"manual_l3"])) %>%
  dplyr::mutate(celltype = ifelse(celltype %in% c("Classical monocytes", "Macrophages VCAN+", "Macrophages C1Q+", "Macrophages VCAN+C1Q+", "Macrophages SPP1+"), "Mono-macs", celltype),
                color = ifelse(celltype %in% c("Mono-macs"), "#E31A1C", color)) %>%
  unique() %>%
  dplyr::mutate(number_subset = paste0(1:nrow(.), ". ", celltype))

celltype_order_pf_l4 <- readxl::read_excel(celltype_markers_xlsx, col_names = T) %>%
  dplyr::filter(level == "manual_l4",
                modality == "gene") %>%
  dplyr::select(celltype, color) %>%
  unique() %>%
  dplyr::filter(celltype %in% unique(seurat_hc_pf@meta.data[,"manual_l4"])) %>%
  dplyr::mutate(number_subset = paste0(1:nrow(.), ". ", celltype))
```

```{r colors}
tissue_colors <- c(PF = "#224FBD", PBMC = "#F30000", Colon = "#7B3F00", Liver = "#AA336A")

manual_l2_colors <- celltype_order_l2$color
names(manual_l2_colors) <- celltype_order_l2$celltype

manual_l2_l3_number_pf_colors <- celltype_order_pf_l2_l3$color
names(manual_l2_l3_number_pf_colors) <- celltype_order_pf_l2_l3$number_subset

manual_l4_colors_pf <- celltype_order_pf_l4$color
names(manual_l4_colors_pf) <- celltype_order_pf_l4$celltype
```

## Prepare data

```{r prepare subset}
seurat_hc_pf_l2l3 <- seurat_hc_pf

seurat_hc_pf_l2l3@meta.data <- seurat_hc_pf_l2l3@meta.data %>%
  dplyr::mutate(celltype = as.character(manual_l3),
                celltype = ifelse(celltype %in% c("Classical monocytes", "Macrophages C1Q+", "Macrophages SPP1+", "Macrophages VCAN+", "Macrophages VCAN+C1Q+"), "Mono-macs", as.character(celltype)),
                celltype = factor(celltype, levels = celltype_order_pf_l2_l3$celltype),
                celltype_number = as.numeric(celltype),
                celltype_w_number = paste0(as.numeric(celltype), ". ", celltype),
                celltype_w_number = factor(celltype_w_number, levels = celltype_order_pf_l2_l3$number_subset))
```

## Figures

### Dotplot GEX HC PBMC colored by l3

Here we present a dotplot representing the percentage expressing cells and median expression of curated genes for the HC PBMC-derived l3 subsets.

```{r dotplot gex hc pbmc col_cellmarkers, fig.width = 6.5, fig.height=15}
marker_gex <- GetAssayData(seurat_hc_pbmc, assay = "RNA")[which(rownames(GetAssayData(seurat_hc_pbmc, assay = "RNA")) %in% unique(marker_genes$FeatureID)), ]

dotplot_hc_pbmc_cellmarkers_gex_ggplotobj <- data.frame(FeatureID = rownames(marker_gex), marker_gex) %>%
  tidyr::pivot_longer(-c(FeatureID), names_to = "CB", values_to = "nUMIs") %>%
  dplyr::mutate(CB = stringr::str_replace(CB, "\\.", "-")) %>%
  dplyr::inner_join(data.frame(CB = rownames(seurat_hc_pbmc@meta.data), 
                               Celltype = seurat_hc_pbmc@meta.data[,"manual_l3"]), 
                    by = "CB") %>%
  dplyr::mutate(Celltype = factor(Celltype), 
                FeatureID = factor(FeatureID)) %>%
  dplyr::group_by(Celltype, FeatureID, .drop = F) %>%
  dplyr::summarize(Median = median(log1p(nUMIs)),
                   Percentage = mean(nUMIs>0)*100) %>%
  dplyr::mutate(Median = ifelse(is.na(Median), 0, Median),
                Percentage = ifelse(is.na(Percentage), 0, Percentage),
                Celltype = factor(Celltype, manual_l3_order),
                FeatureID = factor(FeatureID, rev(marker_genes$FeatureID))) %>% 
  ggplot(aes(x = Celltype, y = FeatureID, col = Median)) +
  geom_tile(alpha = 0, col = "grey") +
  geom_point(aes(size = Percentage)) +
  theme_bw() +
  theme(legend.position = "bottom", 
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.spacing.x=unit(0, "lines"))

pdf(file.path(figS2Dir, "dotplot_hc_pbmc_cellmarkers_gex_ggplotobj.pdf"), width = 6, height=15)
print(dotplot_hc_pbmc_cellmarkers_gex_ggplotobj)
dev.off()
```

### Boxplot HC PBMC PF lineage relative to all immune cells

Here we prepare the data for scattered boxplots depicting the HC-derived PBMC and PF immune lineages relative to the total immune population. Actual data was plotted in Graphpad Prism.

```{r hc pbmc pf l1rl0 data}
seurat_hc_pbmc_pf@meta.data %>%
  dplyr::group_by(manual_l1, SampleID, Tissue, .drop = T) %>%
  dplyr::summarise(Ncells = n()) %>%
  dplyr::ungroup() %>%
  tidyr::complete(tidyr::nesting(SampleID), tidyr::nesting(manual_l1), fill = list(Ncells = 0)) %>%
  dplyr::group_by(SampleID, Tissue) %>%
  dplyr::mutate(Nlrefsample = sum(Ncells),
                Ncellprop = Ncells/Nlrefsample,
                Ncellprop = ifelse(is.na(Ncellprop), 0, Ncellprop),
                Ncellperc = Ncellprop*100,
                Donor = gsub("^(pt[0-9]{2})_.+$", "\\1", SampleID)) %>%
  readr::write_csv(file = file.path(figS2Dir, "hc_pbmc_pf_l1rl0.csv"))
```

### UMAP HC PF colored by l3

Here we color the PF-derived immune cells by l2/l3.

```{r umap hc pf col_l2l3 lab_l2l3, fig.width=7.5, fig.height=7.5}
umap_hc_pf_coll2l3_labl2l3_ggplotobj <- data.frame(CB = colnames(seurat_hc_pf_l2l3),
                                                   Embeddings(seurat_hc_pf_l2l3[["wnn.umap"]]),
                                                   seurat_hc_pf_l2l3@meta.data) %>%
  ggplot(aes(x = wnnUMAP_1, y = wnnUMAP_2)) +
  geom_point_rast(show.legend = T, size = 0.25, aes(col = celltype_w_number)) +
  labs(y = "",
       x = "") +
  geom_label_repel(data = . %>%
                     dplyr::group_by(celltype, celltype_number, celltype_w_number) %>%
                     summarize(x = median(x = wnnUMAP_1),
                               y = median(x = wnnUMAP_2)),
                   mapping = aes(label = celltype_number, x = x, y = y),
                   alpha = 1,
                   size = 8,
                   show.legend = F, 
                   label.size = 0.25) +
  guides(colour = guide_legend(override.aes = list(size = 3))) +
  scale_color_manual(values = manual_l2_l3_number_pf_colors) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.x = element_text(face = "bold"))

pdf(file.path(figS2Dir, "umap_hc_pf_coll2l3_labl2l3.pdf"), width = 7.5, height = 7.5)
print(umap_hc_pf_coll2l3_labl2l3_ggplotobj)
dev.off()
```

### Boxplot HC PBMC PF celltype relative to all immune cells

Here we prepare the data for scattered boxplots depicting the HC-derived PBMC and PF immune cell types relative to the total immune population. Actual data was plotted in Graphpad Prism.

```{r hc pbmc pf l3rl0 data}
seurat_hc_pbmc_pf@meta.data %>%
  dplyr::mutate(celltype = manual_l3,
                celltype = ifelse(celltype %in% c("Classical monocytes", "Macrophages C1Q+", "Macrophages SPP1+", "Macrophages VCAN+", "Macrophages VCAN+C1Q+"), "Mono-macs", celltype),
                celltype = factor(celltype, levels = celltype_order_pf_l2_l3$celltype)) %>%
  dplyr::group_by(celltype, SampleID, Tissue, .drop = T) %>%
  dplyr::summarise(Ncells = n()) %>%
  dplyr::ungroup() %>%
  tidyr::complete(tidyr::nesting(SampleID), tidyr::nesting(celltype), fill = list(Ncells = 0)) %>%
  dplyr::group_by(SampleID, Tissue) %>%
  dplyr::mutate(Nlrefsample = sum(Ncells),
                Ncellprop = Ncells/Nlrefsample,
                Ncellprop = ifelse(is.na(Ncellprop), 0, Ncellprop),
                Ncellperc = Ncellprop*100,
                Donor = gsub("^(pt[0-9]{2})_.+$", "\\1", SampleID),
                Tissue = gsub("^pt[0-9]{2}_(.+)$", "\\1", SampleID)) %>%
  readr::write_csv(file = file.path(figS2Dir, "hc_pbmc_pf_l2l3rl0.csv"))
```

### UMAP HC PF macrophages colored by monocyte-macrophage GEX markers

Here we characterize the PF-derived CD163+ macrophages by superimposing the gene expression of the following genes on the respective UMAP configuration:
- Monocytes/Monocyte-macrophages/Macrophages (CD14)
- Monocytes/Monocyte-macrophages/Macrophages (CCR2)
- Peritoneal resident macrophages (TREM2)

```{r umap hc pf macrophages col_monomacgexmarkers, fig.width=9, fig.height=3}
umap_hc_pf_macrophages_colmonomacgexmarkers_ggplotobj_list <- lapply(
  data.frame(CB = colnames(seurat_hc_pf_macrophages),
             Embeddings(seurat_hc_pf_macrophages[["gex.umap"]]),
             seurat_hc_pf_macrophages@meta.data,
             expr = GetAssayData(seurat_hc_pf_macrophages, assay = "RNA")["CD14",],
             Feature = "CD14") %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_hc_pf_macrophages),
                                  Embeddings(seurat_hc_pf_macrophages[["gex.umap"]]),
                                  seurat_hc_pf_macrophages@meta.data,
                                  expr = GetAssayData(seurat_hc_pf_macrophages, assay = "RNA")["CCR2",],
                                  Feature = "CCR2")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_hc_pf_macrophages),
                                  Embeddings(seurat_hc_pf_macrophages[["gex.umap"]]),
                                  seurat_hc_pf_macrophages@meta.data,
                                  expr = GetAssayData(seurat_hc_pf_macrophages, assay = "RNA")["TREM2",],
                                  Feature = "TREM2")) %>%
    dplyr::mutate(expr_rank = rank(expr, ties.method="first"),
                  Feature = factor(Feature, levels = c("CD14", "CCR2", "TREM2"))) %>%
    split(., .$Feature), FUN = function(plotdf){
      plotdf %>% dplyr::arrange(expr) %>%
        ggplot(aes(x = gexUMAP_1, y = gexUMAP_2)) +
        geom_point_rast(show.legend = F, size = 0.25, col = "#d3d3d3") +
        geom_point_rast(data = . %>%
                          dplyr::filter(expr>0), aes(col = expr), show.legend = T) +
        guides(colour = guide_legend(override.aes = list(size = 3))) +
        facet_wrap(~Feature) +
        guides(colour = guide_legend(override.aes = list(size = 3),
                                     title = "Expression")) +
        scale_color_viridis() +
        theme_bw() +
        theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              legend.position = "none",
              axis.title.x = element_blank(),
              axis.title.y = element_blank(),
              strip.text.x = element_text(face = "bold"))
    }
)

pdf(file.path(figS2Dir, "umap_hc_pf_macrophages_colmonomacgexmarkers.pdf"), width = 9, height = 3)
print(ggarrange(plotlist = umap_hc_pf_macrophages_colmonomacgexmarkers_ggplotobj_list, ncol = 3))
dev.off()
```

### UMAP HC PF macrophages colored by monocyte-macrophage CITE markers

Here we characterize the PF-derived CD163+ macrophages by superimposing the gene expression of the following genes on the respective UMAP configuration:
- Monocytes/Monocyte-macrophages/Macrophages (CD14)
- Monocytes/Monocyte-macrophages/Macrophages (CD192)

```{r umap hc pf macrophages col_monomaccitemarkers, fig.width=6, fig.height=3}
umap_hc_pf_macrophages_colmonomaccitemarkers_ggplotobj_list <- lapply(
  data.frame(CB = colnames(seurat_hc_pf_macrophages),
             Embeddings(seurat_hc_pf_macrophages[["gex.umap"]]),
             seurat_hc_pf_macrophages@meta.data,
             expr = GetAssayData(seurat_hc_pf_macrophages, assay = "CITE")["Hu.CD14-M5E2",],
             Feature = "CD14") %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_hc_pf_macrophages),
                                  Embeddings(seurat_hc_pf_macrophages[["gex.umap"]]),
                                  seurat_hc_pf_macrophages@meta.data,
                                  expr = GetAssayData(seurat_hc_pf_macrophages, assay = "CITE")["Hu.CD192",],
                                  Feature = "CD192")) %>%
    dplyr::mutate(expr_rank = rank(expr, ties.method="first"),
                  Feature = factor(Feature, levels = c("CD14", "CD192"))) %>%
    split(., .$Feature), FUN = function(plotdf){
      plotdf %>% dplyr::arrange(expr) %>%
        ggplot(aes(x = gexUMAP_1, y = gexUMAP_2)) +
        geom_point_rast(show.legend = F, size = 0.25, col = "#d3d3d3") +
        geom_point_rast(data = . %>%
                          dplyr::filter(expr>0), aes(col = expr), show.legend = T) +
        guides(colour = guide_legend(override.aes = list(size = 3))) +
        facet_wrap(~Feature) +
        guides(colour = guide_legend(override.aes = list(size = 3),
                                     title = "Expression")) +
        scale_colour_gradient(low = "#ffffff", high = "#FF0000") +
        theme_bw() +
        theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              legend.position = "none",
              axis.title.x = element_blank(),
              axis.title.y = element_blank(),
              strip.text.x = element_text(face = "bold"))
    }
)

pdf(file.path(figS2Dir, "umap_hc_pf_macrophages_colmonomaccitemarkers.pdf"), width = 6, height = 3)
print(ggarrange(plotlist = umap_hc_pf_macrophages_colmonomaccitemarkers_ggplotobj_list, ncol = 2))
dev.off()
```

### Boxplot HC PF macrophages entropy

Here we plot the entropy of HC-derived macrophages, depicting the degree of "maturity". 

```{r boxplot hc pf macrophages entropy col_l4, fig.width=5, fig.height=5}
boxplot_hc_pf_macrophages_entropy_ggplotobj <- colData(sce_hc_pf_macrophages_trajectory) %>%
  data.frame() %>%
  ggplot(aes(x = forcats::fct_reorder(manual_l4, -entropy), y = entropy, fill = manual_l4)) +
  geom_point_rast(position = position_jitterdodge(), alpha = 0.3, size = 0.05) +
  geom_boxplot(outlier.shape = NA) +
  labs(y = "Entropy") +
  theme_bw() +
  scale_fill_manual(values = manual_l4_colors_pf) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

pdf(file.path(figS2Dir, "boxplot_hc_pf_macrophages_entropy.pdf"), width = 3, height=4)
print(boxplot_hc_pf_macrophages_entropy_ggplotobj)
dev.off()
```

Here we calculate the differential entropy in a pairwise fashion across the HC-derived macrophages by using a linear mixed effects model where donor is a random effects. This data was added post-hoc to the visualization. 

```{r hc pf macrophages entropy differential analysis}
hc_pf_macrophages_trajectory_coldata <- colData(sce_hc_pf_macrophages_trajectory) %>%
  data.frame() %>%
  dplyr::filter(manual_l2 != "Monocytes") %>%
  dplyr::mutate(Donor = as.factor(Donor),
                Tissue = as.factor(Tissue))

hc_pf_macrophages_trajectory_lme <- lme(entropy ~ manual_l4, random = ~1|Donor, data = hc_pf_macrophages_trajectory_coldata)
hc_pf_macrophages_trajectory_lme_emmeans <- emmeans(hc_pf_macrophages_trajectory_lme, pairwise ~ manual_l4)
```

### Stackedbarplot HC PBMC, PF, Liver and Colon l2rl0 colored by l2

Here we plot the cell types relative to the total immune population for HC-PBMC, HC-PF, Liver, and Colon. The Liver and Colon dataset were obtained from the Liver Cell Atlas and the Gut Cell Atlas, respectively, and were collected from (seemingly) healthy donors.

```{r stackedbarplot hc pbmc pf liver colon l2rl0 col_l2, fig.width=5, fig.height=5}
stackedbarplot_hc_pbmc_pf_liver_colon_l2rl0_coll2_ggplotobj <- seurat_pbmc_pf_liver_colon@meta.data %>%
  dplyr::group_by(manual_l2, Tissue, .drop = T) %>% 
  dplyr::summarise(Ncells = n()) %>%
  dplyr::ungroup() %>%
  tidyr::complete(tidyr::nesting(Tissue), tidyr::nesting(manual_l2), fill = list(Ncells = 0)) %>%
  dplyr::group_by(Tissue) %>%
  dplyr::mutate(manual_l2 = factor(manual_l2, levels = celltype_order_l2$celltype),
                Nsample = sum(Ncells),
                Ncellprop = Ncells/Nsample,
                Ncellprop = ifelse(is.na(Ncellprop), 0, Ncellprop),
                Ncellperc = Ncellprop*100,
                Tissue = factor(Tissue, levels = c("PBMC", "PF", "Colon", "Liver"))) %>%
  dplyr::rename(Celltype = manual_l2) %>%
  ggplot(aes(x = Tissue, y = Ncellperc)) +
  geom_bar(position="stack", stat="identity", aes(fill = Celltype)) +
  labs(y = "%CD45+") +
  scale_fill_manual(values = manual_l2_colors) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "bottom",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

pdf(file.path(figS2Dir, "stackedbarplot_hc_pbmc_pf_liver_colon_l2rl0_coll2.pdf"), width = 5, height = 5)
print(stackedbarplot_hc_pbmc_pf_liver_colon_l2rl0_coll2_ggplotobj)
dev.off()

seurat_pbmc_pf_liver_colon@meta.data %>%
  dplyr::group_by(manual_l2, Tissue, .drop = T) %>% 
  dplyr::summarise(Ncells = n()) %>%
  dplyr::ungroup() %>%
  tidyr::complete(tidyr::nesting(Tissue), tidyr::nesting(manual_l2), fill = list(Ncells = 0)) %>%
  dplyr::group_by(Tissue) %>%
  dplyr::mutate(manual_l2 = factor(manual_l2, levels = celltype_order_l2$celltype),
                Nsample = sum(Ncells),
                Ncellprop = Ncells/Nsample,
                Ncellprop = ifelse(is.na(Ncellprop), 0, Ncellprop),
                Ncellperc = Ncellprop*100,
                Tissue = factor(Tissue, levels = c("PBMC", "PF", "Colon", "Liver"))) %>%
  dplyr::rename(Celltype = manual_l2) %>%
  readr::write_csv(file.path(figS2Dir, "stackedbarplot_hc_pbmc_pf_liver_colon_l2rl0.csv"))
```
