---
title: 'Figure SZ: IS Liver'
author: "Andrew Y.F. Li Yim"
date: '2024-07-31'
output: html_document
---

```{r libraries, include=FALSE}
library(Seurat)
library(ggplot2)
library(dplyr)
library(ggrastr)
library(ggrepel)
library(ggpubr)
library(viridis)
library(ggblend)
```

If the snakemake pipeline was run as-is, the following files will be generated in the folder from which the pipeline was run. If necessary, adjust paths accordingly.

## Setup

### Paths

```{r paths}
base_path <- "/mnt/smb/ltytgat/ayliyim/MI-group/ayliyim/projects/TIPRJ0000015_MAPS/maps_crc"
#base_path <- "/media/hdd1/andrewliyim/maps_crc"

# Subsets
seurat_liver_rds <- file.path(base_path, "resources/liver/liver_seuratobject.rds")
seurat_liver_myeloid_rds <- file.path(base_path, "output/q1_pf_characterization/subsets/liver_mnp_SeuratObject.Rds")
seurat_liver_macrophages_rds <- file.path(base_path, "output/q1_pf_characterization/subsets/liver_macrophages_SeuratObject.Rds")

# Plotting parameters
celltype_markers_xlsx <- file.path(base_path, "config/order/celltype_markers.xlsx")
pf_heatmap_order_xlsx <- file.path(base_path, "config/order/q1_pf_heatmap_order.xlsx")
```

### Import data

```{r load subsets}
seurat_liver <- readRDS(seurat_liver_rds)
seurat_liver_myeloid <- readRDS(seurat_liver_myeloid_rds)
seurat_liver_macrophages <- readRDS(seurat_liver_macrophages_rds)
```

## Figures

### Dotplot liver colored by l2

```{r dotplot liver col_l2, fig.width = 3.75, fig.height=15}
marker_genes <- readxl::read_excel(pf_heatmap_order_xlsx) %>%
  dplyr::filter(Assay == "RNA")
marker_gex <- GetAssayData(seurat_liver, assay = "RNA")[which(rownames(GetAssayData(seurat_liver, assay = "RNA")) %in% unique(marker_genes$FeatureID)), ]

dotplot_liver_l2_cellmarkers_gex_ggplotobj <- data.frame(FeatureID = rownames(marker_gex), marker_gex) %>%
  tidyr::pivot_longer(-c(FeatureID), names_to = "CB", values_to = "nUMIs") %>%
  dplyr::mutate(CB = stringr::str_replace(CB, "\\.", "-")) %>%
  dplyr::inner_join(data.frame(CB = rownames(seurat_liver@meta.data), 
                               Celltype = seurat_liver@meta.data[,"cell_type_MAPS_L2"]), 
                    by = "CB") %>%
  dplyr::mutate(Celltype = factor(Celltype, levels = c("CD4T", "CD8T", "NK CD16-", "B naive", "B intermediate", "B memory", "Plasma B", "Monocytes", "Macrophages", "CDCs", "PDCs", "Granulocytes")), 
                FeatureID = factor(FeatureID)) %>%
  dplyr::group_by(Celltype, FeatureID, .drop = F) %>%
  dplyr::summarize(Median = median(log1p(nUMIs)),
                   Percentage = mean(nUMIs>0)*100) %>%
  dplyr::mutate(Median = ifelse(is.na(Median), 0, Median),
                Percentage = ifelse(is.na(Percentage), 0, Percentage),
                Celltype = factor(Celltype),
                FeatureID = factor(FeatureID, rev(marker_genes$FeatureID))) %>% 
  ggplot(aes(x = Celltype, y = FeatureID, col = Median)) +
  geom_tile(alpha = 0, col = "grey") +
  geom_point(aes(size = Percentage)) +
  theme_bw() +
  theme(legend.position = "bottom", 
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.spacing.x=unit(0, "lines"))

pdf("dotplot_liver_l2_cellmarkers_gex_ggplotobj.pdf", width = 3.75, height=15)
print(dotplot_liver_l2_cellmarkers_gex_ggplotobj)
dev.off()
```

### UMAP liver colored by lineage genes

```{r umap liver col_lineage_markers, fig.width=10, fig.height=3}
umap_liver_collineagegenemarkers_ggplotobj_list <- lapply(
  data.frame(CB = colnames(seurat_liver),
             Embeddings(seurat_liver[["umap"]]),
             seurat_liver@meta.data,
             expr = GetAssayData(seurat_liver, layers = "data")["CD3D",],
             Feature = "CD3D") %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver),
                                  Embeddings(seurat_liver[["umap"]]),
                                  seurat_liver@meta.data,
                                  expr = GetAssayData(seurat_liver, layers = "data")["NCAM1",],
                                  Feature = "NCAM1")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver),
                                  Embeddings(seurat_liver[["umap"]]),
                                  seurat_liver@meta.data,
                                  expr = GetAssayData(seurat_liver, layers = "data")["MS4A1",],
                                  Feature = "MS4A1")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver),
                                  Embeddings(seurat_liver[["umap"]]),
                                  seurat_liver@meta.data,
                                  expr = GetAssayData(seurat_liver, layers = "data")["ITGAM",],
                                  Feature = "ITGAM")) %>%
    dplyr::mutate(expr_rank = rank(expr, ties.method="first"),
                  Feature = factor(Feature, levels = c("CD3D", "NCAM1", "MS4A1", "ITGAM"))) %>%
    split(., .$Feature), FUN = function(plotdf){
      plotdf %>% dplyr::arrange(expr) %>%
        ggplot(aes(x = UMAP_1, y = UMAP_2)) +
        geom_point_rast(show.legend = F, size = 0.25, col = "#d3d3d3") +
        geom_point_rast(data = . %>%
                          dplyr::filter(expr>0), aes(col = expr), show.legend = T) +
        guides(colour = guide_legend(override.aes = list(size = 3))) +
        facet_wrap(~Feature) +
        guides(colour = guide_legend(override.aes = list(size = 3),
                                     title = "Expression")) +
        scale_color_viridis() +
        #scale_colour_gradient(low = "#ffffff", high = "#FF0000") +
        theme_bw() +
        theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              legend.position = "none",
              axis.text.x = element_blank(),
              axis.text.y = element_blank(),
              axis.ticks = element_blank(),
              axis.title.x = element_blank(),
              axis.title.y = element_blank(),
              strip.text.x = element_text(face = "bold"))
    }
)

print(ggarrange(plotlist = umap_liver_collineagegenemarkers_ggplotobj_list, nrow = 1, ncol = 4))

pdf("umap_liver_collineagegenemarkers_colviridis.pdf", width = 10, height = 3)
print(ggarrange(plotlist = umap_liver_collineagegenemarkers_ggplotobj_list, nrow = 1, ncol = 4))
dev.off()
```

### UMAP liver mononuclear phagocytes colored by marker genes

```{r umap liver col_myeloidgenemarkers}
umap_liver_myeloid_colmyeloidgenemarkers_ggplotobj_list <- lapply(
  data.frame(CB = colnames(seurat_liver_myeloid),
             Embeddings(seurat_liver_myeloid[["umap"]]),
             seurat_liver_myeloid@meta.data,
             expr = GetAssayData(seurat_liver_myeloid, layers = "data")["CD1C",],
             Feature = "CD1C") %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid,  layers = "data")["CLEC9A",],
                                  Feature = "CLEC9A")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid,  layers = "data")["CLEC4C",],
                                  Feature = "CLEC4C")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid,  layers = "data")["SPP1",],
                                  Feature = "SPP1")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid,  layers = "data")["MARCO",],
                                  Feature = "MARCO")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid,  layers = "data")["CD163",],
                                  Feature = "CD163")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid,  layers = "data")["VCAN",],
                                  Feature = "VCAN")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid,  layers = "data")["C1QA",],
                                  Feature = "C1QA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid,  layers = "data")["VSIG4",],
                                  Feature = "VSIG4")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid,  layers = "data")["TREM2",],
                                  Feature = "TREM2")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid,  layers = "data")["FN1",],
                                  Feature = "FN1")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid,  layers = "data")["CSF1R",],
                                  Feature = "CSF1R")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid,  layers = "data")["ITGAM",],
                                  Feature = "ITGAM")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid,  layers = "data")["ITGAX",],
                                  Feature = "ITGAX")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid,  layers = "data")["CD14",],
                                  Feature = "CD14")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = 0, #Expression is undetectable
                                  Feature = "GATA6")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid,  layers = "data")["TIMD4",],
                                  Feature = "TIMD4")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid,  layers = "data")["MRC1",],
                                  Feature = "MRC1")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid,  layers = "data")["CCR2",],
                                  Feature = "CCR2")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid,  layers = "data")["LYVE1",],
                                  Feature = "LYVE1")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid,  layers = "data")["XCR1",],
                                  Feature = "XCR1")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid,  layers = "data")["S100A8",],
                                  Feature = "S100A8")) %>%
    dplyr::mutate(expr_rank = rank(expr, ties.method="first"),
                  Feature = factor(Feature, levels = c("CD1C", "CLEC9A", "CLEC4C", "SPP1", "MARCO", "CD163", "VCAN", "C1QA", "VSIG4", "TREM2", "FN1", "CSF1R", "ITGAM", "ITGAX", "CD14", "GATA6", "TIMD4", "MRC1", "CCR2", "LYVE1", "XCR1", "S100A8"))) %>%
    dplyr::filter(manual_l2 %in% c("Macrophages", "CDCs", "PDCs")) %>%
    split(., .$Feature), FUN = function(plotdf){
      plotdf %>% dplyr::arrange(expr) %>%
        ggplot(aes(x = UMAP_1, y = UMAP_2)) +
        geom_point_rast(show.legend = F, size = 0.25, col = "#d3d3d3") +
        geom_point_rast(data = . %>%
                          dplyr::filter(expr>0), aes(col = expr), show.legend = T) +
        guides(colour = guide_legend(override.aes = list(size = 3))) +
        facet_wrap(~Feature) +
        guides(colour = guide_legend(override.aes = list(size = 3),
                                     title = "Expression")) +
        # xlim(-10,11) +
        # ylim(-10,8) +
        scale_color_viridis() +
        #scale_colour_gradient(low = "#ffffff", high = "#FF0000") +
        theme_bw() +
        theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              legend.position = "none",
              # axis.text.x = element_blank(),
              # axis.ticks.x = element_blank(),
              axis.title.x = element_blank(),
              # axis.text.y = element_blank(),
              # axis.ticks.y = element_blank(),
              axis.title.y = element_blank(),
              strip.text.x = element_text(face = "bold"))
    }
)

print(ggarrange(plotlist = umap_liver_myeloid_colmyeloidgenemarkers_ggplotobj_list, nrow = 5, ncol = 5))

pdf("umap_liver_myeloid_colmyeloidgenemarkers_colviridis.pdf", width = 15, height = 15)
print(ggarrange(plotlist = umap_liver_myeloid_colmyeloidgenemarkers_ggplotobj_list, nrow = 5, ncol = 5))
dev.off()
```

### UMAP liver macrophages colored by marker genes

```{r umap liver macrophages col_myeloidgenemarkers, fig.width=12, fig.height=12}
umap_liver_macrophages_colmyeloidgenemarkers_ggplotobj_list <- lapply(
  data.frame(CB = colnames(seurat_liver_macrophages ),
             Embeddings(seurat_liver_macrophages[["umap"]]),
             seurat_liver_macrophages@meta.data,
             expr = GetAssayData(seurat_liver_macrophages, layers = "data")["CD1C",],
             Feature = "CD1C",
             Modality = "RNA") %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_macrophages ),
                                  Embeddings(seurat_liver_macrophages[["umap"]]),
                                  seurat_liver_macrophages@meta.data,
                                  expr = 0, #Expression is undetectable
                                  Feature = "SPP1",
                                  Modality = "RNA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_macrophages ),
                                  Embeddings(seurat_liver_macrophages[["umap"]]),
                                  seurat_liver_macrophages@meta.data,
                                  expr = GetAssayData(seurat_liver_macrophages, layers = "data")["MARCO",],
                                  Feature = "MARCO",
                                  Modality = "RNA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_macrophages ),
                                  Embeddings(seurat_liver_macrophages[["umap"]]),
                                  seurat_liver_macrophages@meta.data,
                                  expr = GetAssayData(seurat_liver_macrophages, layers = "data")["CD163",],
                                  Feature = "CD163",
                                  Modality = "RNA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_macrophages ),
                                  Embeddings(seurat_liver_macrophages[["umap"]]),
                                  seurat_liver_macrophages@meta.data,
                                  expr = GetAssayData(seurat_liver_macrophages, layers = "data")["VCAN",],
                                  Feature = "VCAN",
                                  Modality = "RNA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_macrophages ),
                                  Embeddings(seurat_liver_macrophages[["umap"]]),
                                  seurat_liver_macrophages@meta.data,
                                  expr = GetAssayData(seurat_liver_macrophages, layers = "data")["C1QA",],
                                  Feature = "C1QA",
                                  Modality = "RNA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_macrophages ),
                                  Embeddings(seurat_liver_macrophages[["umap"]]),
                                  seurat_liver_macrophages@meta.data,
                                  expr = GetAssayData(seurat_liver_macrophages, layers = "data")["VSIG4",],
                                  Feature = "VSIG4",
                                  Modality = "RNA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_macrophages ),
                                  Embeddings(seurat_liver_macrophages[["umap"]]),
                                  seurat_liver_macrophages@meta.data,
                                  expr = GetAssayData(seurat_liver_macrophages, layers = "data")["TREM2",],
                                  Feature = "TREM2",
                                  Modality = "RNA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_macrophages ),
                                  Embeddings(seurat_liver_macrophages[["umap"]]),
                                  seurat_liver_macrophages@meta.data,
                                  expr = GetAssayData(seurat_liver_macrophages, layers = "data")["FN1",],
                                  Feature = "FN1",
                                  Modality = "RNA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_macrophages ),
                                  Embeddings(seurat_liver_macrophages[["umap"]]),
                                  seurat_liver_macrophages@meta.data,
                                  expr = GetAssayData(seurat_liver_macrophages, layers = "data")["CSF1R",],
                                  Feature = "CSF1R",
                                  Modality = "RNA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_macrophages ),
                                  Embeddings(seurat_liver_macrophages[["umap"]]),
                                  seurat_liver_macrophages@meta.data,
                                  expr = GetAssayData(seurat_liver_macrophages, layers = "data")["ITGAM",],
                                  Feature = "ITGAM",
                                  Modality = "RNA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_macrophages ),
                                  Embeddings(seurat_liver_macrophages[["umap"]]),
                                  seurat_liver_macrophages@meta.data,
                                  expr = GetAssayData(seurat_liver_macrophages, layers = "data")["ITGAX",],
                                  Feature = "ITGAX",
                                  Modality = "RNA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_macrophages ),
                                  Embeddings(seurat_liver_macrophages[["umap"]]),
                                  seurat_liver_macrophages@meta.data,
                                  expr = GetAssayData(seurat_liver_macrophages, layers = "data")["CD14",],
                                  Feature = "CD14",
                                  Modality = "RNA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_macrophages ),
                                  Embeddings(seurat_liver_macrophages[["umap"]]),
                                  seurat_liver_macrophages@meta.data,
                                  expr = 0, #Expression is undetectable
                                  Feature = "GATA6",
                                  Modality = "RNA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_macrophages ),
                                  Embeddings(seurat_liver_macrophages[["umap"]]),
                                  seurat_liver_macrophages@meta.data,
                                  expr = GetAssayData(seurat_liver_macrophages, layers = "data")["TIMD4",],
                                  Feature = "TIMD4",
                                  Modality = "RNA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_macrophages ),
                                  Embeddings(seurat_liver_macrophages[["umap"]]),
                                  seurat_liver_macrophages@meta.data,
                                  expr = GetAssayData(seurat_liver_macrophages, layers = "data")["MRC1",],
                                  Feature = "MRC1",
                                  Modality = "RNA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_macrophages ),
                                  Embeddings(seurat_liver_macrophages[["umap"]]),
                                  seurat_liver_macrophages@meta.data,
                                  expr = 0, #Expression is undetectable
                                  Feature = "CCR2",
                                  Modality = "RNA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_macrophages ),
                                  Embeddings(seurat_liver_macrophages[["umap"]]),
                                  seurat_liver_macrophages@meta.data,
                                  expr = GetAssayData(seurat_liver_macrophages, layers = "data")["LYVE1",],
                                  Feature = "LYVE1",
                                  Modality = "RNA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_macrophages ),
                                  Embeddings(seurat_liver_macrophages[["umap"]]),
                                  seurat_liver_macrophages@meta.data,
                                  expr = 0, #Expression is undetectable
                                  Feature = "XCR1",
                                  Modality = "RNA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_macrophages ),
                                  Embeddings(seurat_liver_macrophages[["umap"]]),
                                  seurat_liver_macrophages@meta.data,
                                  expr = GetAssayData(seurat_liver_macrophages, layers = "data")["S100A8",],
                                  Feature = "S100A8",
                                  Modality = "RNA")) %>%
    dplyr::mutate(expr_rank = rank(expr, ties.method="first"),
                  Feature = factor(Feature, levels = c("CD1C", "SPP1", "MARCO", "CD163", "VCAN", "C1QA", "VSIG4", "TREM2", "FN1", "CSF1R", "ITGAM", "ITGAX", "CD14", "GATA6", "TIMD4", "MRC1", "CCR2", "LYVE1", "XCR1", "S100A8"))) %>%
    split(., .$Feature), FUN = function(plotdf){
      plotdf %>% dplyr::arrange(expr) %>%
        ggplot(aes(x = UMAP_1, y = UMAP_2)) +
        geom_point_rast(show.legend = F, size = 0.25, col = "#d3d3d3") +
        geom_point_rast(data = . %>%
                          dplyr::filter(expr>0), aes(col = expr), show.legend = T) +
        guides(colour = guide_legend(override.aes = list(size = 3))) +
        facet_wrap(~Feature) +
        guides(colour = guide_legend(override.aes = list(size = 3),
                                     title = "Expression")) +
        scale_color_viridis() +
        #scale_colour_gradient(low = "#ffffff", high = "#FF0000") +
        theme_bw() +
        theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              legend.position = "none",
              # axis.text.x = element_blank(),
              # axis.ticks.x = element_blank(),
              axis.title.x = element_blank(),
              # axis.text.y = element_blank(),
              # axis.ticks.y = element_blank(),
              axis.title.y = element_blank(),
              strip.text.x = element_text(face = "bold"))
    }
)

print(ggarrange(plotlist = umap_liver_macrophages_colmyeloidgenemarkers_ggplotobj_list, nrow = 5, ncol = 5))

pdf("umap_liver_macrophages_colmyeloidgenemarkers_colviridis.pdf", width = 15, height = 15)
print(ggarrange(plotlist = umap_liver_macrophages_colmyeloidgenemarkers_ggplotobj_list, nrow = 5, ncol = 5))
dev.off()
```