---
title: 'Supplementary Figure 3: Characterization of macrophages in healthy colon and liver'
author: "Andrew Y.F. Li Yim"
date: '2024-07-31'
output: html_document
---

```{r libraries, include=FALSE}
library(Seurat)
library(ggplot2)
library(dplyr)
library(ggrastr)
library(ggrepel)
library(ggpubr)
library(viridis)
library(ggblend)
```

If the snakemake pipeline was run as-is, the following files will be generated in the folder from which the pipeline was run. If necessary, adjust paths accordingly.

## Setup

### Paths

```{r paths}
base_path <- "/mnt/smb/ltytgat/ayliyim/MI-group/ayliyim/projects/TIPRJ0000015_MAPS/maps_crc"

# Subsets
seurat_colon_rds <- file.path(base_path, "resources/colon/colon_seuratobject.rds")
seurat_colon_myeloid_rds <- file.path(base_path, "output/q1_pf_characterization/subsets/colon_mnp_SeuratObject.Rds")
seurat_colon_macrophages_rds <- file.path(base_path, "output/q1_pf_characterization/subsets/colon_macrophages_SeuratObject.Rds")
seurat_liver_rds <- file.path(base_path, "resources/liver/liver_seuratobject.rds")
seurat_liver_myeloid_rds <- file.path(base_path, "output/q1_pf_characterization/subsets/liver_mnp_SeuratObject.Rds")
seurat_liver_macrophages_rds <- file.path(base_path, "output/q1_pf_characterization/subsets/liver_macrophages_SeuratObject.Rds")

# Plotting parameters
celltype_markers_xlsx <- file.path(base_path, "config/order/celltype_markers.xlsx")
pf_heatmap_order_xlsx <- file.path(base_path, "config/order/q1_pf_heatmap_order.xlsx")
```

### Import data

```{r load subsets}
seurat_colon <- readRDS(seurat_colon_rds)
seurat_colon_myeloid <- readRDS(seurat_colon_myeloid_rds)
seurat_colon_macrophages <- readRDS(seurat_colon_macrophages_rds)
seurat_liver <- readRDS(seurat_liver_rds)
seurat_liver_myeloid <- readRDS(seurat_liver_myeloid_rds)
seurat_liver_macrophages <- readRDS(seurat_liver_macrophages_rds)
```

```{r load marker genes}
marker_genes <- readxl::read_excel(pf_heatmap_order_xlsx) %>%
  dplyr::filter(Assay == "RNA")
```

### Output

```{r figures}
figS3Dir <- file.path(base_path, "docs/manuscript/figures/figS3")
dir.create(figS3Dir, recursive = T)
```

## Figures

### UMAP colon colored by lineage genes

Here we characterize the Gut Cell Atlas healthy colon-derived immune cells by superimposing the gene expression of the following genes on the respective UMAP configuration:
- T (CD3D)
- NK/ILC (NCAM1)
- B (MS4A1)
- Myeloid (ITGAM)

```{r umap colon col_lineagemarkers, fig.width=10, fig.height=3}
umap_colon_collineagegenemarkers_ggplotobj_list <- lapply(
  data.frame(CB = colnames(seurat_colon),
             Embeddings(seurat_colon[["umap"]]),
             seurat_colon@meta.data,
             expr = GetAssayData(seurat_colon, layers = "data")["CD3D",],
             Feature = "CD3D") %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_colon),
                                  Embeddings(seurat_colon[["umap"]]),
                                  seurat_colon@meta.data,
                                  expr = GetAssayData(seurat_colon, layers = "data")["NCAM1",],
                                  Feature = "NCAM1")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_colon),
                                  Embeddings(seurat_colon[["umap"]]),
                                  seurat_colon@meta.data,
                                  expr = GetAssayData(seurat_colon, layers = "data")["MS4A1",],
                                  Feature = "MS4A1")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_colon),
                                  Embeddings(seurat_colon[["umap"]]),
                                  seurat_colon@meta.data,
                                  expr = GetAssayData(seurat_colon, layers = "data")["ITGAM",],
                                  Feature = "ITGAM")) %>%
    dplyr::mutate(expr_rank = rank(expr, ties.method="first"),
                  Feature = factor(Feature, levels = c("CD3D", "NCAM1", "MS4A1", "ITGAM"))) %>%
    split(., .$Feature), FUN = function(plotdf){
      plotdf %>% dplyr::arrange(expr) %>%
        ggplot(aes(x = UMAP_1, y = UMAP_2)) +
        geom_point_rast(show.legend = F, size = 0.25, col = "#d3d3d3") +
        geom_point_rast(data = . %>%
                          dplyr::filter(expr>0), aes(col = expr), show.legend = T) +
        guides(colour = guide_legend(override.aes = list(size = 3))) +
        facet_wrap(~Feature) +
        guides(colour = guide_legend(override.aes = list(size = 3),
                                     title = "Expression")) +
        scale_color_viridis() +
        theme_bw() +
        theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              legend.position = "none",
              axis.text.x = element_blank(),
              axis.text.y = element_blank(),
              axis.ticks = element_blank(),
              axis.title.x = element_blank(),
              axis.title.y = element_blank(),
              strip.text.x = element_text(face = "bold"))
    }
)

pdf(file.path(figS3Dir, "umap_colon_collineagegenemarkers.pdf"), width = 10, height = 3)
print(ggarrange(plotlist = umap_colon_collineagegenemarkers_ggplotobj_list, nrow = 1, ncol = 4))
dev.off()
```

### UMAP liver colored by lineage genes

Here we characterize the Liver Cell Atlas healthy colon-derived immune cells by superimposing the gene expression of the following genes on the respective UMAP configuration:
- T (CD3D)
- NK/ILC (NCAM1)
- B (MS4A1)
- Myeloid (ITGAM)

```{r umap liver col_lineagemarkers, fig.width=10, fig.height=3}
umap_liver_collineagegenemarkers_ggplotobj_list <- lapply(
  data.frame(CB = colnames(seurat_liver),
             Embeddings(seurat_liver[["umap"]]),
             seurat_liver@meta.data,
             expr = GetAssayData(seurat_liver, layers = "data")["CD3D",],
             Feature = "CD3D") %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver),
                                  Embeddings(seurat_liver[["umap"]]),
                                  seurat_liver@meta.data,
                                  expr = GetAssayData(seurat_liver, layers = "data")["NCAM1",],
                                  Feature = "NCAM1")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver),
                                  Embeddings(seurat_liver[["umap"]]),
                                  seurat_liver@meta.data,
                                  expr = GetAssayData(seurat_liver, layers = "data")["MS4A1",],
                                  Feature = "MS4A1")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver),
                                  Embeddings(seurat_liver[["umap"]]),
                                  seurat_liver@meta.data,
                                  expr = GetAssayData(seurat_liver, layers = "data")["ITGAM",],
                                  Feature = "ITGAM")) %>%
    dplyr::mutate(expr_rank = rank(expr, ties.method="first"),
                  Feature = factor(Feature, levels = c("CD3D", "NCAM1", "MS4A1", "ITGAM"))) %>%
    split(., .$Feature), FUN = function(plotdf){
      plotdf %>% dplyr::arrange(expr) %>%
        ggplot(aes(x = UMAP_1, y = UMAP_2)) +
        geom_point_rast(show.legend = F, size = 0.25, col = "#d3d3d3") +
        geom_point_rast(data = . %>%
                          dplyr::filter(expr>0), aes(col = expr), show.legend = T) +
        guides(colour = guide_legend(override.aes = list(size = 3))) +
        facet_wrap(~Feature) +
        guides(colour = guide_legend(override.aes = list(size = 3),
                                     title = "Expression")) +
        scale_color_viridis() +
        theme_bw() +
        theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              legend.position = "none",
              axis.text.x = element_blank(),
              axis.text.y = element_blank(),
              axis.ticks = element_blank(),
              axis.title.x = element_blank(),
              axis.title.y = element_blank(),
              strip.text.x = element_text(face = "bold"))
    }
)

pdf(file.path(figS3Dir, "umap_liver_collineagegenemarkers.pdf"), width = 10, height = 3)
print(ggarrange(plotlist = umap_liver_collineagegenemarkers_ggplotobj_list, nrow = 1, ncol = 4))
dev.off()
```

### UMAP colon mononuclear phagocytes colored by mononuclear phagocytes marker genes

Here we characterize the Gut Cell Atlas healthy colon-derived mononuclear phagocytes by superimposing the gene expression of the following genes on the respective UMAP configuration:
- Plasmacytoid dendritic cells (CLEC4C)
- Conventional dendritic cells type 1 (XCR1)
- Conventional dendritic cells type 2 (CD1C)
- Myeloid (ITGAM)
- Macrophages (MARCO)
- Macrophages (CD163)
- Monocyte-macrophages (CD14)
- Monocytes (CCR2)

```{r umap colon col_mnpmarkers, fig.width=6, fig.height=12}
umap_colon_mnp_colmnpgenemarkers_ggplotobj_list <- lapply(
  data.frame(CB = colnames(seurat_colon_myeloid),
             Embeddings(seurat_colon_myeloid[["umap"]]),
             seurat_colon_myeloid@meta.data,
             expr = GetAssayData(seurat_colon_myeloid, layers = "data")["CLEC4C",],
             Feature = "CLEC4C") %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_colon_myeloid ),
                                  Embeddings(seurat_colon_myeloid[["umap"]]),
                                  seurat_colon_myeloid@meta.data,
                                  expr = GetAssayData(seurat_colon_myeloid, layers = "data")["XCR1",],
                                  Feature = "XCR1")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_colon_myeloid ),
                                  Embeddings(seurat_colon_myeloid[["umap"]]),
                                  seurat_colon_myeloid@meta.data,
                                  expr = GetAssayData(seurat_colon_myeloid, layers = "data")["CD1C",],
                                  Feature = "CD1C")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_colon_myeloid ),
                                  Embeddings(seurat_colon_myeloid[["umap"]]),
                                  seurat_colon_myeloid@meta.data,
                                  expr = GetAssayData(seurat_colon_myeloid, layers = "data")["ITGAM",],
                                  Feature = "ITGAM")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_colon_myeloid ),
                                  Embeddings(seurat_colon_myeloid[["umap"]]),
                                  seurat_colon_myeloid@meta.data,
                                  expr = GetAssayData(seurat_colon_myeloid, layers = "data")["MARCO",],
                                  Feature = "MARCO")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_colon_myeloid ),
                                  Embeddings(seurat_colon_myeloid[["umap"]]),
                                  seurat_colon_myeloid@meta.data,
                                  expr = GetAssayData(seurat_colon_myeloid, layers = "data")["CD163",],
                                  Feature = "CD163")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_colon_myeloid ),
                                  Embeddings(seurat_colon_myeloid[["umap"]]),
                                  seurat_colon_myeloid@meta.data,
                                  expr = GetAssayData(seurat_colon_myeloid, layers = "data")["CD14",],
                                  Feature = "CD14")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_colon_myeloid ),
                                  Embeddings(seurat_colon_myeloid[["umap"]]),
                                  seurat_colon_myeloid@meta.data,
                                  expr = GetAssayData(seurat_colon_myeloid, layers = "data")["CCR2",],
                                  Feature = "CCR2")) %>%
    dplyr::mutate(expr_rank = rank(expr, ties.method="first"),
                  Feature = factor(Feature, levels = c("CLEC4C", "XCR1", "CD1C", "ITGAM", "MARCO", "CD163", "CD14", "CCR2"))) %>%
    dplyr::filter(manual_l2 %in% c("Macrophages", "CDCs", "PDCs")) %>%
    split(., .$Feature), FUN = function(plotdf){
      plotdf %>% dplyr::arrange(expr) %>%
        ggplot(aes(x = UMAP_1, y = UMAP_2)) +
        geom_point_rast(show.legend = F, size = 0.25, col = "#d3d3d3") +
        geom_point_rast(data = . %>%
                          dplyr::filter(expr>0), aes(col = expr), show.legend = T) +
        guides(colour = guide_legend(override.aes = list(size = 3))) +
        facet_wrap(~Feature) +
        guides(colour = guide_legend(override.aes = list(size = 3),
                                     title = "Expression")) +
        scale_color_viridis() +
        theme_bw() +
        theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              legend.position = "none",
              axis.title.x = element_blank(),
              axis.title.y = element_blank(),
              strip.text.x = element_text(face = "bold"))
    }
)

pdf(file.path(figS3Dir, "umap_colon_mnp_colmnpgenemarkers.pdf"), width = 6, height = 12)
print(ggarrange(plotlist = umap_colon_mnp_colmnpgenemarkers_ggplotobj_list, nrow = 4, ncol = 2))
dev.off()
```

### UMAP liver mononuclear phagocytes colored by mononuclear phagocytes marker genes

Here we characterize the Liver Cell Atlas healthy liver-derived mononuclear phagocytes by superimposing the gene expression of the following genes on the respective UMAP configuration:
- Plasmacytoid dendritic cells (CLEC4C)
- Conventional dendritic cells type 1 (XCR1)
- Conventional dendritic cells type 2 (CD1C)
- Myeloid (ITGAM)
- Macrophages (MARCO)
- Macrophages (CD163)
- Monocyte-macrophages (CD14)
- Monocytes (CCR2)

```{r umap liver col_mnpmarkers, fig.width=6, fig.height=12}
umap_liver_mnp_colmnpgenemarkers_ggplotobj_list <- lapply(
  data.frame(CB = colnames(seurat_liver_myeloid),
             Embeddings(seurat_liver_myeloid[["umap"]]),
             seurat_liver_myeloid@meta.data,
             expr = GetAssayData(seurat_liver_myeloid, layers = "data")["CLEC4C",],
             Feature = "CLEC4C") %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid, layers = "data")["XCR1",],
                                  Feature = "XCR1")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid, layers = "data")["CD1C",],
                                  Feature = "CD1C")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid, layers = "data")["ITGAM",],
                                  Feature = "ITGAM")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid, layers = "data")["MARCO",],
                                  Feature = "MARCO")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid, layers = "data")["CD163",],
                                  Feature = "CD163")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid, layers = "data")["CD14",],
                                  Feature = "CD14")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_myeloid ),
                                  Embeddings(seurat_liver_myeloid[["umap"]]),
                                  seurat_liver_myeloid@meta.data,
                                  expr = GetAssayData(seurat_liver_myeloid, layers = "data")["CCR2",],
                                  Feature = "CCR2")) %>%
    dplyr::mutate(expr_rank = rank(expr, ties.method="first"),
                  Feature = factor(Feature, levels = c("CLEC4C", "XCR1", "CD1C", "ITGAM", "MARCO", "CD163", "CD14", "CCR2"))) %>%
    dplyr::filter(manual_l2 %in% c("Macrophages", "CDCs", "PDCs")) %>%
    split(., .$Feature), FUN = function(plotdf){
      plotdf %>% dplyr::arrange(expr) %>%
        ggplot(aes(x = UMAP_1, y = UMAP_2)) +
        geom_point_rast(show.legend = F, size = 0.25, col = "#d3d3d3") +
        geom_point_rast(data = . %>%
                          dplyr::filter(expr>0), aes(col = expr), show.legend = T) +
        guides(colour = guide_legend(override.aes = list(size = 3))) +
        facet_wrap(~Feature) +
        guides(colour = guide_legend(override.aes = list(size = 3),
                                     title = "Expression")) +
        scale_color_viridis() +
        theme_bw() +
        theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              legend.position = "none",
              axis.title.x = element_blank(),
              axis.title.y = element_blank(),
              strip.text.x = element_text(face = "bold"))
    }
)

pdf(file.path(figS3Dir, "umap_liver_mnp_colmnpgenemarkers_colviridis.pdf"), width = 6, height = 12)
print(ggarrange(plotlist = umap_liver_mnp_colmnpgenemarkers_ggplotobj_list, nrow = 4, ncol = 2))
dev.off()
```

### UMAP colon macrophages colored by macrophage marker genes

Here we characterize the Gut Cell Atlas healthy colon-derived macrophages by superimposing the gene expression of the following genes on the respective UMAP configuration:
- Macrophages (CD163)
- Macrophages C1Q+ [aka "CMs C1Q+"] (C1QA)
- Macrophages VCAN+ [aka "mono-CMs"] (VCAN)
- Monocytes (CCR2)
- Macrophages SPP1+ [aka "CMs SPP1+"] (SPP1)

```{r umap colon macrophages col_macrophagegenemarkers, fig.width=6, fig.height=9}
umap_colon_macrophages_colmacrophagegenemarkers_ggplotobj_list <- lapply(
  data.frame(CB = colnames(seurat_colon_macrophages ),
             Embeddings(seurat_colon_macrophages[["umap"]]),
             seurat_colon_macrophages@meta.data,
             expr = GetAssayData(seurat_colon_macrophages, layers = "data")["CD163",],
             Feature = "CD163") %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_colon_macrophages ),
                                  Embeddings(seurat_colon_macrophages[["umap"]]),
                                  seurat_colon_macrophages@meta.data,
                                  expr = GetAssayData(seurat_colon_macrophages, layers = "data")["C1QA",],
                                  Feature = "C1QA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_colon_macrophages ),
                                  Embeddings(seurat_colon_macrophages[["umap"]]),
                                  seurat_colon_macrophages@meta.data,
                                  expr = GetAssayData(seurat_colon_macrophages, layers = "data")["VCAN",],
                                  Feature = "VCAN")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_colon_macrophages ),
                                  Embeddings(seurat_colon_macrophages[["umap"]]),
                                  seurat_colon_macrophages@meta.data,
                                  expr = GetAssayData(seurat_colon_macrophages, layers = "data")["CCR2",],
                                  Feature = "CCR2")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_colon_macrophages ),
                                  Embeddings(seurat_colon_macrophages[["umap"]]),
                                  seurat_colon_macrophages@meta.data,
                                  expr = GetAssayData(seurat_colon_macrophages, layers = "data")["SPP1",],
                                  Feature = "SPP1")) %>%
    dplyr::mutate(expr_rank = rank(expr, ties.method="first"),
                  Feature = factor(Feature, levels = c("CD163", "C1QA", "VCAN", "CCR2", "SPP1"))) %>%
    split(., .$Feature), FUN = function(plotdf){
      plotdf %>% dplyr::arrange(expr) %>%
        ggplot(aes(x = UMAP_1, y = UMAP_2)) +
        geom_point_rast(show.legend = F, size = 0.25, col = "#d3d3d3") +
        geom_point_rast(data = . %>%
                          dplyr::filter(expr>0), aes(col = expr), show.legend = T) +
        guides(colour = guide_legend(override.aes = list(size = 3))) +
        facet_wrap(~Feature) +
        guides(colour = guide_legend(override.aes = list(size = 3),
                                     title = "Expression")) +
        scale_color_viridis() +
        theme_bw() +
        theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              legend.position = "none",
              axis.title.x = element_blank(),
              axis.title.y = element_blank(),
              strip.text.x = element_text(face = "bold"))
    }
)

pdf(file.path(figS3Dir, "umap_colon_macrophages_colmacrophagegenemarkers.pdf"), width = 6, height = 9)
print(ggarrange(plotlist = umap_colon_macrophages_colmacrophagegenemarkers_ggplotobj_list, nrow = 3, ncol = 2))
dev.off()
```

### UMAP liver macrophages colored by macrophage marker genes

Here we characterize the Liver Cell Atlas healthy liver-derived macrophages by superimposing the gene expression of the following genes on the respective UMAP configuration:
- Macrophages (CD163)
- Macrophages C1Q+ [aka "CMs C1Q+"] (C1QA)
- Macrophages VCAN+ [aka "mono-CMs"] (VCAN)
- Monocytes (CCR2)
- Macrophages SPP1+ [aka "CMs SPP1+"] (SPP1)

```{r umap liver macrophages col_macrophagegenemarkers, fig.width=6, fig.height=9}
umap_liver_macrophages_colmacrophagegenemarkers_ggplotobj_list <- lapply(
  data.frame(CB = colnames(seurat_liver_macrophages ),
             Embeddings(seurat_liver_macrophages[["umap"]]),
             seurat_liver_macrophages@meta.data,
             expr = GetAssayData(seurat_liver_macrophages, layers = "data")["CD163",],
             Feature = "CD163") %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_macrophages ),
                                  Embeddings(seurat_liver_macrophages[["umap"]]),
                                  seurat_liver_macrophages@meta.data,
                                  expr = GetAssayData(seurat_liver_macrophages, layers = "data")["C1QA",],
                                  Feature = "C1QA")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_macrophages ),
                                  Embeddings(seurat_liver_macrophages[["umap"]]),
                                  seurat_liver_macrophages@meta.data,
                                  expr = GetAssayData(seurat_liver_macrophages, layers = "data")["VCAN",],
                                  Feature = "VCAN")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_macrophages ),
                                  Embeddings(seurat_liver_macrophages[["umap"]]),
                                  seurat_liver_macrophages@meta.data,
                                  expr = 0, #CCR2 is not detected in the selected cells.
                                  Feature = "CCR2")) %>%
    dplyr::rows_append(data.frame(CB = colnames(seurat_liver_macrophages ),
                                  Embeddings(seurat_liver_macrophages[["umap"]]),
                                  seurat_liver_macrophages@meta.data,
                                  expr = 0, #SPP1 is not detected in the selected cells.
                                  Feature = "SPP1")) %>%
    dplyr::mutate(expr_rank = rank(expr, ties.method="first"),
                  Feature = factor(Feature, levels = c("CD163", "C1QA", "VCAN", "CCR2", "SPP1"))) %>%
    split(., .$Feature), FUN = function(plotdf){
      plotdf %>% dplyr::arrange(expr) %>%
        ggplot(aes(x = UMAP_1, y = UMAP_2)) +
        geom_point_rast(show.legend = F, size = 0.25, col = "#d3d3d3") +
        geom_point_rast(data = . %>%
                          dplyr::filter(expr>0), aes(col = expr), show.legend = T) +
        guides(colour = guide_legend(override.aes = list(size = 3))) +
        facet_wrap(~Feature) +
        guides(colour = guide_legend(override.aes = list(size = 3),
                                     title = "Expression")) +
        scale_color_viridis() +
        theme_bw() +
        theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              legend.position = "none",
              axis.title.x = element_blank(),
              axis.title.y = element_blank(),
              strip.text.x = element_text(face = "bold"))
    }
)

pdf(file.path(figS3Dir, "umap_liver_macrophages_colmacrophagegenemarkers.pdf"), width = 6, height = 9)
print(ggarrange(plotlist = umap_liver_macrophages_colmacrophagegenemarkers_ggplotobj_list, nrow = 3, ncol = 2))
dev.off()
```

### Dotplot colon colored by l2

Here we present a dotplot representing the percentage expressing cells and median expression of curated genes for the Gut Cell Atlas healthy colon-derived l2 subsets.

```{r dotplot colon col_l2, fig.width = 3.75, fig.height=15}
marker_gex_colon <- GetAssayData(seurat_colon, assay = "RNA")[which(rownames(GetAssayData(seurat_colon, assay = "RNA")) %in% unique(marker_genes$FeatureID)), ]

dotplot_colon_l2_cellgexmarkers_ggplotobj <- data.frame(FeatureID = rownames(marker_gex_colon), marker_gex_colon) %>%
  tidyr::pivot_longer(-c(FeatureID), names_to = "CB", values_to = "nUMIs") %>%
  dplyr::mutate(CB = stringr::str_replace(CB, "\\.", "-")) %>%
  dplyr::inner_join(data.frame(CB = rownames(seurat_colon@meta.data), 
                               Celltype = seurat_colon@meta.data[,"cell_type_MAPS_L2"]), 
                    by = "CB") %>%
  dplyr::mutate(Celltype = factor(Celltype, levels = c("CD4T", "CD8T", "other T", "NK CD16-", "ILCs", "B naive", "B intermediate", "B memory", "Plasma B", "Monocytes", "Macrophages", "CDCs", "Granulocytes")), 
                FeatureID = factor(FeatureID)) %>%
  dplyr::group_by(Celltype, FeatureID, .drop = F) %>%
  dplyr::summarize(Median = median(log1p(nUMIs)),
                   Percentage = mean(nUMIs>0)*100) %>%
  dplyr::mutate(Median = ifelse(is.na(Median), 0, Median),
                Percentage = ifelse(is.na(Percentage), 0, Percentage),
                Celltype = factor(Celltype),
                FeatureID = factor(FeatureID, rev(marker_genes$FeatureID))) %>% 
  ggplot(aes(x = Celltype, y = FeatureID, col = Median)) +
  geom_tile(alpha = 0, col = "grey") +
  geom_point(aes(size = Percentage)) +
  theme_bw() +
  theme(legend.position = "bottom", 
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.spacing.x=unit(0, "lines"))

pdf(file.path(figS3Dir, "dotplot_colon_l2_cellgexmarkers.pdf"), width = 3.75, height=15)
print(dotplot_colon_l2_cellgexmarkers_ggplotobj)
dev.off()
```

### Dotplot liver colored by l2

Here we present a dotplot representing the percentage expressing cells and median expression of curated genes for the Liver Cell Atlas healthy liver-derived l2 subsets.

```{r dotplot liver col_l2, fig.width = 3.75, fig.height=15}
marker_liver_gex <- GetAssayData(seurat_liver, assay = "RNA")[which(rownames(GetAssayData(seurat_liver, assay = "RNA")) %in% unique(marker_genes$FeatureID)), ]

dotplot_liver_l2_cellgexmarkers_ggplotobj <- data.frame(FeatureID = rownames(marker_liver_gex), marker_liver_gex) %>%
  tidyr::pivot_longer(-c(FeatureID), names_to = "CB", values_to = "nUMIs") %>%
  dplyr::mutate(CB = stringr::str_replace(CB, "\\.", "-")) %>%
  dplyr::inner_join(data.frame(CB = rownames(seurat_liver@meta.data), 
                               Celltype = seurat_liver@meta.data[,"cell_type_MAPS_L2"]), 
                    by = "CB") %>%
  dplyr::mutate(Celltype = factor(Celltype, levels = c("CD4T", "CD8T", "NK CD16-", "B naive", "B intermediate", "B memory", "Plasma B", "Monocytes", "Macrophages", "CDCs", "PDCs", "Granulocytes")), 
                FeatureID = factor(FeatureID)) %>%
  dplyr::group_by(Celltype, FeatureID, .drop = F) %>%
  dplyr::summarize(Median = median(log1p(nUMIs)),
                   Percentage = mean(nUMIs>0)*100) %>%
  dplyr::mutate(Median = ifelse(is.na(Median), 0, Median),
                Percentage = ifelse(is.na(Percentage), 0, Percentage),
                Celltype = factor(Celltype),
                FeatureID = factor(FeatureID, rev(marker_genes$FeatureID))) %>% 
  ggplot(aes(x = Celltype, y = FeatureID, col = Median)) +
  geom_tile(alpha = 0, col = "grey") +
  geom_point(aes(size = Percentage)) +
  theme_bw() +
  theme(legend.position = "bottom", 
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.spacing.x=unit(0, "lines"))

pdf(file.path(figS3Dir, "dotplot_liver_l2_cellgexmarkers.pdf"), width = 3.75, height=15)
print(dotplot_liver_l2_cellgexmarkers_ggplotobj)
dev.off()
```